---
title: "ngs101_scRNAseq_walkthrough_4 - Part 1. Manual and automated cell type annotation"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-4-cell-type-identification/>

Walkthrough 3 produced an integrated, batch-corrected, and clustered Seurat
object. Clustering only groups transcriptionally similar cells together -
it says nothing about *what those cells are*. This walkthrough converts
"Cluster 0, 1, 2 ..." into biologically meaningful labels such as "CD4+ T
cells" or "Classical Monocytes."

There is no single correct way to do this, so the original tutorial (and
this walkthrough) uses **four complementary approaches** and then manually
reconciles the results into a consensus:

1. **Manual annotation** - canonical marker genes plus data-driven marker
   discovery (`FindAllMarkers`). Time-consuming but the closest thing to a
   gold standard, because it incorporates biological reasoning that no
   algorithm currently replicates.
2. **SingleR** - reference-based annotation. Correlates each cell's
   expression profile against a curated, pre-labeled reference atlas
   (`celldex`) and transfers the best-matching label.
3. **scType** - marker-based gene set scoring. Uses a curated database of
   positive/negative marker genes per cell type and scores each cluster's
   enrichment.
4. **scCATCH** - tissue-specific automated annotation. Automatically finds
   cluster marker genes and matches them against a tissue-specific reference
   database (`cellmatch`).

Different methods fail in different ways, so agreement across methods builds
confidence, and disagreement flags clusters that need closer inspection.
This first part of the walkthrough runs all four methods; Part 2 compares
them, builds a manual consensus, and saves the final annotated object.

**Dataset reminder**: GSE174609 is a periodontitis study (8 samples: 4
Healthy + 4 Post-Treatment PBMC samples). Peripheral blood is an ideal
dataset for learning annotation because it has well-characterized cell
types, clear canonical markers, and mature reference atlases.

## Libraries and directories

```{r step1_libs_data}

#-----------------------------------------------
# STEP 1: Load required libraries
#-----------------------------------------------

# Core single-cell analysis
library(Seurat)
library(SeuratObject)

# Automated annotation methods
library(SingleR)
library(celldex)
library(scCATCH)

# Data structures
library(SingleCellExperiment)

# Marker processing (scType dependency)
library(HGNChelper)
library(openxlsx)

# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(scales)
library(viridis)

# Set working directory
base.dir <- "~/GSE174609_scRNA"
integr.dir <- file.path(base.dir, "integration_analysis")
annot.dir <- file.path(base.dir, "cell_type_annotation")

# Create output directories
dir.create(file.path(annot.dir,"plots"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(annot.dir,"plots/manual_annotation"), showWarnings = FALSE)
dir.create(file.path(annot.dir,"plots/automated_annotation"), showWarnings = FALSE)
dir.create(file.path(annot.dir,"plots/method_comparison"), showWarnings = FALSE)
dir.create(file.path(annot.dir,"annotations"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(42)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Load integrated data from Part 3

```{r step2_load_objs}

#-----------------------------------------------
# STEP 2: Load integrated data from Part 3
#-----------------------------------------------

# Object file name from walkthrough 3, Part 3 (already batch-corrected,
# clustered at the chosen resolution, with sample_id/condition/patient_id
# metadata attached)
integrated.file <- "seurat_integrated_clustered.rds"

cat("Reading in integrated Seurat object:", integrated.file, "\n")
merged_seurat_integrated <- readRDS(file.path(integr.dir,"integrated_data",integrated.file))

# Retrieve the integration method and reduction chosen in Part 2 of
# walkthrough 3 (chosen_method, chosen_reduction)
load(file.path(integr.dir,"metadata","chosen_methods.RData"))

cat("\nUsing", chosen_method, "integration; UMAP reduction:", chosen_reduction, "\n")
cat("Cells:", ncol(merged_seurat_integrated), " Genes:", nrow(merged_seurat_integrated), "\n")
```

### Fallback: auto-detect the UMAP reduction

If `chosen_methods.RData` isn't available for some reason, the original
tutorial's approach of scanning the object's reductions for a UMAP by name
is a reasonable fallback. Not evaluated here since `chosen_reduction` was
loaded above, but kept for reference.

```{r step2b_reduction_fallback, eval = FALSE}
available_reductions <- names(merged_seurat_integrated@reductions)
if ("umap.harmony" %in% available_reductions) {
  chosen_reduction <- "umap.harmony"
} else if ("umap.cca" %in% available_reductions) {
  chosen_reduction <- "umap.cca"
} else if ("umap.rpca" %in% available_reductions) {
  chosen_reduction <- "umap.rpca"
} else if ("umap.mnn" %in% available_reductions) {
  chosen_reduction <- "umap.mnn"
} else {
  chosen_reduction <- "umap"
}
```

## Visualize starting point

```{r step3_starting_clusters, fig.width = 14, fig.height = 10}

#-----------------------------------------------
# STEP 3: Visualize clusters before annotation
#-----------------------------------------------

p1 <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
              group.by = "seurat_clusters", label = TRUE, label.size = 5,
              pt.size = 0.1) +
  ggtitle("Clusters (Pre-Annotation)") +
  theme(plot.title = element_text(face = "bold", size = 14))

p2 <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
              group.by = "sample_id", pt.size = 0.1) +
  ggtitle("Samples") +
  theme(legend.text = element_text(size = 7))

p3 <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
              group.by = "condition", pt.size = 0.1) +
  ggtitle("Condition") +
  scale_color_manual(values = c("Healthy" = "#2E86AB",
                                 "Post_Treatment" = "#F18F01"))

(p_overview <- (p1 | p2) / (p3 | plot_spacer()))

ggsave(file.path(annot.dir,"plots/00_starting_clusters.png"), p_overview,
       width = 14, height = 10, dpi = 300)
```

Top-left shows the numbered clusters we're about to name. Top-right checks
that samples are still well-mixed within clusters (a sanity check on the
Part 3 integration). Bottom-left shows how the two conditions distribute
across clusters - useful context once cell types are assigned.

## Preparing SCT-normalized data for annotation

**Deviation from the original tutorial**: the source article calls
`NormalizeData()` and `ScaleData()` directly on the integrated object,
which assumes the standard `RNA` assay workflow. This walkthrough series
has used `SCTransform` throughout (see walkthrough 3, Part 1), so
`merged_seurat_integrated` already carries normalized (`data`) and scaled
(`scale.data`) values in its `SCT` assay - re-running `NormalizeData` /
`ScaleData` is unnecessary and would not target the right assay.

```{r step4_assay_check}
DefaultAssay(merged_seurat_integrated) <- "SCT"
cat("Default assay:", DefaultAssay(merged_seurat_integrated), "\n")
cat("Assays present:", paste(Assays(merged_seurat_integrated), collapse = ", "), "\n")
```

### A caveat: `return.only.var.genes` and missing canonical markers

Walkthrough 3's `SCTransform` call used `return.only.var.genes = TRUE`
(the default), which means only genes selected as highly variable have
corrected values in the `data` and `scale.data` layers of the `SCT` assay.
Canonical marker genes used for manual annotation are not guaranteed to be
highly variable (e.g. broadly-expressed lineage markers can have low
cell-to-cell variance), so some may be missing from these layers and would
silently fail to plot with `VlnPlot`/`FeaturePlot`/`DotPlot`/`DoHeatmap`.

`Seurat::GetResidual()` computes corrected values for specific genes on
demand, regardless of whether they were originally flagged as variable.
Run this before checking marker presence below.

```{r step4b_get_residual}
canonical_markers_flat <- c(
  "CD3D", "CD3E", "CD3G", "CD4", "IL7R", "CD8A", "CD8B",
  "CD79A", "MS4A1", "CD19", "NKG7", "GNLY", "NCAM1",
  "CD14", "LYZ", "S100A8", "S100A9", "FCGR3A", "MS4A7",
  "FCER1A", "CST3", "IL3RA", "GZMB", "SERPINF1",
  "PPBP", "PF4", "GP9"
)

markers_present_in_object <- canonical_markers_flat[canonical_markers_flat %in% rownames(merged_seurat_integrated)]
markers_missing_from_data <- setdiff(
  markers_present_in_object,
  rownames(LayerData(merged_seurat_integrated, assay = "SCT", layer = "scale.data"))
)

if (length(markers_missing_from_data) > 0) {
  cat("Computing SCT residuals for", length(markers_missing_from_data),
      "canonical marker(s) not in the variable feature set:\n")
  cat(" ", paste(markers_missing_from_data, collapse = ", "), "\n")
  merged_seurat_integrated <- GetResidual(merged_seurat_integrated,
                                          features = markers_missing_from_data,
                                          verbose = FALSE)
} else {
  cat("All canonical markers already present in SCT scale.data\n")
}
```

## Manual cell type annotation: the gold standard

### Understanding PBMC cell types

Approximate expected composition of a healthy human PBMC sample (varies
with disease, age, and technical factors):

| Cell Type | Expected % | Key Markers | Function |
|---|---|---|---|
| T cells | 50-70% | CD3D, CD3E, CD3G | Adaptive immunity |
| &nbsp;&nbsp;CD4+ T cells | 30-40% | CD3+, CD4+, IL7R | Helper T cells |
| &nbsp;&nbsp;CD8+ T cells | 20-30% | CD3+, CD8A, CD8B | Cytotoxic T cells |
| B cells | 10-15% | CD79A, MS4A1 (CD20), CD19 | Antibody production |
| NK cells | 5-10% | NKG7, GNLY, NCAM1 (CD56) | Innate cytotoxicity |
| Monocytes | 10-20% | CD14, LYZ, S100A8, S100A9 | Phagocytosis |
| &nbsp;&nbsp;Classical | ~80% of mono | CD14+, FCGR3A- (CD16-) | CD14+ monocytes |
| &nbsp;&nbsp;Non-classical | ~20% of mono | CD14low, FCGR3A+ (CD16+) | CD16+ monocytes |
| Dendritic cells | 1-2% | FCER1A, CST3 | Antigen presentation |
| Plasmacytoid DC | <1% | IL3RA, GZMB, SERPINF1 | Type I interferon |
| Platelets | Variable | PPBP, PF4, GP9 | Usually contamination |

### Step 1: canonical marker genes

```{r step5_canonical_markers}

#-----------------------------------------------
# STEP 5: Define and check canonical markers
#-----------------------------------------------

# Define comprehensive marker panel for PBMC cell types
canonical_markers <- list(
  "T_cells" = c("CD3D", "CD3E", "CD3G"),
  "CD4_T" = c("CD3D", "CD4", "IL7R"),
  "CD8_T" = c("CD3D", "CD8A", "CD8B"),
  "B_cells" = c("CD79A", "MS4A1", "CD19"),
  "NK_cells" = c("NKG7", "GNLY", "NCAM1"),
  "Monocytes" = c("CD14", "LYZ", "S100A8", "S100A9"),
  "Classical_Mono" = c("CD14", "S100A8", "FCGR3A"),
  "Non_Classical_Mono" = c("FCGR3A", "MS4A7"),
  "Dendritic_cells" = c("FCER1A", "CST3"),
  "Plasmacytoid_DC" = c("IL3RA", "GZMB", "SERPINF1"),
  "Platelets" = c("PPBP", "PF4", "GP9")
)

# Check which markers are present in dataset
all_genes <- rownames(merged_seurat_integrated)

for (cell_type in names(canonical_markers)) {
  markers <- canonical_markers[[cell_type]]
  present <- markers %in% all_genes

  cat(sprintf("%-20s: %d/%d markers present\n",
              cell_type, sum(present), length(markers)))

  if (!all(present)) {
    missing <- markers[!present]
    cat(sprintf("  Missing: %s\n", paste(missing, collapse = ", ")))
  }
}
```

If a gene is entirely absent from `all_genes` (rather than just absent from
`scale.data`), it was filtered out upstream during QC gene filtering
(walkthrough 2) - `GetResidual()` above cannot help there, and that cell
type will need to rely more heavily on its remaining markers or on the
automated methods below.

### Step 2: visualizing canonical markers

```{r step6_visualize_markers}

#-----------------------------------------------
# STEP 6: Visualize canonical markers
#-----------------------------------------------

# Function to create violin plots for a marker set
plot_violin_markers <- function(markers, cell_type_name, filename_suffix) {
  present_markers <- markers[markers %in% rownames(merged_seurat_integrated)]

  if (length(present_markers) == 0) {
    cat("  No markers present for", cell_type_name, "\n")
    return(NULL)
  }

  cat(" ", cell_type_name, ":", length(present_markers), "markers\n")

  p <- VlnPlot(merged_seurat_integrated,
               features = present_markers,
               group.by = "seurat_clusters",
               pt.size = 0,
               ncol = 3) &
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_blank(),
          plot.title = element_text(size = 11, face = "bold"))

  n_rows <- ceiling(length(present_markers) / 3)
  height <- max(4, n_rows * 2.5)

  ggsave(file.path(annot.dir,paste0("plots/manual_annotation/violin_", filename_suffix, ".png")),
         p, width = 14, height = height, dpi = 300)

  return(p)
}

# Function to create UMAP feature plots for a marker set
plot_umap_markers <- function(markers, cell_type_name, filename_suffix) {
  present_markers <- markers[markers %in% rownames(merged_seurat_integrated)]

  if (length(present_markers) == 0) {
    return(NULL)
  }

  p <- FeaturePlot(merged_seurat_integrated,
                   features = present_markers,
                   reduction = chosen_reduction,
                   ncol = 4,
                   pt.size = 0.1) &
    theme(plot.title = element_text(size = 10),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.position = "right")

  n_rows <- ceiling(length(present_markers) / 4)
  height <- max(4, n_rows * 3)

  ggsave(file.path(annot.dir,paste0("plots/manual_annotation/umap_", filename_suffix, ".png")),
         p, width = 16, height = height, dpi = 300)

  return(p)
}

# Generate plots for ALL cell types
for (cell_type_key in names(canonical_markers)) {
  markers <- canonical_markers[[cell_type_key]]
  cell_type_name <- gsub("_", " ", cell_type_key)

  plot_violin_markers(markers, cell_type_name, cell_type_key)
  plot_umap_markers(markers, cell_type_name, cell_type_key)
}

# Create comprehensive DotPlot showing all key markers
all_markers <- unique(unlist(canonical_markers))
all_markers_present <- all_markers[all_markers %in% rownames(merged_seurat_integrated)]

p_dotplot <- DotPlot(merged_seurat_integrated,
                     features = all_markers_present,
                     group.by = "seurat_clusters") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        axis.text.y = element_text(size = 8)) +
  labs(title = "All Canonical Markers by Cluster",
       subtitle = paste(length(all_markers_present), "markers across all PBMC cell types"))

ggsave(file.path(annot.dir,"plots/manual_annotation/dotplot_all_canonical_markers.png"), p_dotplot,
       width = 12, height = max(8, length(all_markers_present) * 0.15), dpi = 300)

p_dotplot
```

Reading the violin plots: a wide/tall violin with values well above 1-2
indicates high, consistent expression in that cluster; a thin violin
hugging zero indicates the gene is essentially absent. A marker is
cluster-specific when only one or a few clusters show clear elevation.

Reading the DotPlot: dot size encodes the percentage of cells in a cluster
expressing the gene; dot color encodes average expression level. Look for
*combinations* of markers that uniquely define a cluster - single genes are
rarely sufficient (e.g. `CD4` alone doesn't separate CD4+ T cells from some
myeloid populations).

### Step 3: data-driven marker discovery

`FindAllMarkers` performs differential expression of each cluster against
all remaining cells, which is a useful complement to (not a replacement
for) canonical markers - it can reveal markers you didn't think to check.

Because `merged_seurat_integrated` was created by running `SCTransform`
across multiple sample layers (walkthrough 3, Part 1), the `SCT` assay
holds separate correction models per sample. `PrepSCTFindMarkers()`
re-corrects counts to a common sequencing depth so that differential
expression p-values and fold-changes are comparable across samples - this
step is required before `FindAllMarkers` on an `SCT` assay and is *not*
part of the standard-`NormalizeData` workflow the original tutorial
assumes.

```{r step7_find_markers}

#-----------------------------------------------
# STEP 7: Find cluster-specific marker genes
#-----------------------------------------------

# Recorrect SCT counts to a common sequencing depth across samples before DE
merged_seurat_integrated <- PrepSCTFindMarkers(merged_seurat_integrated, verbose = FALSE)

# Find markers for each cluster vs all other cells
cluster_markers <- FindAllMarkers(
  merged_seurat_integrated,
  assay = "SCT",
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

# Remove ribosomal and mitochondrial genes - these commonly top DE lists
# but reflect cell quality/metabolic state rather than cell identity
cluster_markers_filtered <- cluster_markers %>%
  filter(!grepl("^RP[SL]", gene)) %>%
  filter(!grepl("^MT-", gene))

# Get top 5 markers per cluster
top_markers <- cluster_markers_filtered %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC) %>%
  arrange(cluster, desc(avg_log2FC))

cat("\nTop 5 markers per cluster:\n")
print(top_markers %>% select(cluster, gene, avg_log2FC, pct.1, pct.2))

# Save all markers
write.csv(cluster_markers_filtered,
          file.path(annot.dir,"annotations/cluster_markers_all.csv"),
          row.names = FALSE)

write.csv(top_markers,
          file.path(annot.dir,"annotations/top5_markers_per_cluster.csv"),
          row.names = FALSE)

# Create heatmap of top markers
top_genes <- top_markers$gene

p_heatmap <- DoHeatmap(
  merged_seurat_integrated,
  features = top_genes,
  group.by = "seurat_clusters",
  size = 3
) +
  theme(axis.text.y = element_text(size = 6)) +
  labs(title = "Top 5 Markers per Cluster")

ggsave(file.path(annot.dir,"plots/manual_annotation/06_heatmap_top_markers.png"), p_heatmap,
       width = 12, height = 14, dpi = 300)

p_heatmap
```

Interpreting the `FindAllMarkers` output: `avg_log2FC` is higher when a gene
is more specific to that cluster; `pct.1` is the percentage of cells *in*
the cluster expressing the gene; `pct.2` is the percentage of cells *outside*
the cluster expressing it. A good marker has high `pct.1`, low `pct.2`, and
a large positive `avg_log2FC`. `min.pct = 0.25` requires the gene be
detected in at least 25% of cells in the cluster before being tested at all.

### Step 4: manual cluster assignment

The mapping below is an **example** built from the violin plots, DotPlot,
and `FindAllMarkers` results above, following this process: examine marker
expression per cluster -> cross-check against `FindAllMarkers` -> find the
combination of markers that most uniquely identifies each cluster ->
merge clusters that share the same identifying combination. **Customize
this mapping for your own cluster numbers and marker patterns** - it will
not transfer directly to a different clustering resolution or dataset.

```{r step8_manual_assignment}

#-----------------------------------------------
# STEP 8: Manual cell type assignment
#-----------------------------------------------

# EXAMPLE MAPPING - CUSTOMIZE FOR YOUR DATA
cluster_to_celltype <- c(
  "0" = "CD4+ T cells",
  "1" = "CD4+ T cells",
  "2" = "NK cells",
  "3" = "NK cells",
  "4" = "CD4+ T cells",
  "5" = "NK cells",
  "6" = "Classical Mono",
  "7" = "B Cells",
  "8" = "CD8+ T Cells",
  "9" = "B Cells",
  "10" = "CD4+ T cells",
  "11" = "CD4+ T cells",
  "12" = "Monocytes",
  "13" = "CD4+ T cells",
  "14" = "NK cells",
  "15" = "NK cells",
  "16" = "Platelets"
)

# Apply to all cells
cluster_ids <- as.character(merged_seurat_integrated$seurat_clusters)
merged_seurat_integrated$manual_annotation <- unname(cluster_to_celltype[cluster_ids])

# Visualize manual annotation
p_manual <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "manual_annotation",
  label = TRUE,
  label.size = 4,
  pt.size = 0.1,
  repel = TRUE
) +
  ggtitle("Manual Cell Type Annotation") +
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave(file.path(annot.dir,"plots/manual_annotation/07_manual_annotation_umap.png"), p_manual,
       width = 10, height = 8, dpi = 300)

# Split by condition
p_manual_split <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "manual_annotation",
  split.by = "condition",
  pt.size = 0.1,
  ncol = 2
) +
  ggtitle("Manual Annotation by Condition")

ggsave(file.path(annot.dir,"plots/manual_annotation/08_manual_annotation_by_condition.png"),
       p_manual_split, width = 14, height = 6, dpi = 300)

p_manual
```

A simple decision framework for cross-checking assignments, using a
handful of markers per row:

| Cluster | CD3D | CD4 | CD8A | CD14 | CD79A | NKG7 | Assignment |
|---|---|---|---|---|---|---|---|
| 0 | High | High | Low | Low | Low | Low | CD4+ T cells |
| 1 | High | Low | High | Low | Low | Low | CD8+ T cells |
| 2 | Low | Low | Low | Low | High | Low | B cells |
| 3 | Low | Low | Low | High | Low | Low | Monocytes |
| 4 | Low | Low | Low | Low | Low | High | NK cells |

## SingleR: reference-based annotation

### Understanding SingleR's algorithm

For each cell, SingleR calculates its correlation with every cell type in a
reference dataset, then fine-tunes by re-scoring using only genes variable
within the top-scoring types. It reports both a predicted label and a
confidence ("delta") score - the difference between the best and
second-best match - which flags ambiguous cells.

### Available reference datasets (`celldex`)

| Reference | Cell Types | Source | Best For |
|---|---|---|---|
| HumanPrimaryCellAtlas | 38 types | Microarray | Broad tissue coverage |
| MonacoImmuneData | 29 immune types | RNA-seq | Immune cells (PBMCs) |
| DatabaseImmuneCellExpression | 15 immune types | Microarray | Basic immune |
| BlueprintEncodeData | 43 types | RNA-seq | Mixed tissues |
| NovershternHematopoieticData | 38 hematopoietic | Microarray | Blood differentiation |

For PBMC data, `MonacoImmuneData` is typically the best fit - it's
immune-specific and generated with RNA-seq, matching our own technology.

### Testing multiple references

```{r step9_singler}

#-----------------------------------------------
# STEP 9: SingleR with multiple reference datasets
#-----------------------------------------------

# Convert Seurat to SingleCellExperiment for SingleR
sce <- as.SingleCellExperiment(merged_seurat_integrated, assay = "SCT")

# Reference 1: HumanPrimaryCellAtlas (Broad)
hpca_ref <- celldex::HumanPrimaryCellAtlasData()

singler_hpca <- SingleR(
  test = sce,
  ref = hpca_ref,
  labels = hpca_ref$label.main,
  assay.type.test = "logcounts"
)

merged_seurat_integrated$singler_hpca <- singler_hpca$labels

# Reference 2: MonacoImmuneData (Immune-Specific)
monaco_ref <- celldex::MonacoImmuneData()

singler_monaco <- SingleR(
  test = sce,
  ref = monaco_ref,
  labels = monaco_ref$label.main,
  assay.type.test = "logcounts"
)

merged_seurat_integrated$singler_monaco <- singler_monaco$labels

# Reference 3: DatabaseImmuneCellExpression
dice_ref <- celldex::DatabaseImmuneCellExpressionData()

singler_dice <- SingleR(
  test = sce,
  ref = dice_ref,
  labels = dice_ref$label.main,
  assay.type.test = "logcounts"
)

merged_seurat_integrated$singler_dice <- singler_dice$labels
```

HPCA covers 38 broad cell types but includes many non-immune types that can
cause misclassification in a pure-PBMC sample. Monaco is immune-specific
(29 types) and RNA-seq-based, usually the best match here. DICE is simpler
(15 basic immune categories) - less granular but a good sanity check.

### Comparing references

```{r step10_compare_references, fig.width = 16, fig.height = 12}

#-----------------------------------------------
# STEP 10: Compare SingleR references
#-----------------------------------------------

p_hpca <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
                  group.by = "singler_hpca", pt.size = 0.1) +
  labs(title = "HPCA Reference") +
  theme(legend.text = element_text(size = 7))

p_monaco <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
                    group.by = "singler_monaco", pt.size = 0.1) +
  labs(title = "Monaco Reference") +
  theme(legend.text = element_text(size = 7))

p_dice <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
                  group.by = "singler_dice", pt.size = 0.1) +
  labs(title = "DICE Reference") +
  theme(legend.text = element_text(size = 7))

p_manual_ref <- DimPlot(merged_seurat_integrated, reduction = chosen_reduction,
                        group.by = "manual_annotation", pt.size = 0.1) +
  labs(title = "Manual") +
  theme(legend.text = element_text(size = 7))

(p_ref_comparison <- (p_manual_ref | p_hpca) / (p_monaco | p_dice))

ggsave(file.path(annot.dir,"plots/automated_annotation/10_singler_reference_comparison.png"),
       p_ref_comparison, width = 16, height = 12, dpi = 300)
```

Evaluate references by: how closely they resemble the manual annotation,
whether the assigned names make biological sense for blood, granularity
(too many rare labels suggests over-classification; too few suggests
under-classification), and spatial coherence on the UMAP (a good reference
gives compact, contiguous regions rather than scattered "salt-and-pepper"
assignments).

```{r step10b_choose_reference}
# For this tutorial, we'll use Monaco as the primary SingleR annotation
merged_seurat_integrated$singler_annotation <- merged_seurat_integrated$singler_monaco

# Visualize chosen reference
p_singler_final <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "singler_annotation",
  label = TRUE,
  label.size = 3,
  pt.size = 0.1,
  repel = TRUE
) +
  ggtitle("SingleR Annotation (Monaco Reference)") +
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave(file.path(annot.dir,"plots/automated_annotation/11_singler_final_annotation.png"),
       p_singler_final, width = 11, height = 8, dpi = 300)

p_singler_final
```

## scType: marker-based gene set scoring

scType scores each cluster's enrichment for known positive markers (minus
depletion of known negative markers) using a curated marker database. It
is transparent - you can always see exactly which markers drove a call.

### Loading scType functions

The scType project distributes two small R scripts rather than an
installable package.

```{r step11_sctype_functions}

#-----------------------------------------------
# STEP 11: Load scType functions
#-----------------------------------------------

# Load scType functions from GitHub
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```

### Loading marker database

```{r step12_sctype_db}

#-----------------------------------------------
# STEP 12: Load scType marker database
#-----------------------------------------------

# Download the full database
db_url <- "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_full.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
download.file(db_url, destfile = temp_file, mode = "wb")

# Read the database for immune system
db <- read.xlsx(temp_file)
db_immune <- db[db$tissueType == "Immune system", ]

# Prepare gene sets
gs_list <- gene_sets_prepare(temp_file, "Immune system")

cat("Available cell types in database:\n")
print(names(gs_list$gs_positive))
```

The database provides, per tissue, both positive markers (should be
expressed) and negative markers (should *not* be expressed - useful for
distinguishing similar types). We use the "Immune system" tissue subset,
which covers common PBMC populations and subtypes.

### Calculating scType scores

```{r step13_sctype_scores}

#-----------------------------------------------
# STEP 13: Calculate scType scores
#-----------------------------------------------

# Get scaled data (SCT assay; residuals for any missing canonical markers
# were already computed above via GetResidual())
scaled_data <- as.matrix(LayerData(merged_seurat_integrated, assay = "SCT", layer = "scale.data"))

# Calculate scores
es_max <- sctype_score(
  scRNAseqData = scaled_data,
  scaled = TRUE,
  gs = gs_list$gs_positive,
  gs2 = gs_list$gs_negative
)

# Assign cell types to clusters
clusters <- merged_seurat_integrated$seurat_clusters
cL_results <- do.call("rbind", lapply(unique(clusters), function(cl) {
  es_max_cl <- sort(rowSums(es_max[, clusters == cl]), decreasing = TRUE)

  top_score <- es_max_cl[1]
  top_type <- names(es_max_cl)[1]

  data.frame(
    cluster = cl,
    type = top_type,
    scores = top_score,
    ncells = sum(clusters == cl)
  )
}))

# Filter low-confidence assignments - clusters scoring below the 60th
# percentile of all cluster scores are relabeled "Unknown" rather than
# forced into a possibly-wrong category
score_threshold <- quantile(cL_results$scores, 0.6)
cL_results$type[cL_results$scores < score_threshold] <- "Unknown"

# Create mapping and apply to cells
cluster_to_sctype <- setNames(cL_results$type, cL_results$cluster)
merged_seurat_integrated$sctype_annotation <- unname(cluster_to_sctype[
  as.character(merged_seurat_integrated$seurat_clusters)
])

# Visualize
p_sctype <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "sctype_annotation",
  label = TRUE,
  label.size = 4,
  pt.size = 0.1,
  repel = TRUE
) +
  ggtitle("scType: Marker-Based Annotation") +
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave(file.path(annot.dir,"plots/automated_annotation/12_sctype_annotation.png"), p_sctype,
       width = 11, height = 8, dpi = 300)

# Save scType results
write.csv(cL_results, file.path(annot.dir,"annotations/sctype_cluster_scores.csv"), row.names = FALSE)

p_sctype
```

A positive score means strong enrichment for that type's positive markers
(net of negative-marker depletion); a low or negative score means weak or
absent evidence. "Unknown" clusters may represent cell types absent from
the scType database, mixed populations that need finer clustering
resolution, low-quality cells that should have been caught during QC, or
genuinely transitional/activation states.

## scCATCH: tissue-specific automated annotation

scCATCH ("Single Cell Cluster-based Annotation Toolkit for Cellular
Heterogeneity") automatically identifies cluster marker genes and matches
them against a tissue-specific reference database (`cellmatch`), so no
manual marker curation is required.

### Running scCATCH

```{r step14_sccatch}

#-----------------------------------------------
# STEP 14: scCATCH annotation
#-----------------------------------------------

# Prepare data for scCATCH (SCT normalized data layer)
obj_sccatch <- createscCATCH(
  data = LayerData(merged_seurat_integrated, assay = "SCT", layer = "data"),
  cluster = as.character(merged_seurat_integrated$seurat_clusters)
)

# Find marker genes for each cluster
obj_sccatch <- findmarkergene(
  object = obj_sccatch,
  species = "Human",
  marker = cellmatch,
  tissue = "Blood",
  cancer = "Normal",
  cell_min_pct = 0.25,
  logfc = 0.25,
  pvalue = 0.05
)

# Find cell types based on markers
obj_sccatch <- findcelltype(obj_sccatch)

# Extract results
sccatch_celltype <- obj_sccatch@celltype

# Create cluster to cell type mapping
cluster_to_sccatch <- setNames(
  sccatch_celltype$cell_type,
  sccatch_celltype$cluster
)

# Apply to all cells
merged_seurat_integrated$sccatch_annotation <- unname(cluster_to_sccatch[
  as.character(merged_seurat_integrated$seurat_clusters)
])

# Handle unmapped clusters
merged_seurat_integrated$sccatch_annotation[
  is.na(merged_seurat_integrated$sccatch_annotation)
] <- "Unknown"

# Visualize
p_sccatch <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "sccatch_annotation",
  label = TRUE,
  label.size = 4,
  pt.size = 0.1,
  repel = TRUE
) +
  ggtitle("scCATCH: Tissue-Specific Database") +
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave(file.path(annot.dir,"plots/automated_annotation/13_sccatch_annotation.png"), p_sccatch,
       width = 11, height = 8, dpi = 300)

# Save detailed results
write.csv(sccatch_celltype,
          file.path(annot.dir,"annotations/sccatch_cluster_assignments.csv"),
          row.names = FALSE)

# Save marker evidence
marker_evidence <- obj_sccatch@markergene
write.csv(marker_evidence,
          file.path(annot.dir,"annotations/sccatch_marker_evidence.csv"),
          row.names = FALSE)

p_sccatch
```

Parameters: `tissue = "Blood"` selects the appropriate reference subset for
PBMCs; `cancer = "Normal"` excludes tumor-associated cell types;
`cell_min_pct = 0.25` and `logfc = 0.25` mirror the `FindAllMarkers`
thresholds used for manual annotation; `pvalue = 0.05` is the significance
cutoff for candidate markers. The `sccatch_marker_evidence.csv` file
records exactly which markers supported each call, which is useful when
validating or disputing an assignment.

## Save intermediate object

All four annotation methods have now been run and stored as metadata
columns (`manual_annotation`, `singler_annotation`, `sctype_annotation`,
`sccatch_annotation`) on `merged_seurat_integrated`. Part 2 of this
walkthrough compares them and builds a consensus.

```{r save_part1_object}
saveRDS(merged_seurat_integrated,
        file.path(annot.dir,"annotations","seurat_annotated_methods.rds"))
```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut4_annotation_part1_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut4_annotation_part1_sessionInfo_summary.txt"))

sessionInfo()
```
