---
title: "ngs101_scRNAseq_walkthrough_4 - Part 2. Compare annotation methods and build consensus"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-4-cell-type-identification/>

Part 1 of this walkthrough produced four independent cell type annotations
for the same clusters: `manual_annotation`, `singler_annotation`,
`sctype_annotation`, and `sccatch_annotation`. This part compares them
systematically (confusion matrices, Sankey diagrams, side-by-side UMAPs),
then builds a manual consensus and saves the final annotated object.

**Why not automate the consensus step?** The four methods use different
naming conventions for the same biology - e.g. "CD4+ T cells" (manual),
"T cells, CD4+" (SingleR), "T cells" (scType), and "CD4 T cell" (scCATCH)
are all the same call, but a program comparing strings would see four
different labels. Reconciling this reliably requires a human read of the
comparison tables produced below.

## Libraries, directories, and Part 1 object

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)

# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(ggalluvial)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(scales)
library(viridis)

# Set working directory
base.dir <- "~/GSE174609_scRNA"
annot.dir <- file.path(base.dir, "cell_type_annotation")

dir.create(file.path(annot.dir,"plots/method_comparison"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(annot.dir,"annotations"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(42)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

```{r step1b_load_part1}
# Object saved at the end of Part 1, carrying all four annotation columns
merged_seurat_integrated <- readRDS(
  file.path(annot.dir,"annotations","seurat_annotated_methods.rds")
)

load(file.path(base.dir,"integration_analysis","metadata","chosen_methods.RData"))

cat("Loaded object with", ncol(merged_seurat_integrated), "cells\n")
cat("Annotation columns present:",
    paste(intersect(c("manual_annotation","singler_annotation",
                       "sctype_annotation","sccatch_annotation"),
                     colnames(merged_seurat_integrated@meta.data)),
          collapse = ", "), "\n")
```

## Comparing annotation methods

### Confusion matrices

**Bug fix vs. the original tutorial**: the source article's
`create_confusion_matrix` function is written without a `return()`
statement or closing brace, so it does not actually work as published. The
version below adds `return(conf_table)` so the function runs correctly.

```{r step15_confusion_matrices}

#-----------------------------------------------
# STEP 15: Generate confusion matrices
#-----------------------------------------------

# Function to create confusion matrix
create_confusion_matrix <- function(method1_labels, method2_labels,
                                   method1_name, method2_name) {
  conf_table <- table(
    Method1 = method1_labels,
    Method2 = method2_labels
  )
  return(conf_table)
}

# Compare each automated method to manual
conf_manual_vs_singler <- create_confusion_matrix(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$singler_annotation,
  "Manual", "SingleR"
)

conf_manual_vs_sctype <- create_confusion_matrix(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$sctype_annotation,
  "Manual", "scType"
)

conf_manual_vs_sccatch <- create_confusion_matrix(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$sccatch_annotation,
  "Manual", "scCATCH"
)

print(conf_manual_vs_singler)
print(conf_manual_vs_sctype)
print(conf_manual_vs_sccatch)

# Save confusion matrices
write.csv(as.data.frame(conf_manual_vs_singler),
          file.path(annot.dir,"annotations/confusion_matrix_manual_vs_singler.csv"),
          row.names = FALSE)
write.csv(as.data.frame(conf_manual_vs_sctype),
          file.path(annot.dir,"annotations/confusion_matrix_manual_vs_sctype.csv"),
          row.names = FALSE)
write.csv(as.data.frame(conf_manual_vs_sccatch),
          file.path(annot.dir,"annotations/confusion_matrix_manual_vs_sccatch.csv"),
          row.names = FALSE)
```

Rows are Method 1 labels, columns are Method 2 labels, and cell values are
cell counts. Perfect agreement between two methods on a given manual label
shows up as nearly all cells falling in a single column; partial agreement
shows a dominant column with some scatter; disagreement shows counts spread
thinly across many columns with no dominant category.

### Sankey diagrams

```{r step16_sankey_diagrams, fig.width = 12, fig.height = 10}

#-----------------------------------------------
# STEP 16: Create Sankey diagrams
#-----------------------------------------------

# Function to prepare Sankey data
prepare_sankey_data <- function(method1_labels, method2_labels,
                               method1_name, method2_name) {
  sankey_df <- data.frame(
    Method1 = method1_labels,
    Method2 = method2_labels
  ) %>%
    group_by(Method1, Method2) %>%
    summarise(Count = n(), .groups = "drop") %>%
    mutate(Comparison = paste(method1_name, "vs", method2_name))

  return(sankey_df)
}

# Prepare data for each comparison
sankey_manual_singler <- prepare_sankey_data(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$singler_annotation,
  "Manual", "SingleR"
)

sankey_manual_sctype <- prepare_sankey_data(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$sctype_annotation,
  "Manual", "scType"
)

sankey_manual_sccatch <- prepare_sankey_data(
  merged_seurat_integrated$manual_annotation,
  merged_seurat_integrated$sccatch_annotation,
  "Manual", "scCATCH"
)

# Function to create Sankey plot
create_sankey_plot <- function(sankey_data, method1_name, method2_name) {
  p <- ggplot(sankey_data,
              aes(axis1 = Method1, axis2 = Method2, y = Count)) +
    geom_alluvium(aes(fill = Method1), width = 1/12, alpha = 0.7) +
    geom_stratum(width = 1/12, fill = "white", color = "grey") +
    geom_text(stat = "stratum", aes(label = after_stat(stratum)),
              size = 3, fontface = "bold") +
    scale_x_discrete(limits = c(method1_name, method2_name),
                     expand = c(0.05, 0.05)) +
    labs(title = paste(method1_name, "vs", method2_name),
         subtitle = "Straight flows = agreement | Crossed flows = disagreement",
         y = "Number of Cells") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.title.x = element_blank(),
          plot.title = element_text(face = "bold"))

  return(p)
}

# Create Sankey plots
p_sankey_singler <- create_sankey_plot(sankey_manual_singler, "Manual", "SingleR")
ggsave(file.path(annot.dir,"plots/method_comparison/14_sankey_manual_vs_singler.png"),
       p_sankey_singler, width = 12, height = 10, dpi = 300)

p_sankey_sctype <- create_sankey_plot(sankey_manual_sctype, "Manual", "scType")
ggsave(file.path(annot.dir,"plots/method_comparison/15_sankey_manual_vs_sctype.png"),
       p_sankey_sctype, width = 12, height = 10, dpi = 300)

p_sankey_sccatch <- create_sankey_plot(sankey_manual_sccatch, "Manual", "scCATCH")
ggsave(file.path(annot.dir,"plots/method_comparison/16_sankey_manual_vs_sccatch.png"),
       p_sankey_sccatch, width = 12, height = 10, dpi = 300)

p_sankey_singler
p_sankey_sctype
p_sankey_sccatch
```

Straight vertical flows indicate agreement between the two methods; crossed
or diagonal flows indicate disagreement. Flow thickness encodes cell count,
so many thin scattered flows mean widespread disagreement, while a few
thick flows mean strong consensus. Watch particularly for thick flows
landing on "Unknown" or other ambiguous labels - these clusters need
follow-up in the manual comparison workflow below.

### Side-by-side UMAP comparison

```{r step17_umap_comparison, fig.width = 18, fig.height = 14}

#-----------------------------------------------
# STEP 17: Create comprehensive UMAP comparison
#-----------------------------------------------

p_manual_comp <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "manual_annotation",
  pt.size = 0.1
) +
  labs(title = "Manual") +
  theme(legend.text = element_text(size = 7),
        plot.title = element_text(face = "bold"))

p_singler_comp <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "singler_annotation",
  pt.size = 0.1
) +
  labs(title = "SingleR (Monaco)") +
  theme(legend.text = element_text(size = 7),
        plot.title = element_text(face = "bold"))

p_sctype_comp <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "sctype_annotation",
  pt.size = 0.1
) +
  labs(title = "scType (Markers)") +
  theme(legend.text = element_text(size = 7),
        plot.title = element_text(face = "bold"))

p_sccatch_comp <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "sccatch_annotation",
  pt.size = 0.1
) +
  labs(title = "scCATCH (Tissue DB)") +
  theme(legend.text = element_text(size = 7),
        plot.title = element_text(face = "bold"))

# Combine all four methods
(p_all_methods <- (p_manual_comp | p_singler_comp) /
                 (p_sctype_comp | p_sccatch_comp))

ggsave(file.path(annot.dir,"plots/method_comparison/17_all_methods_comparison.png"),
       p_all_methods, width = 18, height = 14, dpi = 300)
```

Look for spatial coherence (contiguous regions rather than scattered
"islands"), consistent boundaries between methods, naming differences that
are really the same biology (e.g. "T cells, CD4+" vs "CD4+ T cells"), and
problem areas - regions where every method disagrees, method-unique labels
that might be errors (or genuine discoveries), and "Unknown" regions that
will need manual curation.

## Building consensus annotations: manual comparison strategy

### Step 1: export annotations for manual review

```{r step18_export_comparison}

#-----------------------------------------------
# STEP 18: Export annotations for manual comparison
#-----------------------------------------------

# Create comprehensive annotation table
annotation_comparison <- data.frame(
  cell_barcode = colnames(merged_seurat_integrated),
  cluster = merged_seurat_integrated$seurat_clusters,
  manual = merged_seurat_integrated$manual_annotation,
  singler = merged_seurat_integrated$singler_annotation,
  sctype = merged_seurat_integrated$sctype_annotation,
  sccatch = merged_seurat_integrated$sccatch_annotation,
  sample_id = merged_seurat_integrated$sample_id,
  condition = merged_seurat_integrated$condition,
  stringsAsFactors = FALSE
)

# Export to CSV for manual review
write.csv(annotation_comparison,
          file.path(annot.dir,"annotations/all_methods_comparison.csv"),
          row.names = FALSE)

# Create cluster-level summary (easier to review)
cluster_summary <- annotation_comparison %>%
  group_by(cluster) %>%
  summarise(
    n_cells = n(),
    manual_type = names(sort(table(manual), decreasing = TRUE))[1],
    singler_type = names(sort(table(singler), decreasing = TRUE))[1],
    sctype_type = names(sort(table(sctype), decreasing = TRUE))[1],
    sccatch_type = names(sort(table(sccatch), decreasing = TRUE))[1],
    .groups = "drop"
  )

write.csv(cluster_summary,
          file.path(annot.dir,"annotations/cluster_level_comparison.csv"),
          row.names = FALSE)

cluster_summary
```

This produces two files: a cell-level table (every cell with all four
method labels, useful for detailed downstream analysis) and a
cluster-level summary (one row per cluster with the modal/dominant label
per method), which is the practical starting point for the manual review
below.

### Step 2: manual comparison workflow

This step has no code - it is a spreadsheet-based decision process, done
outside R:

1. Open `annotations/cluster_level_comparison.csv` in Excel or Google
   Sheets.
2. Add a `consensus_annotation` column for the final decision on each
   cluster.
3. Apply these decision rules per cluster:
   - **All 4 methods agree** (same biology, possibly different naming) ->
     accept the consensus label. Example: Manual "CD4+ T cells", SingleR
     "T cells, CD4+", scType "T cells", scCATCH "CD4 T cell" -> consensus
     "CD4+ T cells".
   - **3 of 4 agree** -> usually accept the majority label, but check
     marker expression (Part 1, Steps 6-7) to confirm before finalizing -
     e.g. if 3 methods say "B cells" and 1 says "Unknown", verify with
     `CD79A`/`MS4A1`.
   - **2 of 4 agree (split decision)** -> requires investigation: return to
     the violin plots and `FindAllMarkers` results, check which markers
     support each candidate assignment, choose the one with the strongest
     evidence, and document the reasoning in a Notes column.
   - **All 4 disagree** -> manual investigation required: the cluster may
     be a mixed population needing higher clustering resolution, a
     transitional state, low-quality cells that should have been filtered
     during QC, or a genuinely novel population needing literature
     research.
4. Save the edited spreadsheet back to CSV (overwriting or renaming
   `cluster_level_comparison.csv`) before continuing to Step 3.

### Step 3: import consensus and finalize

```{r step19_import_consensus}

#-----------------------------------------------
# STEP 19: Import consensus annotations
#-----------------------------------------------

# Read your manually edited cluster consensus file
# (Save your Excel/Google Sheets file as CSV first)
cluster_consensus <- read.csv(file.path(annot.dir,"annotations/cluster_level_comparison.csv"),
                              stringsAsFactors = FALSE)

# Create mapping from cluster to consensus annotation
cluster_to_consensus <- setNames(
  cluster_consensus$consensus_annotation,  # Column you created
  cluster_consensus$cluster
)

# Apply consensus to all cells
merged_seurat_integrated$final_annotation <- unname(cluster_to_consensus[
  as.character(merged_seurat_integrated$seurat_clusters)
])

# Visualize final consensus
p_consensus <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "final_annotation",
  label = TRUE,
  label.size = 4,
  pt.size = 0.1,
  repel = TRUE
) +
  ggtitle("Final Consensus Annotation") +
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave(file.path(annot.dir,"plots/method_comparison/18_final_consensus_annotation.png"),
       p_consensus, width = 11, height = 8, dpi = 300)

# Split by condition to check biological validity
p_consensus_split <- DimPlot(
  merged_seurat_integrated,
  reduction = chosen_reduction,
  group.by = "final_annotation",
  split.by = "condition",
  pt.size = 0.1,
  ncol = 2
) +
  ggtitle("Final Consensus by Condition")

ggsave(file.path(annot.dir,"plots/method_comparison/19_consensus_by_condition.png"),
       p_consensus_split, width = 14, height = 6, dpi = 300)

p_consensus
```

Before moving on, check the final plots for: spatial coherence, plausible
proportions for each cell type (see the composition table in Part 1),
sensible condition-related differences (some cell type proportions are
expected to shift Healthy vs. Post-Treatment), and ideally no remaining
"Unknown" labels.

Note: the code chunk above requires `annotations/cluster_level_comparison.csv`
to already contain a manually completed `consensus_annotation` column (Step
2), so this chunk cannot run unattended in a fully automated pipeline -
it's included here as `eval = TRUE` on the assumption the manual review has
been done between Parts of this walkthrough. Set `eval = FALSE` above if
running the whole document non-interactively before that review step.

## Saving annotated data

```{r step20_save_final}

#-----------------------------------------------
# STEP 20: Save final annotated Seurat object
#-----------------------------------------------

# Save the complete annotated Seurat object
saveRDS(merged_seurat_integrated,
        file.path(annot.dir,"annotations","integrated_annotated_seurat.rds"))

# Create metadata export
metadata_export <- merged_seurat_integrated@meta.data %>%
  select(
    sample_id, condition, patient_id,
    seurat_clusters,
    manual_annotation,
    singler_annotation,
    sctype_annotation,
    sccatch_annotation,
    final_annotation
  )

write.csv(metadata_export,
          file.path(annot.dir,"annotations/cell_metadata_final.csv"),
          row.names = TRUE)
```

This produces `integrated_annotated_seurat.rds` (the complete Seurat object
carrying every annotation column, ready for downstream analysis such as
differential expression by cell type and condition) and
`cell_metadata_final.csv` (a lightweight table suitable for sharing or
depositing alongside a publication). The comparison files saved earlier in
this document (`all_methods_comparison.csv`, `cluster_level_comparison.csv`,
the confusion matrices, and the scType/scCATCH marker evidence files from
Part 1) together document the reasoning behind the final consensus calls.

## Best practices for cell type annotation

**General principles**

1. *Multiple lines of evidence* - never rely on a single marker or method;
   combine automated methods with manual validation, and use both positive
   and negative markers.
2. *Biological context* - account for tissue type, species differences, and
   disease-state effects on marker expression.
3. *Appropriate resolution* - start broad and refine to subtypes only when
   confident; don't over-split without supporting evidence.
4. *Conservative when uncertain* - label ambiguous cells "Unknown" rather
   than guess, and merge unclear distinctions. A broader-but-correct call
   beats a fine-grained-but-wrong one.

**Method selection guidelines**

- **Manual annotation** when domain expertise is available, the tissue or
  condition is novel (no good reference exists), the dataset is small
  (<5,000 cells), or the analysis requires deep biological understanding.
- **Reference-based (SingleR)** when working with a standard, well-atlased
  tissue, a large dataset (>50,000 cells), or when fast/scalable annotation
  is needed and a reference matches the tissue and technology.
- **Marker-based (scType/scCATCH)** for well-characterized tissues without
  an appropriate reference, or when transparent/interpretable assignments
  (seeing exactly which markers drove a call) matter.
- **Multiple methods** whenever publication-quality annotation is needed,
  important biological conclusions depend on the cell type calls, the
  dataset is challenging (mixed states or transitions), or the time budget
  allows thorough validation.

**Reference selection for SingleR**

- PBMCs: best = `MonacoImmuneData`; alternative = `DatabaseImmuneCellExpression`;
  avoid `HumanPrimaryCellAtlasData` (too broad, includes irrelevant
  non-immune types).
- Other tissues: brain -> `HumanPrimaryCellAtlasData` (good neural coverage);
  mixed tissues -> `BlueprintEncodeData`; developmental studies ->
  `NovershternHematopoieticData`.
- Always verify tissue match, technology match (RNA-seq vs. microarray),
  and that expected cell types are actually included in the reference -
  and test more than one reference, as done in Part 1.

**Common pitfalls to avoid**

1. *Over-reliance on automated methods* - algorithms can be confidently
   wrong (e.g. SingleR calling stressed cells "activated"); always validate
   with marker expression.
2. *Wrong reference dataset* - e.g. using a brain reference for blood data;
   match reference tissue and technology to your sample.
3. *Ignoring biological context* - markers can differ between disease and
   health, and between activation/stress states.
4. *Over-splitting cell types* - e.g. 20+ clusters for a tissue with 8-12
   expected populations; merge clusters that share markers and function.
5. *Trusting single markers* - single genes are rarely specific enough;
   require marker combinations.
6. *Not validating rare cell types* - artifacts and doublets can masquerade
   as rare populations; give <1% clusters extra scrutiny.
7. *Automated consensus without manual review* - naming differences between
   methods prevent reliable automated comparison; always reconcile by hand
   as done in this Part.

**Troubleshooting common issues**

- *All markers have extremely high or low values*: usually caused by using
  raw UMI counts instead of normalized/scaled values - this walkthrough
  avoids the issue by working from the SCT assay's `data`/`scale.data`
  layers throughout Part 1.
- *Methods strongly disagree*: return to marker validation, check for
  batch effects or low-quality clusters, consider broader/more conservative
  categories, and try alternative references or marker lists.
- *Automated methods label everything "Unknown"*: check for a
  reference/tissue mismatch, incomplete marker databases, QC that was too
  lenient, or genuinely novel cell types absent from any database.
- *Too many cell types assigned*: usually over-splitting from batch effects,
  an overly granular reference, clustering resolution set too high, or
  doublets creating artificial hybrid types.
- *Expected cell types missing*: check whether QC filtering was too strict,
  whether a rare population got merged into a more abundant one, or whether
  clustering resolution needs to be increased for rare types.

## Conclusion: from clusters to cell types

Four complementary approaches were used to move from numbered clusters to
labeled cell types: manual annotation (biological expertise), SingleR
(reference-based, multiple atlases tested), scType (marker-based gene set
scoring), and scCATCH (tissue-specific automated identification).

**Key takeaways**

1. Agreement across diverse methods (reference-based, marker-based, and
   manual) builds confidence; disagreement flags ambiguous populations that
   need further investigation.
2. Manual validation remains essential - even the best automated methods
   can be wrong, so always cross-check with marker expression and
   biological common sense.
3. Context matters - the same gene can mean different things in different
   tissues, species, or disease states, so choose references and marker
   sets that match the specific context of the data.
4. When uncertain, prefer conservative, broader categories over guessing at
   fine subtypes - subtypes can always be refined later with more evidence.

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut4_annotation_part2_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut4_annotation_part2_sessionInfo_summary.txt"))

sessionInfo()
```
