---
title: "ngs101_scRNAseq_walkthrough_3_QC - integration and clustering"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-3-integration-and-clustering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)
library(glmGamPoi) # speeds up SCTransform
 
# Integration methods
library(harmony)
library(batchelor)
library(SeuratWrappers)
 
# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
 
# Quality metrics and utilities
library(FNN)              # K-nearest neighbor calculations for mixing metrics
library(cluster)          # Silhouette scores for clustering quality
library(reshape2)         # Data reshaping for visualization
 
# Parallel processing
library(future)
options(future.globals.maxSize = 20 * 1024^3)  # Increase to 20GB for large datasets
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
# qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
# if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)
integr.dir <- file.path(base.dir, "integration_analysis")

# Create output directories
dir.create("plots", showWarnings = FALSE)
dir.create("plots/integration_comparison", showWarnings = FALSE)
dir.create("plots/clustering", showWarnings = FALSE)
dir.create("integrated_data", showWarnings = FALSE)
dir.create("metadata", showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Define sample paths and load QC-filtered Seurat objects

The default is to load in the cell- and gene-filtered separate sample objects from the previous walkthrough. However, by defining the `suffix` parameter below, it would be possible to choose between cell-filtered only `_qc_part2_filtered.rds` or cell- and gene-filtered `_qc_final_filtered.rds`.

**This may be important** if all possible genes are desired in the aggregated Seurat object (e.g. to help identify or discover rare cell populations that are only present in some samples in the project).

```{r step2}

suffix <- "_qc_final_filtered.rds" # could also be "_qc_part2_filtered.rds"

# Define sample metadata
sample_metadata <- data.frame(
  sample_names = c(
    "Healthy_1", "Healthy_2", "Healthy_3", "Healthy_4",
    "Post_Patient_1", "Post_Patient_2", "Post_Patient_3", "Post_Patient_4"
  ),
  condition = c(
    rep("Healthy", 4),
    rep("Post_Treatment", 4)
  ),
  patient_id = c(
    "Donor_1", "Donor_2", "Donor_3", "Donor_4",
    "Patient_1", "Patient_2", "Patient_3", "Patient_4"
  )
)

# Load QC-filtered Seurat objects

seurat_list <- lapply(sample_metadata$sample_names, function(sample_name) {
  obj <- readRDS(file.path(filt.dir, paste0(sample_name, suffix)))
 
  # Ensure metadata is consistent
  obj$sample_id <- sample_name
  obj$condition <- sample_metadata$condition[sample_metadata$sample_names == sample_name]
  obj$patient_id <- sample_metadata$patient_id[sample_metadata$sample_names == sample_name]
 
  return(obj)
})

names(seurat_list) <- sample_metadata$sample_names
```

### Ensure all samples have the same genes

This is critical for e.g. FastMNN integration. See note above on bypassing initial gene-level filtering in the QC process for specific purposes.

```{r gene_check}

# Ensure all samples have the same genes (critical for FastMNN integration)
all_genes <- lapply(seurat_list, rownames)
common_genes <- Reduce(intersect, all_genes)
 
# Subset all samples to common genes
seurat_list <- lapply(seurat_list, function(obj) {
  obj[common_genes, ]
})
 
# Report dimensions
cat("\nLoaded", length(seurat_list), "QC-filtered samples:\n")
for (sample_id in names(seurat_list)) {
  cat(sprintf("  %s: %d cells × %d genes\n", 
              sample_id, 
              ncol(seurat_list[[sample_id]]), 
              nrow(seurat_list[[sample_id]])))
}
 
cat("\nTotal cells across all samples:", 
    sum(sapply(seurat_list, ncol)), "\n")
```

## Naive merge without integration

```{r naive_merge}
#-----------------------------------------------
# STEP 3: Naive merge without integration
#-----------------------------------------------
 
# Merge all samples
merged_naive <- merge(
  x = seurat_list[[1]],
  y = seurat_list[-1],
  add.cell.ids = names(seurat_list),
  project = "GSE174609_Naive_Merge"
)
 
cat("Merged dataset:", ncol(merged_naive), "cells ×", nrow(merged_naive), "genes\n")
# Merged dataset: 72649 cells × 18861 genes
 
# Standard Seurat workflow
merged_naive <- NormalizeData(merged_naive, verbose = FALSE)
merged_naive <- FindVariableFeatures(merged_naive, nfeatures = 2000, verbose = FALSE)
merged_naive <- ScaleData(merged_naive, verbose = FALSE)
merged_naive <- RunPCA(merged_naive, npcs = 50, verbose = FALSE)
merged_naive <- RunUMAP(merged_naive, dims = 1:30, reduction = "pca", verbose = FALSE)
merged_naive <- FindNeighbors(merged_naive, dims = 1:30, verbose = FALSE)
merged_naive <- FindClusters(merged_naive, resolution = 0.6, verbose = FALSE)
 
cat("Clustering complete:", length(unique(merged_naive$seurat_clusters)), "clusters identified\n")

```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut3_integration_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut3_integration_sessionInfo_summary.txt"))

sessionInfo()
```




