---
title: "ngs101_scRNAseq_walkthrough_3 - Part 1. Merge and prepare integration"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-3-integration-and-clustering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)
library(BPCells)
library(glmGamPoi) # speeds up SCTransform
 
# Integration methods
library(harmony)
library(batchelor)
library(SeuratWrappers)
 
# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
 
# Quality metrics and utilities
library(FNN)              # K-nearest neighbor calculations for mixing metrics
library(cluster)          # Silhouette scores for clustering quality
library(reshape2)         # Data reshaping for visualization
 
# Parallel processing
library(future)
options(future.globals.maxSize = 40 * 1024^3)  # Increase to 40GB for large datasets - needed during integration steps
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
# qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
# if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)
integr.dir <- file.path(base.dir, "integration_analysis")

# Create output directories
dir.create(file.path(integr.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/integration_comparison"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/clustering"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"integrated_data"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"metadata"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Define sample paths and load QC-filtered Seurat objects

The default is to load in the cell- and gene-filtered separate sample objects from the previous walkthrough. However, by defining the `suffix` parameter below, it would be possible to choose between cell-filtered only `_qc_part2_filtered.rds` or cell- and gene-filtered `_qc_final_filtered.rds`.

**This may be important** if all possible genes are desired in the aggregated Seurat object (e.g. to help identify or discover rare cell populations that are only present in some samples in the project).

```{r step2}

suffix <- "_qc_final_filtered.rds" # could also be "_qc_part2_filtered.rds"

# Define sample metadata
sample_metadata <- data.frame(
  sample_names = c(
    "Healthy_1", "Healthy_2", "Healthy_3", "Healthy_4",
    "Post_Patient_1", "Post_Patient_2", "Post_Patient_3", "Post_Patient_4"
  ),
  condition = c(
    rep("Healthy", 4),
    rep("Post_Treatment", 4)
  ),
  patient_id = c(
    "Donor_1", "Donor_2", "Donor_3", "Donor_4",
    "Patient_1", "Patient_2", "Patient_3", "Patient_4"
  )
)

# Load QC-filtered Seurat objects

seurat_list <- lapply(sample_metadata$sample_names, function(sample_name) {
  cat("Reading in Seurat object",paste0(sample_name,suffix),"\n")
  obj <- readRDS(file.path(filt.dir, paste0(sample_name, suffix)))
 
  # Ensure metadata is consistent
  obj$sample_id <- sample_name
  obj$condition <- sample_metadata$condition[sample_metadata$sample_names == sample_name]
  obj$patient_id <- sample_metadata$patient_id[sample_metadata$sample_names == sample_name]
 
  return(obj)
})

names(seurat_list) <- sample_metadata$sample_names
```

### Ensure all samples have the same genes

This is critical for e.g. FastMNN integration. See note above on bypassing initial gene-level filtering in the QC process for specific purposes.

```{r gene_check}

# Ensure all samples have the same genes (critical for FastMNN integration)
all_genes <- lapply(seurat_list, rownames)
common_genes <- Reduce(intersect, all_genes)
 
# Subset all samples to common genes
seurat_list <- lapply(seurat_list, function(obj) {
  obj[common_genes, ]
})
 
# Report dimensions
cat("\nLoaded", length(seurat_list), "QC-filtered samples:\n")
for (sample_id in names(seurat_list)) {
  cat(sprintf("  %s: %d cells × %d genes\n", 
              sample_id, 
              ncol(seurat_list[[sample_id]]), 
              nrow(seurat_list[[sample_id]])))
}
 
cat("\nTotal cells across all samples:", 
    sum(sapply(seurat_list, ncol)), "\n")
```

## Naive merge without integration, with basic cluster determination

Use `SCTransform` here by default rather than the standard Seurat normalization workflow for basic clustering.

For snRNAseq, consider calculating MALAT1 ratios for each separate sample object in the earlier QC process, and then using MALAT1 ratios as an additional parameter in SCTransform `vars.to.regress`

The `glmGamPoi` package substantially improves speed and is used by default if installed.

**After the merge**: this is the place to convert the object to on-disk via `BPCells` and thus facilitate further operations. The benefit of on-disk storage is apparent with the standard Seurat normalization workflow, but `SCTransform` requires all data in memory (at least for Seurat 5.4; https://github.com/satijalab/sctransform/issues/207)

```{r naive_merge}
#-----------------------------------------------
# STEP 3: Naive merge without integration
#-----------------------------------------------
 
# Merge all samples
merged_naive <- merge(
  x = seurat_list[[1]],
  y = seurat_list[-1],
  add.cell.ids = names(seurat_list),
  project = "GSE174609_Naive_Merge"
)
 
cat("Merged dataset:", ncol(merged_naive), "cells ×", nrow(merged_naive), "genes\n")
# Merged dataset: 72649 cells × 18861 genes
```

### Check whether layers are already split

```{r check_layer_split, eval=FALSE}
# Check if layers are already split
current_layers <- Layers(merged_naive)
if (length(current_layers) > 1) {
  cat("Layers already split by merge operation (", length(current_layers), " layers)\n", sep = "")
} else {
  # Split layers by sample if not already split
  cat("Splitting layers by sample\n")
  merged_naive <- split(merged_naive, f = merged_naive$sample_id)
}
```

### Write layers out with BPCells and normalize

```{r layers_normalization}
for (layer in Layers(merged_naive, search = "counts")) {
  mat <- LayerData(merged_naive, layer = layer)
  path <- file.path(integr.dir,"bpcells_naive_merge", layer)
  write_matrix_dir(mat, path)
  LayerData(merged_naive, layer = layer) <- open_matrix_dir(path)
}

## (Standard workflow commands are commented out)
# merged_naive <- NormalizeData(merged_naive, verbose = FALSE)
# merged_naive <- FindVariableFeatures(merged_naive, nfeatures = 2000, verbose = FALSE)
# merged_naive <- ScaleData(merged_naive, verbose = FALSE)

## SCTransform
merged_naive <- SCTransform(merged_naive,
  vars.to.regress = c("percent.mt"),
  return.only.var.genes = TRUE,
  verbose = TRUE,
  conserve.memory = TRUE
)

merged_naive <- RunPCA(merged_naive, npcs = 50, verbose = FALSE)
merged_naive <- RunUMAP(merged_naive, dims = 1:30, reduction = "pca", verbose = FALSE)
merged_naive <- FindNeighbors(merged_naive, dims = 1:30, verbose = FALSE)
merged_naive <- FindClusters(merged_naive, resolution = 0.6, verbose = FALSE)
 
cat("Clustering complete:", length(unique(merged_naive$seurat_clusters)), "clusters identified with default parameters\n")

saveRDS(merged_naive,file.path(integr.dir,"integrated_data","merged_naive.rds"))
```

### Visualize basic clusters

```{r naive_merge_basic_cluster_viz, fig.width = 16, fig.height = 12}

#-----------------------------------------------
# STEP 4: Visualize batch effects in naive merge
#-----------------------------------------------
 
# Define color palettes for visualization
# 8 samples need 8 distinct, visible colors
sample_colors <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",  # Healthy 1-4: Red, Blue, Green, Purple
  "#FF7F00", "#A65628", "#F781BF", "#999999"   # Post_Patient 1-4: Orange, Brown, Pink, Gray
)
names(sample_colors) <- sample_metadata$sample_id
 
# 2 conditions need 2 distinct colors
condition_colors <- c(
  "Healthy" = "#2E86AB",       # Blue
  "Post_Treatment" = "#F18F01"  # Orange
)
 
# UMAP colored by sample - shows batch effects
p1_naive <- DimPlot(merged_naive, reduction = "umap", group.by = "sample_id", 
                     pt.size = 0.05, cols = sample_colors) +
  ggtitle("Naive Merge: Colored by Sample") +
  theme(legend.position = "right", legend.text = element_text(size = 8))
 
# UMAP colored by condition
p2_naive <- DimPlot(merged_naive, reduction = "umap", group.by = "condition",
                     pt.size = 0.05, cols = condition_colors) +
  ggtitle("Naive Merge: Colored by Condition")
 
# UMAP colored by clusters
p3_naive <- DimPlot(merged_naive, reduction = "umap", group.by = "seurat_clusters",
                     pt.size = 0.05, label = TRUE, label.size = 5) +
  ggtitle("Naive Merge: Clusters") +
  NoLegend()
 
# Split by sample to see separation
p4_naive <- DimPlot(merged_naive, reduction = "umap", group.by = "condition",
                     split.by = "sample_id", pt.size = 0.05, ncol = 4, 
                     cols = condition_colors) +
  ggtitle("Naive Merge: Split by Sample") +
  theme(strip.text = element_text(size = 9, face = "bold"))
 
# Combine plots
(combined_naive <- (p1_naive | p2_naive) / (p3_naive | p4_naive))
ggsave(file.path(integr.dir,"plots/01_naive_merge_batch_effects.png"), combined_naive, 
       width = 16, height = 12, dpi = 300)
```

### Quantify apparent batch effects

The criterion used in this walkthrough for excessive contribution of a single sample to a cluster is >50% of cluster membership.

```{r naive_merge_batch_effects}
#-----------------------------------------------
# STEP 5: Quantify batch effects with mixing metrics
#-----------------------------------------------
 
# Calculate per-cluster sample composition
cluster_composition <- table(merged_naive$seurat_clusters, merged_naive$sample_id)
cluster_composition_pct <- prop.table(cluster_composition, margin = 1) * 100
 
# Find clusters dominated by single samples (>50% from one sample = batch-driven)
dominant_sample_clusters <- apply(cluster_composition_pct, 1, max) > 50
n_batch_clusters <- sum(dominant_sample_clusters)
 
cat("\nBatch effect assessment:\n")
cat("  Clusters dominated by single sample (>50%):", n_batch_clusters, 
    "/", nrow(cluster_composition), "\n")
# Clusters dominated by single sample (>50%): 1 / 19
 
if (n_batch_clusters > 0) {
  cat("    Naive merge shows significant batch effects\n")
  cat("  → Integration is necessary\n")
}
```
## Integration procedures

The `merged_naive` Seurat object has had metadata added in prior steps. It's ready for further steps. A copy will be made as `merged_seurat_subset` and the integration procedures wil be run on this new object.

**Note**: the original tutorial performed standard Seurat normalization (`NormalizeData`, `FindVariableFeatures`, `ScaleData`) followed by `RunPCA` to prepare the object for integration. For this modified walkthough (2026-05) I'm using `SCTransform` followed by PCA reduction.

```{r copy_object}
merged_seurat_subset <- merged_naive
```

Code is presented below to subset the object, sequentially according to biosample (cell donor) so as not to skew the subset toward biosamples with larger cell numbers. Default `eval = FALSE`

### Subsampling code

Run 2026-05-10. The subsetted object would be used to evaluate integration method performance, and then the chosen method would be applied to the original object with all cells.

```{r subset_Seurat_object, eval = TRUE}

sample_ids <- unique(merged_seurat_subset@meta.data$sample_id)
cell_frac <- 0.20

cells_to_keep <- NULL
for (s in seq_along(sample_ids)) {
  tmp_sub <- subset(merged_seurat_subset, subset = sample_id == sample_ids[s])
  tmp_keep <- sample(colnames(tmp_sub), round(ncol(tmp_sub) * cell_frac))
  cells_to_keep <- c(cells_to_keep, tmp_keep)
}
rm(tmp_sub, tmp_keep)

merged_seurat_subset <- subset(merged_seurat_subset, cells = cells_to_keep)
```

### Preserve integration layers within the same object

The original tutorial produced a separate Seurat object for each integration. This should not be necessary, so I've performed each integration within the `merged_seurat_subset` object. The UMAPs for each integration can be retrieved by name.

There are `saveRDS` commands at the end of each integration section to preserve work done in case of crashes. These can be commented out.

### CCA integration

```{r run_cca}

#-----------------------------------------------
# STEP 7: CCA Integration
#-----------------------------------------------
 
# Integrate using CCA
merged_seurat_subset <- IntegrateLayers(
  object = merged_seurat_subset,
  method = CCAIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.cca",
  normalization.method = "SCT",
  dims = 1:30,
  verbose = FALSE
)
 
# Standard downstream workflow
merged_seurat_subset <- FindNeighbors(merged_seurat_subset, reduction = "integrated.cca", dims = 1:30, verbose = FALSE)
merged_seurat_subset <- FindClusters(merged_seurat_subset, resolution = 0.6, cluster.name = "cca_clusters", verbose = FALSE)
merged_seurat_subset <- RunUMAP(merged_seurat_subset, reduction = "integrated.cca", dims = 1:30, 
                          reduction.name = "umap.cca", verbose = FALSE)
 
cat("CCA integration complete:", length(unique(merged_seurat_subset$cca_clusters)), "clusters\n")
# CCA integration complete: 23 clusters

saveRDS(merged_seurat_subset,file.path(integr.dir,"integrated_data","merged_seurat_subset_with_integrations.rds"))
```

### RPCA integration

```{r rpca_integration}
#-----------------------------------------------
# STEP 8: RPCA Integration
#-----------------------------------------------
 
# Integrate using RPCA
merged_seurat_subset <- IntegrateLayers(
  object = merged_seurat_subset,
  method = RPCAIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.rpca",
  normalization.method = "SCT",
  dims = 1:30,
  verbose = FALSE
)
 
# Downstream workflow
merged_seurat_subset <- FindNeighbors(merged_seurat_subset, reduction = "integrated.rpca", dims = 1:30, verbose = FALSE)
merged_seurat_subset <- FindClusters(merged_seurat_subset, resolution = 0.6, cluster.name = "rpca_clusters", verbose = FALSE)
merged_seurat_subset <- RunUMAP(merged_seurat_subset, reduction = "integrated.rpca", dims = 1:30,
                           reduction.name = "umap.rpca", verbose = FALSE)
 
cat("RPCA integration complete:", length(unique(merged_seurat_subset$rpca_clusters)), "clusters\n")
# RPCA integration complete: 22 clusters

saveRDS(merged_seurat_subset,file.path(integr.dir,"integrated_data","merged_seurat_subset_with_integrations.rds"))
```

### Harmony integration

```{r harmony}
# Run Harmony (corrects batch effects on PCA embedding)
# Harmony works directly on PCA embedding

merged_seurat_subset <- RunHarmony(merged_seurat_subset,
                            group.by.vars = "sample_id", # covariates for which harmony will remove effect on data
                            reduction.use = "pca",
                            dims.use = 1:50,
                            reduction.save = "integrated.harmony"
                            )
 
# Downstream workflow
merged_seurat_subset <- FindNeighbors(merged_seurat_subset, reduction = "integrated.harmony", dims = 1:30, verbose = FALSE)
merged_seurat_subset <- FindClusters(merged_seurat_subset, resolution = 0.6, cluster.name = "harmony_clusters", verbose = FALSE)
merged_seurat_subset <- RunUMAP(merged_seurat_subset, reduction = "integrated.harmony", dims = 1:30,
                                reduction.name = "umap.harmony", verbose = FALSE)
 
cat("Harmony integration complete:", length(unique(merged_seurat_subset$harmony_clusters)), "clusters\n")
# Harmony integration complete: 21 clusters

saveRDS(merged_seurat_subset,file.path(integr.dir,"integrated_data","merged_seurat_subset_with_integrations.rds"))
```

### FastMNN integration

Note that unlike the other three methods evaluated here, FastMNN (in the `batchelor` package) also corrects counts, which are placed in a new Seurat Assay (`mnn.reconstructed` by default). This in addition to the embeddings encoded in `integrated.mnn`.

The code chunk below is for FastMNN integration after the **standard normalization** workflow, and with an identical number of genes in each layer established by upstream chunk 'gene check'.

```{r fastmnn_integration_nfs}
#-----------------------------------------------
# STEP 10: FastMNN Integration
#-----------------------------------------------
DefaultAssay(merged_seurat_subset) <- "RNA"

merged_seurat_subset <- IntegrateLayers(
  object = merged_seurat_subset,
  method = FastMNNIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.mnn",
  verbose = FALSE
)
 
cat("FastMNN integration complete\n")
 
# Downstream workflow
merged_seurat_subset <- FindNeighbors(merged_seurat_subset, reduction = "integrated.mnn", dims = 1:30, verbose = FALSE)
merged_seurat_subset <- FindClusters(merged_seurat_subset, resolution = 0.6, cluster.name = "fastmnn_clusters", verbose = FALSE)
merged_seurat_subset <- RunUMAP(merged_seurat_subset, reduction = "integrated.mnn", dims = 1:30,
                              reduction.name = "umap.mnn", verbose = FALSE)
 
cat("Clustering complete:", length(unique(merged_seurat_subset$fastmnn_clusters)), "clusters\n")
# FastMNN integration complete: 20 clusters

DefaultAssay(merged_seurat_subset) <- "SCT" # back to previous state

saveRDS(merged_seurat_subset,file.path(integr.dir,"integrated_data","merged_seurat_subset_with_integrations.rds"))
```

#### Difficulties with FastMNN and SCTransform

`SCTransform` can select different numbers of variable genes for each layer or sample (~= cell donor). This differs from standard normalization, for which there has already been a check to ensure each layer has the same number of genes (see upstream chunk 'gene_check').

The `FastMNN` function from `batchelor` is very strict about reconciling genes across layers or batches. An alternative is illustrated by the **`SeuratIntegrate`** package used in Q3 2025, in which the command `DoIntegrate(MNNIntegration...)` calls the `mnnCorrect` function from `batchelor` under the hood. This handles `SCTransform` Assays gracefully, but is very slow.

Another alternative would be to construct a custom `SeuratWrappers` function to integrate `mnnCorrect` into Seurat's `IntegrateLayers` function.

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut3_integration_part1_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut3_integration_part1_sessionInfo_summary.txt"))

sessionInfo()
```




