---
title: "ngs101_scRNAseq_walkthrough_3 - Part 2. Compare integration methods and select"
subtitle: "Subset from merged naive object to facilitate integration evaluation"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-3-integration-and-clustering>

## Libraries, directories, defaults

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)
library(BPCells)
library(glmGamPoi) # speeds up SCTransform
 
# Integration methods
library(harmony)
library(batchelor)
library(SeuratWrappers)
 
# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
 
# Quality metrics and utilities
library(FNN)              # K-nearest neighbor calculations for mixing metrics
library(cluster)          # Silhouette scores for clustering quality
library(reshape2)         # Data reshaping for visualization
 
# Parallel processing
library(future)
options(future.globals.maxSize = 20 * 1024^3)  # Increase to 40GB for large datasets - needed during integration steps
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
# qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
# if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)
integr.dir <- file.path(base.dir, "integration_analysis")

# Create output directories
dir.create(file.path(integr.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/integration_comparison"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/clustering"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"integrated_data"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"metadata"), showWarnings = FALSE)

# Object file names from Part 1 integration analysis
merged_naive.file <- "merged_naive.rds"
merged_integrations.file <- "merged_seurat_subset_with_integrations.rds"

# Set random seed for reproducibility
set.seed(27)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Load integration comparison objects from Part 1

```{r step2_load_objs}

cat("Reading in merged_naive Seurat object:",merged_naive.file,"\n")
merged_naive <- readRDS(file.path(integr.dir,"integrated_data",merged_naive.file))
cat("Reading in merged Seurat object with integrations:",merged_integrations.file,"\n")
merged_seurat_subset <- readRDS(file.path(integr.dir,"integrated_data",merged_integrations.file))

```

## Visual comparison of UMAPs

```{r umap_comparison, fig.width = 16, fig.height = 12}
#-----------------------------------------------
# STEP 11: Generate comparison UMAPs
#-----------------------------------------------

sample_ids <- unique(merged_seurat_subset@meta.data$sample_id)

# Define color palettes for visualization
# 8 samples need 8 distinct, visible colors
sample_colors <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",  # Healthy 1-4: Red, Blue, Green, Purple
  "#FF7F00", "#A65628", "#F781BF", "#999999"   # Post_Patient 1-4: Orange, Brown, Pink, Gray
)
names(sample_colors) <- sample_ids
 
# 2 conditions need 2 distinct colors
condition_colors <- c(
  "Healthy" = "#2E86AB",       # Blue
  "Post_Treatment" = "#F18F01"  # Orange
)

cat("\n=== Generating Comparison Plots ===\n")
 
# Create a function to make standardized UMAP plots
make_comparison_plot <- function(seurat_obj, reduction_name, title, group_by = "sample_id", colors = NULL) {
  p <- DimPlot(seurat_obj, reduction = reduction_name, group.by = group_by, pt.size = 0.05, shuffle = TRUE) + # shuffle randomizes dot plotting order so last plotted group doesn't overlay others
    ggtitle(title) +
    theme(plot.title = element_text(face = "bold", size = 12),
          legend.text = element_text(size = 7),
          legend.key.size = unit(0.3, "cm"))
 
  # Apply custom colors if provided
  if (!is.null(colors)) {
    p <- p + scale_color_manual(values = colors)
  }
 
  return(p)
}
 
# UMAPs colored by sample (assesses mixing)
p_sample_naive <- make_comparison_plot(merged_naive, "umap", "Naive Merge", 
                                       colors = sample_colors)
p_sample_cca <- make_comparison_plot(merged_seurat_subset, "umap.cca", "CCA", 
                                     colors = sample_colors)
p_sample_rpca <- make_comparison_plot(merged_seurat_subset, "umap.rpca", "RPCA", 
                                      colors = sample_colors)
p_sample_harmony <- make_comparison_plot(merged_seurat_subset, "umap.harmony", "Harmony", 
                                         colors = sample_colors)
p_sample_mnn <- make_comparison_plot(merged_seurat_subset, "umap.mnn", "FastMNN", 
                                     colors = sample_colors)
 
# Combine sample-colored UMAPs
(combined_samples <- (p_sample_naive | p_sample_cca | p_sample_rpca) / 
                    (p_sample_harmony | p_sample_mnn | plot_spacer()))
 
ggsave(file.path(integr.dir,"plots/integration_comparison/02_integration_by_sample_using_subset.png"), 
       combined_samples, width = 16, height = 12, dpi = 300)
 
# UMAPs colored by condition (assesses biological preservation)
p_cond_naive <- make_comparison_plot(merged_naive, "umap", "Naive Merge", "condition", 
                                     colors = condition_colors)
p_cond_cca <- make_comparison_plot(merged_seurat_subset, "umap.cca", "CCA", "condition", 
                                   colors = condition_colors)
p_cond_rpca <- make_comparison_plot(merged_seurat_subset, "umap.rpca", "RPCA", "condition", 
                                    colors = condition_colors)
p_cond_harmony <- make_comparison_plot(merged_seurat_subset, "umap.harmony", "Harmony", "condition", 
                                       colors = condition_colors)
p_cond_mnn <- make_comparison_plot(merged_seurat_subset, "umap.mnn", "FastMNN", "condition", 
                                   colors = condition_colors)
 
# Combine condition-colored UMAPs
(combined_conditions <- (p_cond_naive | p_cond_cca | p_cond_rpca) /
                       (p_cond_harmony | p_cond_mnn | plot_spacer()))
 
ggsave(file.path(integr.dir,"plots/integration_comparison/03_integration_by_condition_using_subset.png"),
       combined_conditions, width = 16, height = 12, dpi = 300)
```

## Quantitative integration metrics

### Mixing metric (LISI)

**How the mixing metric works**: For each cell, we find its k=50 nearest neighbors in the UMAP space and calculate the diversity of samples among those neighbors using the inverse Simpson’s Index. If all neighbors are from different samples (perfect mixing), the score is high. If all neighbors are from the same sample (poor integration), the score is low.

k can be configured below

```{r lisi, fig.width = 10, fig.height = 6}
#-----------------------------------------------
# STEP 12: Calculate integration quality metrics
#-----------------------------------------------

k_param <- 50

# Function to calculate mixing metric (local inverse Simpson's Index)
# This measures sample diversity in each cell's k-nearest neighborhood
calculate_mixing_metric <- function(seurat_obj, group_by = "sample_id", reduction = "umap", k = k_param) {
  # Get embedding
  embedding <- Embeddings(seurat_obj, reduction = reduction)
 
  # Calculate k-nearest neighbors using FNN package
  nn_result <- FNN::get.knn(embedding[, 1:2], k = k)
 
  # For each cell, calculate diversity of samples in its neighborhood
  group_labels <- seurat_obj@meta.data[[group_by]]
  mixing_scores <- sapply(1:nrow(nn_result$nn.index), function(i) {
    neighbors <- nn_result$nn.index[i, ]
    neighbor_groups <- group_labels[neighbors]
    props <- table(neighbor_groups) / length(neighbor_groups)
    # Inverse Simpson's Index (higher = more mixed)
    1 / sum(props^2)
  })
 
  return(mean(mixing_scores))
}
 
# Calculate for all methods
methods_list <- list(
  "Naive" = list(obj = merged_naive, reduction = "umap"),
  "CCA" = list(obj = merged_seurat_subset, reduction = "umap.cca"),
  "RPCA" = list(obj = merged_seurat_subset, reduction = "umap.rpca"),
  "Harmony" = list(obj = merged_seurat_subset, reduction = "umap.harmony"),
  "FastMNN" = list(obj = merged_seurat_subset, reduction = "umap.mnn")
)
 
# Calculate mixing metrics
mixing_results <- data.frame(
  method = names(methods_list),
  mixing_score = sapply(methods_list, function(x) {
    calculate_mixing_metric(x$obj, group_by = "sample_id", reduction = x$reduction)
  }),
  n_clusters = sapply(methods_list, function(x) {
    length(unique(x$obj$seurat_clusters))
  })
)
 
# Display results
cat("\nIntegration Quality Metrics:\n")
cat("(Higher mixing score = better sample mixing)\n\n")
print(mixing_results)
 
# Save metrics
write.csv(mixing_results, "metadata/integration_comparison_metrics_using_subset.csv", row.names = FALSE)
 
# Visualize mixing scores
p_mixing <- ggplot(mixing_results, aes(x = reorder(method, -mixing_score), y = mixing_score, fill = method)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(mixing_score, 2)), vjust = -0.5) +
  labs(title = "Integration Method Comparison: Sample Mixing",
       subtitle = "Higher score = better integration (samples mix within cell types)",
       x = "Integration Method",
       y = "Mixing Score (Local Inverse Simpson's Index)") +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")

print(p_mixing)
ggsave("plots/integration_comparison/04_mixing_scores_using_subset.png", p_mixing, 
       width = 10, height = 6, dpi = 300)
```

### Biological signal (treatment group) preservation

Find condition centers: Calculate the average UMAP position for all Healthy cells and all Post_Treatment cells

Measure separation: Calculate the distance between these two centers

Interpret the distance:

High distance = Conditions are well-separated (biological signal preserved) ✓

Low distance = Conditions overlap (might have over-integrated, losing biology) ⚠️

```{r signal_preservation}
#-----------------------------------------------
# STEP 13: Assess biological signal preservation
#-----------------------------------------------
 
# Function to test if conditions are still distinguishable after integration
# Uses UMAP distances between conditions
assess_condition_separation <- function(seurat_obj, reduction) {
  umap_coords <- seurat_obj[[reduction]]@cell.embeddings[, 1:2]
  conditions <- seurat_obj$condition
 
  # Calculate mean UMAP position for each condition
  condition_centers <- aggregate(umap_coords, by = list(conditions), FUN = mean)
 
  # Calculate pairwise distances between condition centers
  dist_matrix <- dist(condition_centers[, -1])
  mean_dist <- mean(dist_matrix)
 
  return(mean_dist)
}
 
# Calculate for all methods
separation_results <- data.frame(
  method = names(methods_list),
  condition_separation = sapply(methods_list, function(x) {
    assess_condition_separation(x$obj, x$reduction)
  })
)
 
mixing_results$condition_separation <- separation_results$condition_separation
 
cat("\nBiological Signal Preservation:\n")
cat("(Higher separation = conditions remain distinguishable)\n\n")
print(mixing_results[, c("method", "mixing_score", "condition_separation", "n_clusters")])
```

### Plot tradeoff

Ideal integration: High mixing + Moderate separation

Over-integration: High mixing + Low separation (conditions lost)

Under-integration: Low mixing + High separation (batches not corrected)

```{r mix_preserve_plot, fig.width = 10, fig.height=7}

# Visualize the trade-off
p_tradeoff <- ggplot(mixing_results, aes(x = mixing_score, y = condition_separation, label = method)) +
  geom_point(size = 4, aes(color = method)) +
  geom_text_repel(size = 4, box.padding = 0.5) +
  labs(title = "Integration Quality: Mixing vs Biological Preservation",
       subtitle = "Ideal: Upper right (high mixing + preserved biology)",
       x = "Sample Mixing Score (higher = better integration)",
       y = "Condition Separation (higher = preserved biology)") +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Set2")

print(p_tradeoff)
ggsave("plots/integration_comparison/05_mixing_vs_preservation_using_subset.png", p_tradeoff,
       width = 10, height = 7, dpi = 300)
```

## Choose between evaluated integration methods

The primary ranking method below is the `mixing_score` (batch correction) rather than condition separation.

```{r integration_choice}
#-----------------------------------------------
# STEP 14: Select optimal integration method
#-----------------------------------------------
 
# Rank methods by mixing score (primary criterion)
mixing_results <- mixing_results[order(-mixing_results$mixing_score), ]
 
cat("\nMethod ranking (by sample mixing):\n")
for (i in 1:nrow(mixing_results)) {
  cat(sprintf("  %d. %s (mixing: %.2f, separation: %.2f)\n",
              i, 
              mixing_results$method[i],
              mixing_results$mixing_score[i],
              mixing_results$condition_separation[i]))
}
 
# Select top method
chosen_method <- mixing_results$method[1]
cat("\nRecommended method:", chosen_method, "\n")
```
*Review metrics above* before imnplementing reduction choice in the following code chunks (set to `eval = FALSE` by default)

```{r confirm_chosen_reduction, eval = FALSE}
# For subsequent analysis, we'll use the best method
# In most cases, this will be Harmony or CCA

## Option for manual override
# chosen_method <- "enter_here"

if (chosen_method == "Harmony") {
  chosen_reduction <- "umap.harmony"
} else if (chosen_method == "CCA") {
  chosen_reduction <- "umap.cca"
} else if (chosen_method == "RPCA") {
  chosen_reduction <- "umap.rpca"
} else if (chosen_method == "FastMNN") {
  chosen_reduction <- "umap.mnn"
} else if (chosen_method == "Naive") {
  chosen_reduction <- "umap"
} else {
  # Default to CCA if method not found
  chosen_reduction <- "umap.cca"
  cat("Warning: Method not recognized, defaulting to CCA\n")
}
 
cat("\nUsing", chosen_method, "for downstream clustering analysis\n")
```

### Implement chosen integration method on full merged_naive object

```{r integrate_merged_naive_object, eval = FALSE}
# Implement the chosen method only
cat("Implementing chosen integration method on merged_naive\n")
if (chosen_method == "CCA") {
  merged_seurat_integrated <- IntegrateLayers(
    object = merged_naive,
    method = CCAIntegration,
    orig.reduction = "pca",
    new.reduction = "integrated.cca",
    normalization.method = "SCT",
    dims = 1:30,
    verbose = TRUW
  )
  
  # Standard downstream workflow
  merged_seurat_integrated <- FindNeighbors(
    merged_seurat_integrated,
    reduction = "integrated.cca",
    dims = 1:30,
    verbose = TRUE
  )
  merged_seurat_integrated <- FindClusters(
    merged_seurat_integrated,
    resolution = 0.6,
    cluster.name = "cca_clusters",
    verbose = TRUE
  )
  merged_seurat_integrated <- RunUMAP(
    merged_seurat_integrated,
    reduction = "integrated.cca",
    dims = 1:30,
    reduction.name = "umap.cca",
    verbose = TRUE
  )
  
  cat("CCA integration complete:", length(unique(
    merged_seurat_integrated$cca_clusters
  )), "clusters\n")
} else if (chosen_method == "RPCA") {
  merged_seurat_integrated <- IntegrateLayers(
    object = merged_naive,
    method = RPCAIntegration,
    orig.reduction = "pca",
    new.reduction = "integrated.rpca",
    normalization.method = "SCT",
    dims = 1:30,
    verbose = TRUE
  )
  
  # Standard downstream workflow
  merged_seurat_integrated <- FindNeighbors(
    merged_seurat_integrated,
    reduction = "integrated.rpca",
    dims = 1:30,
    verbose = TRUE
  )
  merged_seurat_integrated <- FindClusters(
    merged_seurat_integrated,
    resolution = 0.6,
    cluster.name = "rpca_clusters",
    verbose = TRUE
  )
  merged_seurat_integrated <- RunUMAP(
    merged_seurat_integrated,
    reduction = "integrated.rpca",
    dims = 1:30,
    reduction.name = "umap.rpca",
    verbose = TRUE
  )
  
  cat("RPCA integration complete:", length(unique(
    merged_seurat_integrated$rpca_clusters
  )), "clusters\n")
} else if (chosen_method == "Harmony") {
  merged_seurat_integrated <- RunHarmony(merged_naive,
                            group.by.vars = "sample_id", # covariates for which harmony will remove effect on data
                            reduction.use = "pca",
                            dims.use = 1:50,
                            reduction.save = "integrated.harmony"
                            )
  merged_seurat_integrated <- FindNeighbors(merged_seurat_integrated, reduction = "integrated.harmony", dims = 1:30, verbose = TRUE)
merged_seurat_integrated <- FindClusters(merged_seurat_integrated, resolution = 0.6, cluster.name = "harmony_clusters", verbose = TRUE)
merged_seurat_subset <- RunUMAP(merged_seurat_integrated, reduction = "integrated.harmony", dims = 1:30,
                                reduction.name = "umap.harmony", verbose = TRUE)
 
cat("Harmony integration complete:", length(unique(merged_seurat_subset$harmony_clusters)), "clusters\n")
} else if (chosen_method == "FastMNN") {
  DefaultAssay(merged_naive) <- "RNA"
  merged_seurat_integrated <- IntegrateLayers(
    object = merged_naive,
    method = FastMNNIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.mnn",
  verbose = TRUE
  )
  
  # Standard downstream workflow
  merged_seurat_integrated <- FindNeighbors(
    merged_seurat_integrated,
    reduction = "integrated.mnn",
    dims = 1:30,
    verbose = TRUE
  )
  merged_seurat_integrated <- FindClusters(
    merged_seurat_integrated,
    resolution = 0.6,
    cluster.name = "fastmnn_clusters",
    verbose = TRUE
  )
  merged_seurat_integrated <- RunUMAP(
    merged_seurat_integrated,
    reduction = "integrated.mnn",
    dims = 1:30,
    reduction.name = "umap.mnn",
    verbose = TRUE
  )
  
  cat("FastMNN integration complete:", length(unique(
    merged_seurat_integrated$fastmnn_clusters
  )), "clusters\n")
} else if (chosen_method == "Naive") {
  merged_seurat_integrated <- merged_naive
}

# Save out integrated object
cat("Saving Seurat object with chosen integration\n")
saveRDS(merged_seurat_integrated, file.path(integr.dir,"integrated_data","merged_seurat_chosen_integration.rds"))
save(chosen_method,
     chosen_reduction,
     file.path(integr.dir,"metadata","chosen_methods.RData"))
```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut3_integration_part2_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut3_integration_part2_sessionInfo_summary.txt"))

sessionInfo()
```




