---
title: "ngs101_scRNAseq_walkthrough_3 - Part 3. Define optimal clustering resolution and summarize integration"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-3-integration-and-clustering>

## Libraries, directories, previous files, defaults

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)
library(BPCells)
library(glmGamPoi) # speeds up SCTransform
 
# Integration methods
library(harmony)
library(batchelor)
library(SeuratWrappers)

# Silhouette scoring
library(cluster)
 
# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
 
# Quality metrics and utilities
library(FNN)              # K-nearest neighbor calculations for mixing metrics
library(cluster)          # Silhouette scores for clustering quality
library(reshape2)         # Data reshaping for visualization
 
# Parallel processing
library(future)
options(future.globals.maxSize = 20 * 1024^3)  # Increase to 20GB for datasets with large objects
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
# qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
# if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)
integr.dir <- file.path(base.dir, "integration_analysis")

# Create output directories
dir.create(file.path(integr.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/integration_comparison"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"plots/clustering"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"integrated_data"), showWarnings = FALSE)
dir.create(file.path(integr.dir,"metadata"), showWarnings = FALSE)

# Object file names from Parts 1-2 integration analysis
merged_naive.file <- "merged_naive.rds"

## Use _with_integrations if integrations were evaluated on the entire merged_naive object
# merged_integrations.file <- "merged_seurat_with_integrations.rds"

## Use _chosen_integration if integrations were evaluated on a subset of the merged_naive object and a single integration was implemented on the merged_naive object
merged_integrations.file <- "merged_seurat_chosen_integration.rds"

# Set random seed for reproducibility
set.seed(27)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Load integration comparison objects from Parts 1-2

```{r step2_load_objs}

# cat("Reading in merged_naive Seurat object:",merged_naive.file,"\n")
# merged_naive <- readRDS(file.path(integr.dir,"integrated_data",merged_naive.file))

cat("Reading in merged Seurat object with integration:",merged_integrations.file,"\n")
merged_seurat_integrated <- readRDS(file.path(integr.dir,"integrated_data",merged_integrations.file))

load(file.path(integr.dir,"metadata","chosen_methods.RData"))
```

## Examine clustering at multiple resolutions

If using `RNA` assay rather than `SCT`, change `_snn_res` label in cluster calculator and renaming function below

```{r multiple_resolution_clusters, fig.width = 14, fig.height = 14}

#-----------------------------------------------
# STEP 15: Perform clustering at multiple resolutions
#-----------------------------------------------
 
# Test multiple resolutions
resolutions <- c(0.4, 0.6, 0.8, 1.0, 1.2)
 
for (res in resolutions) {
  merged_seurat_integrated <- FindClusters(
    merged_seurat_integrated,
    resolution = res,
    verbose = TRUE
  )
 
  n_clusters <- length(unique(merged_seurat_integrated@meta.data[[paste0("SCT_snn_res.", res)]]))
  cat(sprintf("Resolution %.1f: %d clusters\n", res, n_clusters))
}
 
# Rename cluster columns for clarity
colnames(merged_seurat_integrated@meta.data) <- gsub("SCT_snn_res\\.", "clusters_res_", 
                                              colnames(merged_seurat_integrated@meta.data))

#-----------------------------------------------
# STEP 16: Visualize clustering results across resolutions
#-----------------------------------------------
 
# Create UMAP plots for each resolution
plot_list <- lapply(resolutions, function(res) {
  cluster_col <- paste0("clusters_res_", res)
  n_clusters <- length(unique(merged_seurat_integrated@meta.data[[cluster_col]]))
 
  DimPlot(merged_seurat_integrated, 
          reduction = chosen_reduction,
          group.by = cluster_col,
          label = TRUE,
          label.size = 4,
          pt.size = 0.3) +
    ggtitle(paste0("Resolution ", res, " (", n_clusters, " clusters)")) +
    NoLegend()
})
 
# Combine plots
(combined_resolutions <- wrap_plots(plot_list, ncol = 2))
ggsave(file.path(integr.dir,"plots/clustering/06_multi_resolution_clustering.png"), 
       combined_resolutions, width = 14, height = 14, dpi = 300)
```

## Silhouette scores

The Silhouette Score for a data point is calculated as (b - a) / max(a, b), where a is the average distance to points in its own cluster and b is the average distance to points in the nearest neighboring cluster. (Copilot)

```{r silhouette_scores}
#-----------------------------------------------
# STEP 17: Evaluate optimal clustering resolution
#-----------------------------------------------

# Silhouette score measures how similar cells are to their own cluster vs other clusters
# Values range from -1 to 1 (higher is better)

# Map UMAP reduction name to corresponding integrated reduction name
# We use integrated space (not UMAP) for silhouette calculation because:
# - Integrated space preserves more information (30+ dimensions vs UMAP's 2)
# - Silhouette scores are more meaningful in the space where clustering was performed
# integrated_reduction <- switch(reduction_final,
#   "umap.cca" = "integrated.cca",
#   "umap.rpca" = "integrated.rpca",
#   "umap.harmony" = "harmony",
#   "umap.mnn" = "integrated.mnn",
#   "umap" = "pca"  # Fallback for naive merge
# )
 
cat("\nEvaluating clustering resolutions using", chosen_reduction, "space\n")
 
# For large datasets (>10,000 cells), subsample for silhouette calculation (use 5000 cells in our case)
# Computing distance matrices for all cells is computationally prohibitive
n_cells <- ncol(merged_seurat_integrated)
max_cells_for_silhouette <- 5000
 
if (n_cells > max_cells_for_silhouette) {
  cat("Dataset has", n_cells, "cells - subsampling", max_cells_for_silhouette, "cells for silhouette calculation in existing clusters\n")
  set.seed(42)
  subsample_idx <- sample(1:n_cells, max_cells_for_silhouette)
} else {
  subsample_idx <- 1:n_cells
}
 
silhouette_scores <- sapply(resolutions, function(res) {
  cluster_col <- paste0("clusters_res_", res)
 
  # Get clusters for subsampled cells
  clusters <- as.numeric(merged_seurat_integrated@meta.data[[cluster_col]][subsample_idx])
 
  # Get integrated coordinates for subsampled cells (use first 30 dimensions)
  coords <- Embeddings(merged_seurat_integrated, reduction = chosen_reduction)[subsample_idx, ]
  coords <- coords[, 1:min(30, ncol(coords))]
 
  # Calculate silhouette (only if we have at least 2 clusters)
  if (length(unique(clusters)) > 1) {
    dist_matrix <- dist(coords)
    sil <- silhouette(clusters, dist_matrix)
    mean(sil[, 3])
  } else {
    NA  # Return NA if only 1 cluster
  }
})
 
# Create resolution comparison table
resolution_comparison <- data.frame(
  resolution = resolutions,
  n_clusters = sapply(resolutions, function(res) {
    cluster_col <- paste0("clusters_res_", res)
    length(unique(merged_seurat_integrated@meta.data[[cluster_col]]))
  }),
  silhouette_score = silhouette_scores
)
 
cat("\nClustering resolution comparison:\n")
print(resolution_comparison)

# Recommend optimal resolution
# Choose resolution with good silhouette score and interpretable cluster number
optimal_idx <- which.max(resolution_comparison$silhouette_score)
optimal_resolution <- resolution_comparison$resolution[optimal_idx]
 
cat("\nRecommended resolution:", optimal_resolution, 
    "(", resolution_comparison$n_clusters[optimal_idx], "clusters )\n")
 
# Recommended resolution: 0.4 ( 17 clusters )
```

### Set clusters at the chosen resolution as the default for the object

Choose a different value for `optimal_resolution` if desired

```{r set_default_clusters}
# Set the optimal clustering as the default
merged_seurat_integrated$seurat_clusters <- merged_seurat_integrated@meta.data[[paste0("clusters_res_", optimal_resolution)]]
Idents(merged_seurat_integrated) <- "seurat_clusters"
```

## Cluster quality assessment

```{r cluster_quality_assessment, fig.width = 12, fig.height = 6}
#-----------------------------------------------
# STEP 18: Assess cluster quality and stability
#-----------------------------------------------
 
# Check cluster sizes
cluster_sizes <- table(merged_seurat_integrated$seurat_clusters)
cat("\nCluster sizes:\n")
print(cluster_sizes)
 
# Identify very small clusters (<1% of cells)
min_cluster_size <- 0.01 * ncol(merged_seurat_integrated)
small_clusters <- which(cluster_sizes < min_cluster_size)
 
if (length(small_clusters) > 0) {
  cat("\n  Warning: Small clusters detected (<1% of cells):\n")
  print(cluster_sizes[small_clusters])
  cat("   These may represent rare cell types or over-clustering\n")
}
 
# Check sample distribution across clusters
sample_cluster_table <- table(merged_seurat_integrated$seurat_clusters, merged_seurat_integrated$sample_id)
sample_cluster_pct <- prop.table(sample_cluster_table, margin = 1) * 100
 
# Identify sample-specific clusters (>50% from one sample)
sample_specific <- apply(sample_cluster_pct, 1, max) > 50
if (any(sample_specific)) {
  cat("\n  Warning: Sample-dominated clusters detected (>50% from single sample):\n")
  print(names(which(sample_specific)))
  cat("   These may indicate incomplete batch correction\n")
}
 
# Visualize sample distribution in clusters
sample_dist_data <- as.data.frame.matrix(sample_cluster_pct)
sample_dist_data$cluster <- rownames(sample_dist_data)
sample_dist_long <- reshape2::melt(sample_dist_data, id.vars = "cluster",
                                    variable.name = "sample", value.name = "percentage")
 
p_sample_dist <- ggplot(sample_dist_long, aes(x = cluster, y = percentage, fill = sample)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Sample Distribution Across Clusters",
       subtitle = "Check for sample-dominated clusters (poor integration)",
       x = "Cluster", y = "Percentage of Cells") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_sample_dist)
ggsave(file.path(integr.dir,"plots/clustering/08_sample_distribution_clusters.png"), p_sample_dist,
       width = 12, height = 6, dpi = 300)
 
# Check condition distribution
condition_dist <- table(merged_seurat_integrated$seurat_clusters, merged_seurat_integrated$condition)
condition_dist_pct <- prop.table(condition_dist, margin = 1) * 100
 
cat("\nCondition distribution across clusters:\n")
print(round(condition_dist_pct, 1))
 
# Visualize condition distribution
condition_dist_data <- as.data.frame.matrix(condition_dist_pct)
condition_dist_data$cluster <- rownames(condition_dist_data)
condition_dist_long <- reshape2::melt(condition_dist_data, id.vars = "cluster",
                                      variable.name = "condition", value.name = "percentage")
 
p_condition_dist <- ggplot(condition_dist_long, aes(x = cluster, y = percentage, fill = condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Condition Distribution Across Clusters",
       subtitle = "Check if biological conditions are preserved",
       x = "Cluster", y = "Percentage of Cells") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Healthy" = "#2E86AB", 
                                "Post_Treatment" = "#F18F01"))

print(p_condition_dist)
ggsave(file.path(integr.dir,"plots/clustering/09_condition_distribution_clusters.png"), p_condition_dist,
       width = 12, height = 6, dpi = 300)

```

## Final integrated and clustered visualization

```{r summary_plots, fig.width = 16, fig.height = 12}
#-----------------------------------------------
# STEP 19: Create comprehensive visualization of final results
#-----------------------------------------------
 
# Define color palettes for visualization
# 8 samples need 8 distinct, visible colors
sample_colors <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",  # Healthy 1-4: Red, Blue, Green, Purple
  "#FF7F00", "#A65628", "#F781BF", "#999999"   # Post_Patient 1-4: Orange, Brown, Pink, Gray
)
names(sample_colors) <- unique(merged_seurat_integrated@meta.data$sample_ids)
 
# 2 conditions need 2 distinct colors
condition_colors <- c(
  "Healthy" = "#2E86AB",       # Blue
  "Post_Treatment" = "#F18F01"  # Orange
)

# Main UMAP with clusters
p_final_clusters <- DimPlot(merged_seurat_integrated, 
                            reduction = chosen_reduction,
                            group.by = "seurat_clusters",
                            label = TRUE,
                            label.size = 5,
                            pt.size = 0.05) +
  ggtitle(paste0("Final Clustering (", chosen_method, " Integration)")) +
  theme(plot.title = element_text(face = "bold", size = 14))
 
# Split by condition to show preservation
p_by_condition <- DimPlot(merged_seurat_integrated,
                          reduction = chosen_reduction,
                          group.by = "seurat_clusters",
                          split.by = "condition",
                          label = TRUE,
                          label.size = 4,
                          pt.size = 0.05,
                          ncol = 2) +
  ggtitle("Clusters by Treatment Condition") +
  theme(strip.text = element_text(face = "bold"))
 
# Colored by sample (integration quality check)
p_by_sample <- DimPlot(merged_seurat_integrated,
                       reduction = chosen_reduction,
                       group.by = "sample_id",
                       pt.size = 0.05,
                       cols = sample_colors) +
  ggtitle("Sample Mixing (Integration Quality)") +
  theme(legend.text = element_text(size = 7))
 
# Colored by condition
p_by_condition_single <- DimPlot(merged_seurat_integrated,
                                 reduction = chosen_reduction,
                                 group.by = "condition",
                                 pt.size = 0.05,
                                 cols = condition_colors) +
  ggtitle("Treatment Conditions")
 
# Combine final visualizations
(combined_final <- (p_final_clusters | p_by_condition_single) /
                  (p_by_sample | plot_spacer())
)
 
ggsave(file.path(integr.dir,"plots/clustering/10_final_integrated_clustered.png"), combined_final,
       width = 16, height = 12, dpi = 300)
 
# Split by condition detailed view
ggsave(file.path(integr.dir,"plots/clustering/11_clusters_by_condition_split.png"), p_by_condition,
       width = 18, height = 6, dpi = 300)
```

## Save out summary of integrated and clustered data

Ensure `mixing_results` data frame from earlier .Rmd was retrieved at earlier `load(file.path(integr.dir,"metadata","chosen_methods.RData"))` step.

```{r summary_output}
#-----------------------------------------------
# STEP 20: Save integrated and clustered data
#-----------------------------------------------
 
cat("\n=== Saving Final Results ===\n")
 
# Join layers for downstream analysis (or delay until future .Rmd module)
# merged_seurat_integrated <- JoinLayers(merged_seurat_integrated)
 
# Save the final integrated object
saveRDS(merged_seurat_integrated,
        file.path(integr.dir,"integrated_data","seurat_integrated_clustered.rds")
)
 
# Save metadata
write.csv(merged_seurat_integrated@meta.data, 
          file.path(integr.dir,"metadata/integrated_cell_metadata.csv"),
          row.names = TRUE)
 
# Create comprehensive summary
integration_summary <- data.frame(
  dataset = "GSE174609",
  n_samples = length(unique(merged_seurat_integrated$sample_id)),
  n_cells_total = ncol(merged_seurat_integrated),
  n_genes = nrow(merged_seurat_integrated),
  integration_method = chosen_method,
  optimal_resolution = optimal_resolution,
  n_clusters = length(unique(merged_seurat_integrated$seurat_clusters)),
  mixing_score = mixing_results$mixing_score[mixing_results$method == chosen_method],
  condition_separation = mixing_results$condition_separation[mixing_results$method == chosen_method]
)
 
write.csv(integration_summary, file.path(integr.dir,"metadata/integration_summary.csv"), row.names = FALSE)

```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut3_integration_part3_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut3_integration_part3_sessionInfo_summary.txt"))

sessionInfo()
```




