---
title: "ngs101_scRNAseq_walkthrough_3_QC - integration and clustering"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-3-integration-and-clustering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)
library(SeuratObject)
library(glmGamPoi) # speeds up SCTransform
 
# Integration methods
library(harmony)
library(batchelor)
library(SeuratWrappers)
 
# Visualization and data manipulation
library(ggplot2)
library(ggrepel)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
 
# Quality metrics and utilities
library(FNN)              # K-nearest neighbor calculations for mixing metrics
library(cluster)          # Silhouette scores for clustering quality
library(reshape2)         # Data reshaping for visualization
 
# Parallel processing
library(future)
options(future.globals.maxSize = 20 * 1024^3)  # Increase to 20GB for large datasets
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
# qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
# if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)
integr.dir <- file.path(base.dir, "integration_analysis")

# Create output directories
dir.create("plots", showWarnings = FALSE)
dir.create("plots/integration_comparison", showWarnings = FALSE)
dir.create("plots/clustering", showWarnings = FALSE)
dir.create("integrated_data", showWarnings = FALSE)
dir.create("metadata", showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)

# Configure plotting defaults
theme_set(theme_classic(base_size = 12))
```

## Define sample paths and load QC-filtered Seurat objects

The default is to load in the cell- and gene-filtered separate sample objects from the previous walkthrough. However, by defining the `suffix` parameter below, it would be possible to choose between cell-filtered only `_qc_part2_filtered.rds` or cell- and gene-filtered `_qc_final_filtered.rds`.

**This may be important** if all possible genes are desired in the aggregated Seurat object (e.g. to help identify or discover rare cell populations that are only present in some samples in the project).

```{r step2}

suffix <- "_qc_final_filtered.rds" # could also be "_qc_part2_filtered.rds"

# Define sample metadata
sample_metadata <- data.frame(
  sample_names = c(
    "Healthy_1", "Healthy_2", "Healthy_3", "Healthy_4",
    "Post_Patient_1", "Post_Patient_2", "Post_Patient_3", "Post_Patient_4"
  ),
  condition = c(
    rep("Healthy", 4),
    rep("Post_Treatment", 4)
  ),
  patient_id = c(
    "Donor_1", "Donor_2", "Donor_3", "Donor_4",
    "Patient_1", "Patient_2", "Patient_3", "Patient_4"
  )
)

# Load QC-filtered Seurat objects

seurat_list <- lapply(sample_metadata$sample_names, function(sample_name) {
  cat("Reading in Seurat object",paste0(sample_name,suffix),"\n")
  obj <- readRDS(file.path(filt.dir, paste0(sample_name, suffix)))
 
  # Ensure metadata is consistent
  obj$sample_id <- sample_name
  obj$condition <- sample_metadata$condition[sample_metadata$sample_names == sample_name]
  obj$patient_id <- sample_metadata$patient_id[sample_metadata$sample_names == sample_name]
 
  return(obj)
})

names(seurat_list) <- sample_metadata$sample_names
```

### Ensure all samples have the same genes

This is critical for e.g. FastMNN integration. See note above on bypassing initial gene-level filtering in the QC process for specific purposes.

```{r gene_check}

# Ensure all samples have the same genes (critical for FastMNN integration)
all_genes <- lapply(seurat_list, rownames)
common_genes <- Reduce(intersect, all_genes)
 
# Subset all samples to common genes
seurat_list <- lapply(seurat_list, function(obj) {
  obj[common_genes, ]
})
 
# Report dimensions
cat("\nLoaded", length(seurat_list), "QC-filtered samples:\n")
for (sample_id in names(seurat_list)) {
  cat(sprintf("  %s: %d cells × %d genes\n", 
              sample_id, 
              ncol(seurat_list[[sample_id]]), 
              nrow(seurat_list[[sample_id]])))
}
 
cat("\nTotal cells across all samples:", 
    sum(sapply(seurat_list, ncol)), "\n")
```

## Naive merge without integration, with basic cluster determination

Use `SCTransform` here by default rather than the standard Seurat normalization workflow for basic clustering.

For snRNAseq, consider calculating MALAT1 ratios for each separate sample object in the earlier QC process, and then using MALAT1 ratios as an additional parameter in SCTransform `vars.to.regress`

**After the merge**: this is the place to convert the object to on-disk via `BPCells` and thus facilitate further operations. The benefit of on-disk storage is apparent with the standard Seurat normalization workflow, but `SCTransform` requires all data in memory (at least for Seurat 5.4; https://github.com/satijalab/sctransform/issues/207)

```{r naive_merge}
#-----------------------------------------------
# STEP 3: Naive merge without integration
#-----------------------------------------------
 
# Merge all samples
merged_naive <- merge(
  x = seurat_list[[1]],
  y = seurat_list[-1],
  add.cell.ids = names(seurat_list),
  project = "GSE174609_Naive_Merge"
)
 
cat("Merged dataset:", ncol(merged_naive), "cells ×", nrow(merged_naive), "genes\n")
# Merged dataset: 72649 cells × 18861 genes

for (layer in Layers(merged_naive, search = "counts")) {
  mat <- LayerData(merged_naive, layer = layer)
  path <- file.path(integr.dir,"bpcells_naive_merge", layer)
  write_matrix_dir(mat, path)
  LayerData(merged_naive, layer = layer) <- open_matrix_dir(path)
}

## (Standard workflow commands are commented out)
# merged_naive <- NormalizeData(merged_naive, verbose = FALSE)
# merged_naive <- FindVariableFeatures(merged_naive, nfeatures = 2000, verbose = FALSE)
# merged_naive <- ScaleData(merged_naive, verbose = FALSE)

## SCTransform
options(Seurat.object.assay.version = "v5")
merged_naive <- SCTransform(merged_naive,
  vars.to.regress = c("percent.mt"),
  return.only.var.genes = TRUE,
  verbose = TRUE,
  conserve.memory = TRUE
)

# This procedure of exporting SCT layers out as BPCells matrices will only work if SCT was created as a v5 Assay. I'm not sure this is actually possible despite 'options' command above. Following Claude suggestions 2026-04-29
for (layer in Layers(merged_naive, assay="SCT", search = "counts")) {
  mat <- LayerData(merged_naive, assay="SCT", layer = layer)
  path <- file.path(integr.dir,"bpcells_naive_merge_sct_counts", layer)
  write_matrix_dir(mat, path)
  LayerData(merged_naive, assay="SCT", layer = layer) <- open_matrix_dir(path)
}
merged_naive <- RunPCA(merged_naive, npcs = 50, verbose = FALSE)
merged_naive <- RunUMAP(merged_naive, dims = 1:30, reduction = "pca", verbose = FALSE)
merged_naive <- FindNeighbors(merged_naive, dims = 1:30, verbose = FALSE)
merged_naive <- FindClusters(merged_naive, resolution = 0.6, verbose = FALSE)
 
cat("Clustering complete:", length(unique(merged_naive$seurat_clusters)), "clusters identified\n")

```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"tut3_integration_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"tut3_integration_sessionInfo_summary.txt"))

sessionInfo()
```




