---
title: "ngs101_scRNAseq_walkthrough_2_QC - adapted to multiple samples"
subtitle: "part 3 - gene-level QC: choose thresholds, visualize effects"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)

qc_part1.file <- "qc_part1.RData"
load(qc.dir,qc_part1.file)
```

## Cell-Level Quality Control: Visualization (after threshold setting)

### Set thresholds

In the code chunk below, filtering thresholds are given as vectors, whose elements should follow the same order as in the `sample_names` vector.

An alternative would be to construct a table elsewhere with samples as rows and thresholds as columns, import as a data.frame, and then read from each data.frame column. To reduce the need for code rewriting later, the chunk below gathers together manually coded threshold vectors into a `threshold_df` data.frame and then extracts individual threshold values as needed. The alternative approach is thus to import `threshold_df` from .csv etc.

Another reason to specify thresholds in a table is ease of adjustment after re-visualization.

```{r step_7b_1, fig.width = 8, fig.height = 7, eval = FALSE}
# For an example case - adjust if necessary
nfeature_min <- rep(500, 12)    # Minimum genes per cell
nfeature_max <- rep(5000, 12)   # Maximum genes (above this = likely doublets)
ncount_min <- rep(800, 12)      # Minimum UMI per cell
ncount_max <- rep(20000, 12)    # Maximum UMI (above this = likely doublets)
mt_thresh <- rep(10, 12)        # Maximum mitochondrial % (adjust based on tissue)

removal_upper_limit <- rep(30, 12) # upper % limit on cell removal before warning flag
removal_lower_limit <- rep(5, 12)  # lower % limit on cell removal before warning flag

threshold_df <- data.frame(sample_names = sample_names,
                           nfeature_min = nfeature_min,
                           nfeature_max = nfeature_max,
                           ncount_min = ncount_min,
                           ncount_max = ncount_max,
                           mt_thresh = mt_thresh,
                           removal_upper_limit = removal_upper_limit,
                           removal_lower_limit = removal_lower_limit)
```

### Visualize effect of thresholds (loop)

```{r step_7b_2, fig.width = 8, fig.height = 7, eval = FALSE}

qc_list <- list()
removal_pct <- NULL
warn_icon <- '<img draggable="false" role="img" class="emoji" alt="⚠" src="https://s.w.org/images/core/emoji/17.0.2/svg/26a0.svg">'

for (k in seq_along(sample_names)) {
  cat("Chosen thresholds for", sample_names[k], "\n")
  cat(
    "  nFeature_RNA: [",
    threshold_df$nfeature_min[k],
    ",",
    threshold_df$nfeature_max[k],
    "]\n"
  )
  cat("  nCount_RNA: [",
      threshold_df$ncount_min[k],
      ",",
      threshold_df$ncount_max[k],
      "]\n")
  cat("  percent.mt: <", threshold_df$mt_thresh[k], "%\n\n")
  
  cat("Re-reading Seurat object for", sample_names[k])
  seurat_obj <- readRDS(file = file.path(filt.dir, paste0(
    sample_names[k], "_qc_part1_filtered.rds"
  )))
  
  cat("Generating filtering visualizations for",sample_names[k],"\n")
  # Visualize thresholds on data
  
  qc_list[[k]] <- data.frame(
    nCount_RNA = seurat_obj$nCount_RNA,
    nFeature_RNA = seurat_obj$nFeature_RNA,
    percent.mt = seurat_obj$percent.mt
  )
  names(qc_list)[k] <- sample_names[k]
  
  qc_list[[k]]$pass_qc <- (
    qc_list[[k]]$nCount_RNA >= threshold_df$ncount_min[k] &
      qc_list[[k]]$nCount_RNA <= threshold_df$ncount_max[k] &
      qc_list[[k]]$nFeature_RNA >= threshold_df$nfeature_min[k] &
      qc_list[[k]]$nFeature_RNA <= threshold_df$nfeature_max[k] &
      qc_list[[k]]$percent.mt < threshold_df$mt_thresh[k]
  )
  (
    p8 <- ggplot(qc_list[[k]], aes(
      x = log10(nCount_RNA + 1),
      y = log10(nFeature_RNA + 1),
      color = pass_qc
    )) +
      geom_point(alpha = 0.5, size = 1) +
      geom_vline(
        xintercept = log10(c(
          threshold_df$ncount_min[k], threshold_df$ncount_max[k]
        )),
        linetype = "dashed",
        color = "red"
      ) +
      geom_hline(
        yintercept = log10(c(
          threshold_df$nfeature_min[k], threshold_df$nfeature_max[k]
        )),
        linetype = "dashed",
        color = "red"
      ) +
      scale_color_manual(values = c(
        "TRUE" = "#06D6A0", "FALSE" = "#EF476F"
      )) +
      labs(
        title = "Cell Filtering Thresholds",
        subtitle = paste0(
          sum(qc_list[[k]]$pass_qc),
          " cells pass QC (",
          round(sum(qc_list[[k]]$pass_qc) / nrow(qc_list[[k]]) *
                  100, 1),
          "%)"
        ),
        x = "log10(UMI + 1)",
        y = "log10(Genes + 1)",
        color = "Pass QC"
      ) +
      theme_classic()
  )
  
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[k], "_05_filtering_thresholds.png")
    ),
    p8,
    width = 8,
    height = 7,
    dpi = 300
  )
  
  # Report filtering impact
  removal_pct[k] <- sum(!qc_list[[k]]$pass_qc) / nrow(qc_list[[k]]) * 100
  cat("Filtering impact on sample", sample_names[k], "\n")
  cat("  Cells before:", nrow(qc_list[[k]]), "\n")
  cat("  Cells passing QC:", sum(qc_list[[k]]$pass_qc), "\n")
  cat("  Cells removed:",
      sum(!qc_list[[k]]$pass_qc),
      "(",
      round(removal_pct[k], 1),
      "%)\n\n")
  
  # Breakdown by criterion
  cat("Removal breakdown for sample", sample_names[k], "\n")
  cat(
    "  Low genes (<",
    threshold_df$nfeature_min[k],
    "):",
    sum(qc_list[[k]]$nFeature_RNA < threshold_df$nfeature_min[k]),
    "\n"
  )
  cat(
    "  High genes (>",
    threshold_df$nfeature_max[k],
    "):",
    sum(qc_list[[k]]$nFeature_RNA > threshold_df$nfeature_max[k]),
    "\n"
  )
  cat(
    "  High MT% (>",
    threshold_df$mt_thresh[k],
    "%):",
    sum(qc_list[[k]]$percent.mt > threshold_df$mt_thresh[k]),
    "\n\n"
  )
  
  # Sanity check on removal rate
  if (removal_pct[k] > threshold_df$removal_upper_limit[k]) {
    cat(warn_icon,
        "WARNING: Removing",
        round(removal_pct[k], 1),
        "% of cells is high!\n")
    cat("   Consider relaxing thresholds (especially MT%)\n\n")
  } else if (removal_pct[k] < threshold_df$removal_lower_limit[k]) {
    cat(warn_icon,
        "WARNING: Only removing",
        round(removal_pct, 1),
        "% of cells is low.\n")
    cat("   Check if you're retaining low-quality cells.\n\n")
  } else {
    cat("✓ Removal rate of",
        round(removal_pct[k], 1),
        "% is reasonable (typical: 10-25%)\n\n")
  }
}
```

### Save out QC metric objects generated during threshold setting

Outputs gathered into `qc_part2.RData` object

```{r save_qc_part2, eval = TRUE}
qc_part2.file <- "qc_part2.RData"

save(
sample_names,
threshold_df,
qc_list,
removal_pct,
file = file.path(qc.dir,"qc_metrics", qc_part2.file)
)
```

## Re-import previous QC metric files if needed

Default is `eval = FALSE`

```{r re-import_previous_QC_part2_metrics, eval = FALSE}
qc_part2.file <- "qc_part2.RData"
load(file = file.path(qc.dir,"qc_metrics", qc_part2.file))
```

## Cell-Level Quality Control: Implement filtering thresholds

Inspect visualizations and metrics above, and alter any entries in `threshold_df` for `threshold_df_final` in section below

```{r prep_threshold_final, eval = TRUE}

## Make any alterations necessary within threshold_df_final based on visualizing QC parameters
threshold_df_final <- threshold_df

```

### Implement chosen thresholds on Seurat objects

Run this section (set `eval = TRUE`) only after performing above steps

```{r implement_cell_thresholds, eval = FALSE}

cells_before_cell_qc <- NULL

for (m in seq_along(sample_names)) {
  
 # Track cells before cell-level QC filtering for summary
cells_before_cell_qc[m] <- nrow(qc_list[[m]])
 
# Apply filters
cat("Re-reading Seurat object for",sample_names[m],"\n")
  seurat_obj <- readRDS(file = file.path(filt.dir, paste0(
    sample_names[m], "_qc_part1_filtered.rds"
  )))
seurat_obj <- subset(
  seurat_obj,
  subset = nCount_RNA >= threshold_df_final$ncount_min[m] &
           nCount_RNA <= threshold_df_final$ncount_max[m] &
           nFeature_RNA >= threshold_df_final$nfeature_min[m] &
           nFeature_RNA <= threshold_df_final$nfeature_max[m] &
           percent.mt < threshold_df_final$mt_thresh[m]
)
cat("Before filtering:", nrow(qc_list[[m]]), "cells") 
cat("After filtering:", ncol(seurat_obj), "cells remaining\n") 
cat("Saving new qc_part2_filtered Seurat object for",sample_names[m],"\n")
  saveRDS(seurat_obj,
          file = file.path(filt.dir, paste0(
    sample_names[m], "_qc_part2_filtered.rds"
  )))
  rm(seurat_obj) 
}

```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"QC_part2_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"QC_part2_sessionInfo_summary.txt"))

sessionInfo()
```