---
title: "ngs101_scRNAseq_walkthrough_2_QC"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

### Dataset Overview: GSE174609 PBMC Study

We’re working with the GSE174609 dataset, which examines immune cell responses
in periodontitis patients. This study includes 12 samples total.

For this tutorial, we’ll perform QC on one sample: Healthy_1 (SRR14575500) to
demonstrate the complete workflow.

Reference: See Part 1 of this tutorial series for details on processing FASTQ
files through Cell Ranger to generate the count matrices we’ll use here.

### Why PBMCs Are Ideal for Learning

PBMCs are the “gold standard” for single-cell tutorials because they:

✅ Contain well-characterized cell types\
✅ Have clear marker genes (CD3D for T cells, CD14 for monocytes, etc.)\
✅ Show moderate complexity (not too simple, not too complex)\
✅ Are relatively easy to dissociate (good quality data)\
✅ Have abundant reference datasets for comparison

### Key Quality Metrics to Watch

Throughout this tutorial, we’ll monitor:

nCount_RNA: Total UMI counts per cell (expect: 1,000-50,000) nFeature_RNA:
Unique genes detected per cell (expect: 500-5,000) percent.mt: Mitochondrial
gene percentage (expect: \<5% for healthy cells) Doublet rate: Expected \~0.5-2%
for our cell count Ambient RNA contamination: Expected \<5% for well-prepared
PBMCs

## Step 1: Libraries and data

Primary source:
<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

```{r step1_libs_data}
#-----------------------------------------------
# STEP 1: Load libraries and configure environment
#-----------------------------------------------
 
# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
qc.dir <- file.path(base.dir,"qc_analysis")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)
 
# Verify Seurat 5 installation
cat("Seurat version:", as.character(packageVersion("Seurat")), "\n")
cat("R version:", R.version.string, "\n")
cat("Working directory:", getwd(), "\n\n")
 
# Check for Seurat 5 features
if (packageVersion("Seurat") >= "5.0.0") {
  cat("✓ Seurat 5 detected - layer-based architecture available\n")
} else {
  warning("Please upgrade to Seurat 5 for this tutorial")
}
```

## Step 2: Loading and Initial Inspection of 10x Data

Cell Ranger (10x Genomics’ processing pipeline) outputs three key files that
we’ll load:

matrix.mtx.gz: Sparse UMI count matrix features.tsv.gz: Gene names and IDs
barcodes.tsv.gz: Cell barcodes Understanding Your Cell Ranger Output Structure

```         
~/GSE174609_scRNA/cellranger_output/Healthy_1/outs/
├── filtered_feature_bc_matrix/    # Cell Ranger's filtered cells (~3,000-5,000)
│   ├── barcodes.tsv.gz           # Already filtered for empty droplets
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
├── raw_feature_bc_matrix/         # ALL droplets (~20,000-100,000)
│   ├── barcodes.tsv.gz           # Includes empty droplets + cells
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
└── web_summary.html               # Cell Ranger QC report
```

Key difference explained in Part 1:

- **filtered_feature_bc_matrix/**: Cell Ranger already removed empty droplets and
applied basic ambient RNA correction
- **raw_feature_bc_matrix/**: No filtering – all droplets captured during sequencing

Your choice:

**Use raw → Perform Steps 4-5 (EmptyDrops + SoupX) for maximum control**\
**Use filtered → Skip Steps 4-5, trust Cell Ranger’s filtering**\

```{r step2}
#-----------------------------------------------
# STEP 2: Load 10x Genomics data
#-----------------------------------------------

# Define paths
sample_name <- "Healthy_1"
cellranger_output.dir <- file.path(base.dir,paste0("cellranger_output/",sample_name,"/outs"))

# OPTION A: Load RAW matrix (what this tutorial demonstrates)
counts <- Read10X(data.dir = file.path(cellranger_output.dir, "raw_feature_bc_matrix"))
 
# OPTION B: Load FILTERED matrix (alternative - skip Steps 4-5 if using this)
# counts <- Read10X(data.dir = file.path(cellranger_output, "filtered_feature_bc_matrix"))
 
# Create Seurat object
# If raw: contains ALL droplets (~20,000-100,000)
# If filtered: contains only cells Cell Ranger called (~3,000-5,000)
seurat_obj <- CreateSeuratObject(
  counts = counts,
  project = sample_name,
  min.cells = 0,      # Don't filter genes yet
  min.features = 0    # Don't filter cells yet - we'll do this properly
)
 
# Add sample metadata (consistent with Part 1 tutorial)
seurat_obj$sample_id <- "Healthy_1"
seurat_obj$sra_id <- "SRR14575500"
seurat_obj$condition <- "Healthy"
seurat_obj$patient_id <- "Donor_1"
seurat_obj$time_point <- NA  # Not applicable for healthy donors
 
# Quick check
cat("Loaded:", ncol(seurat_obj), "droplets ×", nrow(seurat_obj), "genes\n")
cat("(Most are empty - EmptyDrops will filter in Step 4)\n")
```

## Step 3: Initial Data Exploration

Before filtering, let’s understand the raw data.

```{r step3}
#-----------------------------------------------
# STEP 3: Initial data exploration
#-----------------------------------------------
 
# Basic statistics
cat("Total droplets:", ncol(seurat_obj), "\n")
cat("Total genes:", nrow(seurat_obj), "\n")
 
# Quick statistics
sparsity <- 1 - (sum(LayerData(seurat_obj, layer = "counts") > 0) / 
                 (nrow(seurat_obj) * ncol(seurat_obj)))
cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")
 
# UMI distribution - will be bimodal (empty droplets + real cells)
umi_counts <- colSums(LayerData(seurat_obj, layer = "counts"))
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500), 
    "(likely cells, will be validated by EmptyDrops)\n")

```

## Step 4: Empty Droplet Detection with DropletUtils

Note: If you loaded the filtered matrix in Step 2, skip this step (and Step 5).
Cell Ranger has already performed empty droplet removal. Jump directly to Step 6
(Doublet Detection).

Not every droplet captured by the 10x system contains a real cell. Some contain
only ambient RNA from lysed cells. While Cell Ranger applies basic UMI-threshold
filtering, EmptyDrops uses a more sophisticated statistical approach.

### The Empty Droplet Problem

Three types of droplets:

-   Cell-containing droplets: High UMI counts, diverse gene expression
-   Empty droplets: Low UMI counts, only ambient RNA
-   Damaged cells: Intermediate UMI counts (need filtering later)
-   The challenge: Rare cell types might have low RNA content, so simple UMI
    thresholds can remove real cells.

### Statistical Detection with EmptyDrops

EmptyDrops uses a statistical approach:

-   Estimates the ambient RNA profile from low-count droplets
-   Tests each droplet: “Is this expression profile different from ambient RNA
    alone?”
-   Assigns FDR (false discovery rate) to each droplet
-   Calls droplets as cells if FDR \< 0.01
-   **Note**: FDR for droplet call is a configurable parameter and should be
    extracted from params in yaml frontmatter

```{r step4}
#-----------------------------------------------
# STEP 4: Empty droplet detection
#-----------------------------------------------
 
# Convert Seurat object to SingleCellExperiment for EmptyDrops
# We already loaded the raw matrix in Step 2
sce <- as.SingleCellExperiment(seurat_obj)
 
# Track initial droplet count for QC summary
initial_droplet_count <- ncol(sce)
initial_gene_count <- nrow(sce)
 
# Run EmptyDrops
set.seed(100)
empty_results <- emptyDrops(
  m = counts(sce),
  lower = 100,        # Droplets with <100 UMI assumed empty (for estimating ambient RNA)
  niters = 10000,     # Iterations for p-value calculation
  test.ambient = TRUE # Test if droplets differ from ambient RNA profile
)
 
# Identify cells (FDR < 0.01)
is_cell <- empty_results$FDR < 0.01
is_cell[is.na(is_cell)] <- FALSE  # Treat NA as empty (very low UMI droplets)
validated_barcodes <- colnames(sce)[is_cell]
 
# Brief feedback
cat("Droplets tested:", ncol(sce), "\n")
cat("Cells called:", sum(is_cell), 
    "(", round(sum(is_cell)/ncol(sce)*100, 1), "%)\n")
cat("Empty droplets removed:", sum(!is_cell), 
    "(", round(sum(!is_cell)/ncol(sce)*100, 1), "%)\n")
 
# Visual QC - the critical check!
empty_df <- data.frame(
  total_umi = colSums(counts(sce)),
  is_cell = is_cell
)
 
p1 <- ggplot(empty_df, aes(x = log10(total_umi + 1), fill = is_cell)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_fill_manual(
    values = c("TRUE" = "#2E86AB", "FALSE" = "#A23B72"),
    labels = c("Empty Droplet", "Cell"),
    name = "Classification"
  ) +
  labs(
    title = "EmptyDrops: Cell vs Empty Droplet Detection",
    subtitle = paste(sample_name, sum(is_cell), "cells called from", ncol(sce), "total droplets"),
    x = "log10(UMI + 1)",
    y = "Number of Droplets"
  ) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 14))
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_01_empty_droplets.png")), p1, width = 10, height = 6, dpi = 300)
 
# NOW filter the Seurat object to keep only validated cells
# IMPORTANT: Save raw counts BEFORE filtering (needed for SoupX in Step 5)
raw_counts_all_droplets <- LayerData(seurat_obj, layer = "counts")
 
seurat_obj <- subset(seurat_obj, cells = validated_barcodes)
 
cat("After EmptyDrops:", ncol(seurat_obj), "cells retained\n")
```

## Step 5: Ambient RNA Correction with SoupX

Note: If you loaded the filtered matrix in Step 2, skip this step. Cell Ranger
has already applied basic ambient RNA correction. Jump directly to Step 6
(Doublet Detection).

Even droplets containing cells are contaminated with ambient RNA—the “soup” of
RNA floating in the suspension from lysed cells.

### Why Ambient RNA Matters

The problem:

-   Lysed cells release their RNA into the suspension
-   This ambient RNA enters all droplets
-   Creates false positive gene expression
-   Especially affects highly expressed genes (hemoglobin in blood, myelin in
    brain)

### Impact

-   Inflates marker gene expression
-   Creates false cell-type assignments
-   Reduces specificity of differential expression
-   Estimating and Removing Contamination

Note: This step is optional. First check the contamination level – if it’s NULL
or \<5% (**configurable**), you can skip correction.

```{r step5}
#-----------------------------------------------
# STEP 5: Ambient RNA correction with SoupX
#-----------------------------------------------

contamination_cutoff <- 0.05 # lowpass filter to decide whether to execute SoupX

# Prepare SoupX channel
# tod (table of droplets) = raw counts from ALL droplets (saved in Step 4)
# toc (table of cells) = filtered counts from validated cells only
tod <- raw_counts_all_droplets
toc <- LayerData(seurat_obj, layer = "counts")
sc <- SoupChannel(tod = tod, toc = toc, calcSoupProfile = TRUE)
 
# Quick standard Seurat clustering for contamination estimation; clusters needed by SoupX
temp_obj <- seurat_obj
temp_obj <- NormalizeData(temp_obj, verbose = FALSE)
temp_obj <- FindVariableFeatures(temp_obj, nfeatures = 2000, verbose = FALSE)
temp_obj <- ScaleData(temp_obj, verbose = FALSE)
temp_obj <- RunPCA(temp_obj, npcs = 30, verbose = FALSE)
temp_obj <- FindNeighbors(temp_obj, dims = 1:30, verbose = FALSE)
temp_obj <- FindClusters(temp_obj, resolution = 0.8, verbose = FALSE)
sc <- setClusters(sc, setNames(as.character(temp_obj$seurat_clusters), colnames(temp_obj)))
 
# Estimate contamination
sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed, setting contamination fraction to NULL\n")
  sc$fit$rho <- NULL
  return(sc)
})
 
contamination_fraction <- sc$fit$rho
 
# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample &/or SoupX autoEstCont failure)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")
 
# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction)) {
  cat("→ Contamination is NULL\n")
  cat("   Proceeding with original counts.\n")
} else if (contamination_fraction < contamination_cutoff) {
  cat("→ Contamination is",round(contamination_fraction * 100, 2),"% - SoupX cutoff is",round(contamination_cutoff * 100, 2),"%")
  cat("   Proceeding with original counts.\n")
} else {
  cat("→ Contamination is",round(contamination_fraction * 100, 2),"% - SoupX cutoff is",round(contamination_cutoff * 100, 2),"%")
 
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  seurat_obj <- SetAssayData(seurat_obj, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}

rm(sc,temp_obj,toc,tod)
```

**When to Skip SoupX**: If contamination is NULL (estimation failed) or <5%, your sample is clean and correction is unnecessary. Values >5% indicate ambient RNA contamination worth correcting, especially in tissues prone to cell lysis (brain, tumor).

## Step 6: Doublet Detection with scDblFinder

Doublets occur when two cells are captured in the same droplet. They appear as artificial “intermediate” cell types and must be removed.

### Understanding Doublets

How doublets form:

- 10x captures ~1% doublets per 1,000 cells loaded
- Higher loading = more doublets (10,000 cells ~6 – 9% doublets)
- Completely random—any two cell types can combine

### Why they’re problematic

- Create false “hybrid” cell types
- Inflate cell counts
- Confound trajectory analysis
- Mess up differential expression

```{r step6}
#-----------------------------------------------
# STEP 6: Doublet detection with scDblFinder
#-----------------------------------------------
 
# Join layers (if multiple samples in assay separated out into layers) and convert to SCE
seurat_obj <- JoinLayers(seurat_obj)
sce <- as.SingleCellExperiment(seurat_obj)
 
# Run scDblFinder (suppress expected warnings about empty layers and xgboost)
set.seed(27)
suppressWarnings({
  sce <- scDblFinder(sce, dbr = NULL, verbose = FALSE)
})
 
# Extract results
seurat_obj$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
seurat_obj$doublet_score <- sce$scDblFinder.score
 
# Brief feedback
n_doublets <- sum(seurat_obj$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(seurat_obj) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")
 
# Track cells before doublet removal for QC summary
cells_before_doublet_removal <- ncol(seurat_obj)
 
# Quick UMAP for visualization
seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
seurat_obj <- FindVariableFeatures(seurat_obj, nfeatures = 2000, verbose = FALSE)
seurat_obj <- ScaleData(seurat_obj, verbose = FALSE)
seurat_obj <- RunPCA(seurat_obj, npcs = 30, verbose = FALSE)
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, verbose = FALSE)
 
# Visualize doublets on UMAP
p2 <- DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_02_doublets_umap.png")), p2, width = 8, height = 7, dpi = 300)
 
# Remove doublets
seurat_obj <- subset(seurat_obj, subset = doublet_class == "Singlet")
cat("After doublet removal:", ncol(seurat_obj), "cells\n")
```

**Note on Warnings**: scDblFinder may generate harmless warnings about empty Seurat layers (normal – it only needs raw counts) and xgboost parameter deprecation (package version mismatch). In the code above, these are suppressed with suppressWarnings() for cleaner output but don’t affect doublet detection accuracy.

**Expected Doublet Rates**: For ~3,000-5,000 cells, expect 0.5-2% doublets. Rates >10% suggest problems with cell loading concentration or the doublet detector being too sensitive. Given that our example sample includes over 9,800 cells, a 10% rate is considered acceptable.

## Step 7: Cell-Level Quality Control: Visualization followed by Manual Threshold Setting

Now we filter individual cells based on their quality metrics. The key is visual inspection – look at the distributions and set biologically meaningful thresholds rather than relying on automatic statistical methods.

This is a three-step procedure:

- Visualization of unfiltered cell distributions, including % MT and % ribo genes
- Parameter selection and visualization of impact on individual cells with before & after plots
- Implement chosen filtering parameters

### Note on filtering in this tutorial: Human data

The % MT and % ribo filtering steps below are based on **human** gene symbols. For future workflow development, select according to species.

**For snRNAseq work**, consider visualization of MALAT1 as an additional violin plot.

### Step 7a: Unfiltered cells

```{r step_7a, fig.width = 15, fig.height = 5}
#-----------------------------------------------
# STEP 7: Calculate QC metrics, visualize, and filter cells
#-----------------------------------------------
 
# Calculate QC metrics
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^RP[SL]")
seurat_obj$log10GenesPerUMI <- log10(seurat_obj$nFeature_RNA) / log10(seurat_obj$nCount_RNA)
 
# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions:\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(seurat_obj$nCount_RNA), 
    "| Q1-Q3:", quantile(seurat_obj$nCount_RNA, 0.25), "-", 
    quantile(seurat_obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(seurat_obj$nFeature_RNA),
    "| Q1-Q3:", quantile(seurat_obj$nFeature_RNA, 0.25), "-",
    quantile(seurat_obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(seurat_obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(seurat_obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(seurat_obj$percent.ribo), 2), "%\n")
 
# Visualize distributions - THE CRITICAL STEP
(p3 <- VlnPlot(
  seurat_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) &
  theme(plot.title = element_text(face = "bold"))
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_03_qc_violins.png")), p3, width = 16, height = 4, dpi = 300)
 
# Scatter plots showing metric relationships
p4 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p5 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p6 <- FeatureScatter(seurat_obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
(p_scatter <- p4 + p5 + p6)
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_04_qc_scatter.png")), p_scatter, width = 15, height = 5, dpi = 300)
 
cat("\n→ EXAMINE plots/03_qc_violins.png and plots/04_qc_scatter.png\n")
cat("→ Look for:\n")
cat("   - Bimodal distributions (good vs bad cells)\n")
cat("   - Outliers in scatter plots\n")
cat("   - Relationship between metrics\n\n")

cat("Guidelines for threshold setting:\n")
cat("  • nFeature_RNA min: Where does the low-quality tail end? (typically 500-1000)\n")
cat("  • nFeature_RNA max: Where do potential doublets start? (typically 4000-7000)\n")
cat("  • percent.mt: PBMCs <10%, Brain <5%, Tumor <20%\n")
cat("  • If distributions are bimodal, set threshold between modes\n")
cat("  • Aim to remove no more than 25% of cells \n\n")
```
### Step 7b: Set thresholds in the following step

```{r step_7b, fig.width = 8, fig.height = 7}
# For an example case - adjust if necessary
nfeature_min <- 500    # Minimum genes per cell
nfeature_max <- 5000   # Maximum genes (above this = likely doublets)
ncount_min <- 800      # Minimum UMI per cell
ncount_max <- 20000    # Maximum UMI (above this = likely doublets)
mt_thresh <- 10        # Maximum mitochondrial % (adjust based on tissue)

removal_upper_limit <- 30 # upper % limit on cell removal before warning flag
removal_lower_limit <- 5  # lower % limit on cell removal before warning flag

cat("Chosen thresholds:\n")
cat("  nFeature_RNA: [", nfeature_min, ",", nfeature_max, "]\n")
cat("  nCount_RNA: [", ncount_min, ",", ncount_max, "]\n")
cat("  percent.mt: <", mt_thresh, "%\n\n")

# Visualize thresholds on data
qc_df <- data.frame(
  nCount_RNA = seurat_obj$nCount_RNA,
  nFeature_RNA = seurat_obj$nFeature_RNA,
  percent.mt = seurat_obj$percent.mt
)
 
qc_df$pass_qc <- (
  qc_df$nCount_RNA >= ncount_min &
  qc_df$nCount_RNA <= ncount_max &
  qc_df$nFeature_RNA >= nfeature_min &
  qc_df$nFeature_RNA <= nfeature_max &
  qc_df$percent.mt < mt_thresh
)
 
(p8 <- ggplot(qc_df, aes(x = log10(nCount_RNA + 1), y = log10(nFeature_RNA + 1), 
                        color = pass_qc)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_vline(xintercept = log10(c(ncount_min, ncount_max)), 
             linetype = "dashed", color = "red") +
  geom_hline(yintercept = log10(c(nfeature_min, nfeature_max)), 
             linetype = "dashed", color = "red") +
  scale_color_manual(values = c("TRUE" = "#06D6A0", "FALSE" = "#EF476F")) +
  labs(
    title = "Cell Filtering Thresholds",
    subtitle = paste0(sum(qc_df$pass_qc), " cells pass QC (", 
                      round(sum(qc_df$pass_qc)/nrow(qc_df)*100, 1), "%)"),
    x = "log10(UMI + 1)",
    y = "log10(Genes + 1)",
    color = "Pass QC"
  ) +
  theme_classic()
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_05_filtering_thresholds.png")), p8, width = 8, height = 7, dpi = 300)
 
# Report filtering impact
removal_pct <- sum(!qc_df$pass_qc) / nrow(qc_df) * 100
cat("Filtering impact:\n")
cat("  Cells before:", nrow(qc_df), "\n")
cat("  Cells passing QC:", sum(qc_df$pass_qc), "\n")
cat("  Cells removed:", sum(!qc_df$pass_qc), 
    "(", round(removal_pct, 1), "%)\n\n")
 
# Breakdown by criterion
cat("Removal breakdown:\n")
cat("  Low genes (<", nfeature_min, "):", 
    sum(qc_df$nFeature_RNA < nfeature_min), "\n")
cat("  High genes (>", nfeature_max, "):", 
    sum(qc_df$nFeature_RNA > nfeature_max), "\n")
cat("  High MT% (>", mt_thresh, "%):", 
    sum(qc_df$percent.mt > mt_thresh), "\n\n")
 
# Sanity check on removal rate

warn_icon <- '<img draggable="false" role="img" class="emoji" alt="⚠" src="https://s.w.org/images/core/emoji/17.0.2/svg/26a0.svg">'

if (removal_pct > removal_upper_limit) {
  cat(warn_icon, "WARNING: Removing", round(removal_pct, 1), "% of cells is high!\n")
  cat("   Consider relaxing thresholds (especially MT%)\n\n")
} else if (removal_pct < removal_lower_limit) {
  cat(warn_icon, "WARNING: Only removing", round(removal_pct, 1), "% of cells is low.\n")
  cat("   Check if you're retaining low-quality cells.\n\n")
} else {
  cat("✓ Removal rate of",round(removal_pct,1),"% is reasonable (typical: 10-25%)\n\n")
}
```

### Step 7c: Implement filtering thresholds

```{r step_7c}
# Track cells before cell-level QC filtering for summary
cells_before_cell_qc <- nrow(qc_df)
 
# Apply filters
seurat_obj <- subset(
  seurat_obj,
  subset = nCount_RNA >= ncount_min &
           nCount_RNA <= ncount_max &
           nFeature_RNA >= nfeature_min &
           nFeature_RNA <= nfeature_max &
           percent.mt < mt_thresh
)
 
cat("After filtering:", ncol(seurat_obj), "cells remaining\n")
```

## Step 8: Gene-Level Quality Control

After filtering cells, we filter genes based on how widely they’re detected across cells.

**Why Remove Hemoglobin?** In PBMC samples, hemoglobin genes (HBA1, HBA2, HBB) are typically from contaminating red blood cells during sample preparation, not true expression in PBMCs.

**Detection Threshold**: The percentage-based approach (e.g., 0.1% of cells) scales with dataset size. For 5,000 cells, 0.1% = 5 cells. This is more robust than fixed “3 cells” cutoffs when analyzing datasets of varying sizes.

### Note on filtering in this tutorial: Human data

The % MT, % ribo filtering and % HB steps below are based on **human** gene symbols. For future workflow development, select according to species.

### Step 8a: Visualization first

```{r step8a, fig.width = 8, fig.height = 5}
#-----------------------------------------------
# STEP 8: Gene-level QC with detection threshold
#-----------------------------------------------
 
# Calculate gene detection
counts_matrix <- LayerData(seurat_obj, layer = "counts")
gene_detection <- rowSums(counts_matrix > 0)
 
gene_qc <- data.frame(
  gene = rownames(seurat_obj),
  n_cells_detected = gene_detection,
  pct_cells_detected = (gene_detection / ncol(seurat_obj)) * 100,
  is_mt = grepl("^MT-", rownames(seurat_obj)),
  is_ribo = grepl("^RP[SL]", rownames(seurat_obj)),
  is_hb = grepl("^HB[AB]", rownames(seurat_obj))
)
 
cat("Total genes:", nrow(gene_qc), "\n")
cat("MT genes:", sum(gene_qc$is_mt), 
    "| Ribo genes:", sum(gene_qc$is_ribo), 
    "| Hb genes:", sum(gene_qc$is_hb), "\n\n")
 
# Visualize gene detection
(p9 <- ggplot(gene_qc, aes(x = pct_cells_detected)) +
  geom_histogram(bins = 50, fill = "#118AB2", alpha = 0.7) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Gene Detection Across Cells",
    subtitle = "Red line: 0.1% of cells threshold",
    x = "% of Cells Expressing Gene (log scale)",
    y = "Number of Genes"
  ) +
  theme_classic()
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_06_gene_detection.png")), p9, width = 8, height = 5, dpi = 300)
 
cat("→ EXAMINE plots/06_gene_detection.png\n")
cat("   Choose threshold based on where detection drops off\n\n")
 
# Set filtering threshold - ADJUST BASED ON YOUR DATA
# Common approaches:
#   - 0.1% of cells (very lenient, keeps most genes)
#   - 1% of cells (moderate, standard approach)
#   - 3 cells minimum (absolute count, conservative)

cat("Threshold options:\n")
cat("  • 0.1% of cells: Lenient (keeps rare cell type markers)\n")
cat("  • 1% of cells: Standard (balances detection vs noise)\n")
cat("  • 3-10 cells minimum: Conservative (removes very rare genes)\n\n")
 
```

### Step 8b: Implement gene per % or # cell threshold

```{r step_8b}
min_pct_cells <- 0.1  # Gene must be in ≥0.1% of cells
# Alternatively: min_cells <- 3  # Gene must be in ≥3 cells
 
cat("Setting gene filter threshold:\n")
cat("  Minimum detection: ≥", min_pct_cells, "% of cells\n")
cat("  (Equivalent to ≥", ceiling(ncol(seurat_obj) * min_pct_cells / 100), 
    "cells with current dataset)\n\n")
 
# Filter genes
genes_to_keep <- (gene_qc$pct_cells_detected >= min_pct_cells) & !gene_qc$is_hb
 
cat("Genes passing filter:", sum(genes_to_keep), "/", nrow(gene_qc), "\n")
cat("Genes removed:\n")
cat("  Low detection:", sum(gene_qc$pct_cells_detected < min_pct_cells), "\n")
cat("  Hemoglobin:", sum(gene_qc$is_hb), "\n\n")
 
seurat_obj <- seurat_obj[gene_qc$gene[genes_to_keep], ]
 
cat("After gene filtering:", nrow(seurat_obj), "genes remaining\n")
```

## Step 9 (optional): First Look at Highly Variable Genes

**Variable Features**: These genes show high variability across cells (not just high expression). They’re likely involved in defining cell types and states, making them most useful for downstream clustering and cell type identification (covered in the next tutorial).

```{r step_9, fig.width - 10, fig.height = 7}
#-----------------------------------------------
# STEP 9: Normalization and variable features
#-----------------------------------------------
 
# Log-normalize
seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", 
                            scale.factor = 10000, verbose = FALSE)
 
# Find highly variable genes
seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", 
                                   nfeatures = 2000, verbose = FALSE)
 
top10 <- head(VariableFeatures(seurat_obj), 10)
cat("Top 10 variable genes:", paste(top10, collapse = ", "), "\n")
# Top 10 variable genes: IGLC2, IGKC, LYPD2, LINGO2, IGLC3, PTGDS, FCER1A, IGHM, BANK1, S100A9
 
# Visualize variable features
p10 <- VariableFeaturePlot(seurat_obj)
p10 <- LabelPoints(plot = p10, points = top10, repel = TRUE)

print(p10)
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_07_variable_features.png")), p10, width = 10, height = 7, dpi = 300)


```

## Step 10: Export Filtered Cells and QC Summary

```{r step_10}
#-----------------------------------------------
# STEP 10: Save processed data and create summary
#-----------------------------------------------
 
# Save the clean Seurat object to filtered data directory
saveRDS(seurat_obj, file = file.path(filt.dir,paste0(sample_name,"_qc_filtered.rds")))
 
# Create ONE comprehensive QC summary (instead of many CSV files)
qc_summary <- data.frame(
  sample = sample_name,
 
  # Cell counts through QC pipeline
  initial_droplets = initial_droplet_count,  # From raw matrix before EmptyDrops
  after_emptydrops = length(validated_barcodes),  # Cells called by EmptyDrops
  after_soupx = length(validated_barcodes),       # SoupX doesn't remove cells
  after_doublets = cells_before_doublet_removal - n_doublets,  # After doublet removal
  after_cell_qc = ncol(seurat_obj),  # After cell-level QC filtering
  final_cells = ncol(seurat_obj),    # Final count (same as after_cell_qc)
 
  # Gene counts
  initial_genes = initial_gene_count,  # From raw matrix
  final_genes = nrow(seurat_obj),      # After gene filtering
 
  # Key metrics (from final filtered data)
  median_umi = median(seurat_obj$nCount_RNA),
  median_genes = median(seurat_obj$nFeature_RNA),
  median_mt_pct = median(seurat_obj$percent.mt),
 
  # QC parameters
  contamination_pct = ifelse(is.null(contamination_fraction), 0, 
                             round(contamination_fraction * 100, 2)),
  doublet_rate_pct = round(doublet_rate, 2),
 
  # Filtering thresholds used
  ncount_min = ncount_min,
  ncount_max = ncount_max,
  nfeature_min = nfeature_min,
  nfeature_max = nfeature_max,
  mt_threshold = mt_thresh
)
 
# Save comprehensive summary
write.csv(qc_summary, file.path(qc.dir,paste0("qc_metrics/",sample_name,"QC_summary.csv")), row.names = FALSE)
write.csv(seurat_obj@meta.data, file.path(filt.dir,paste0(sample_name,"_filt_cell_metadata.csv")))
 
cat("\n=== QC Pipeline Complete ===\n")
cat("Final cells:", ncol(seurat_obj), "\n")
cat("Final genes:", nrow(seurat_obj), "\n")
cat("Plots saved:", length(list.files("plots")), "files\n\n")
 
cat("Files created:\n")
cat(paste0("  • ",filt.dir,"/",sample_name,"_qc_filtered.rds - Clean Seurat object\n"))
cat(paste0("  • ",qc.dir,"/qc_metrics/",sample_name,"_QC_summary.csv - Comprehensive QC summary\n"))
cat(paste0("  • ",filt.dir,"/",sample_name,"_filt_cell_metadata.csv - Cell-level metadata\n"))

```