---
title: "ngs101_scRNAseq_walkthrough_2_QC - adapted to multiple samples"
subtitle: "part 1 - empty droplets, ambient RNA, doublet detection"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)
 
# Verify Seurat 5 installation
cat("Seurat version:", as.character(packageVersion("Seurat")), "\n")
cat("R version:", R.version.string, "\n")
```

## Define sample paths and correction parameters

Also define vector objects to receive QC metrics

```{r step2}

# Define paths
sample_names <- c("Healthy_1",
                  "Healthy_2",
                  "Healthy_3",
                  "Healthy_4",
                  "Pre_Patient_1",
                  "Pre_Patient_2",
                  "Pre_Patient_3",
                  "Pre_Patient_4",
                  "Post_Patient_1",
                  "Post_Patient_2",
                  "Post_Patient_3",
                  "Post_Patient_4") # corresponding to cellranger_output/{SAMPLE}/outs

empty_lower <- 100 # 100 UMIs necessary to be counted as non-empty droplet
empty_niters <- 10000 # iterations for EmptyDrops p-value calculation
empty_FDR <- 0.01 # FDR threshold for detecting non-empty droplets

contamination_cutoff <- 0.05 # lowpass filter to decide whether to execute SoupX

initial_droplet_count <- vector(mode = "numeric", length = length(sample_names))
initial_gene_count <- vector(mode = "numeric", length = length(sample_names))
validated_barcode_count <- vector(mode = "numeric", length = length(sample_names))

n_doublets <- vector(mode = "numeric", length = length(sample_names))
doublet_rate <- vector(mode = "numeric", length = length(sample_names))
cells_before_doublet_removal <- vector(mode = "numeric", length = length(sample_names))

p1 <- list() # for collecting empty droplet ggplots
p2 <- list() # for collecting doublet ggplots

```

## Execute loop for empty droplet, ambient RNA and doublet detection

```{r loop1}

for (i in seq_along(sample_names)) {
  cellranger_output.dir <- file.path(base.dir,
                                     paste0("cellranger_output/", sample_names[i], "/outs"))
  counts <- Read10X(data.dir = file.path(cellranger_output.dir, "raw_feature_bc_matrix"))
  seurat_obj <- CreateSeuratObject(
    counts = counts,
    project = sample_names[i],
    min.cells = 0,
    min.features = 0
  )
  seurat_obj$sample_id <- sample_names[i]
  
  ### EmptyDroplet procedures
  sce <- as.SingleCellExperiment(seurat_obj)
  
  # Track initial droplet count for QC summary
  initial_droplet_count[i] <- ncol(sce)
  initial_gene_count[i] <- nrow(sce)
  
  # Run EmptyDrops
  set.seed(27)
  empty_results <- emptyDrops(
    m = counts(sce),
    lower = empty_lower,
    # Droplets with <100 UMI assumed empty (for estimating ambient RNA)
    niters = empty_niters,
    # Iterations for p-value calculation
    test.ambient = TRUE # Test if droplets differ from ambient RNA profile
  )
  
  # Identify cells (FDR < 0.01)
  is_cell <- empty_results$FDR < empty_FDR
  is_cell[is.na(is_cell)] <- FALSE  # Treat NA as empty (very low UMI droplets)
  validated_barcodes <- colnames(sce)[is_cell]
  validated_barcode_count[i] <- length(validated_barcodes)
  
  # Visual QC - the critical check!
  empty_df <- data.frame(total_umi = colSums(counts(sce)), is_cell = is_cell)
  
  p1[[i]] <- ggplot(empty_df, aes(x = log10(total_umi + 1), fill = is_cell)) +
    geom_histogram(bins = 50,
                   alpha = 0.7,
                   position = "identity") +
    scale_fill_manual(
      values = c("TRUE" = "#2E86AB", "FALSE" = "#A23B72"),
      labels = c("Empty Droplet", "Cell"),
      name = "Classification"
    ) +
    labs(
      title = "EmptyDrops: Cell vs Empty Droplet Detection",
      subtitle = paste(
        sample_names[i],
        sum(is_cell),
        "cells called from",
        ncol(sce),
        "total droplets"
      ),
      x = "log10(UMI + 1)",
      y = "Number of Droplets"
    ) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold", size = 14))
  
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[i], "_01_empty_droplets.png")
    ),
    p1[i],
    width = 10,
    height = 6,
    dpi = 300
  )
  
  # Filter the Seurat object to keep only validated cells
  # IMPORTANT: Save raw counts BEFORE filtering (needed for SoupX)
  raw_counts_all_droplets <- LayerData(seurat_obj, layer = "counts")
  
  seurat_obj <- subset(seurat_obj, cells = validated_barcodes)
  rm(sce)
  
  ### SoupX procedures
  # Prepare SoupX channel
  # tod (table of droplets) = raw counts from ALL droplets
  # toc (table of cells) = filtered counts from validated cells only
  tod <- raw_counts_all_droplets
  toc <- LayerData(seurat_obj, layer = "counts")
  sc <- SoupChannel(tod = tod,
                    toc = toc,
                    calcSoupProfile = TRUE)
  
  # Quick standard Seurat clustering for contamination estimation; clusters needed by SoupX
  temp_obj <- seurat_obj
  temp_obj <- NormalizeData(temp_obj, verbose = FALSE)
  temp_obj <- FindVariableFeatures(temp_obj, nfeatures = 2000, verbose = FALSE)
  temp_obj <- ScaleData(temp_obj, verbose = FALSE)
  temp_obj <- RunPCA(temp_obj, npcs = 30, verbose = FALSE)
  temp_obj <- FindNeighbors(temp_obj, dims = 1:30, verbose = FALSE)
  temp_obj <- FindClusters(temp_obj, resolution = 0.8, verbose = FALSE)
  sc <- setClusters(sc, setNames(
    as.character(temp_obj$seurat_clusters),
    colnames(temp_obj)
  ))
  
  # Estimate contamination
  sc <- tryCatch({
    autoEstCont(sc, verbose = FALSE)
  }, error = function(e) {
    sc$fit$rho <- NULL # if autoEstCont fails
    return(sc)
  })
  
  contamination_fraction[i] <- sc$fit$rho
  
  # Decision: Skip or correct based on contamination level
  if (!is.null(contamination_fraction) &&
      contamination_fraction[i] < contamination_cutoff) {
    # Apply correction
    suppressWarnings({
      corrected_counts <- adjustCounts(sc)
    })
    seurat_obj <- SetAssayData(seurat_obj, layer = "counts", new.data = corrected_counts)
  }
  
  rm(sc, temp_obj, toc, tod)
  
  ### Doublet detection procedure (scDblFinder)
  sce <- as.SingleCellExperiment(seurat_obj)
 
set.seed(27)
# Run scDblFinder (suppress expected warnings about empty layers and xgboost)
suppressWarnings({
  sce <- scDblFinder(sce, dbr = NULL, verbose = FALSE)
})
 
# Extract results
seurat_obj$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
seurat_obj$doublet_score <- sce$scDblFinder.score
 
# Brief feedback
n_doublets[i] <- sum(seurat_obj$doublet_class == "Doublet")
doublet_rate[i] <- n_doublets[i] / ncol(seurat_obj) * 100
 
# Track cells before doublet removal for QC summary
cells_before_doublet_removal[i] <- ncol(seurat_obj)
 
# Quick UMAP for visualization
seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
seurat_obj <- FindVariableFeatures(seurat_obj, nfeatures = 2000, verbose = FALSE)
seurat_obj <- ScaleData(seurat_obj, verbose = FALSE)
seurat_obj <- RunPCA(seurat_obj, npcs = 30, verbose = FALSE)
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, verbose = FALSE)
 
# Visualize doublets on UMAP
p2[[i]] <- DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate[i], 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))
 
ggsave(file.path(qc.dir,paste0("plots/",sample_names[i],"_02_doublets_umap.png")), p2, width = 8, height = 7, dpi = 300)
 
# Remove doublets
seurat_obj <- subset(seurat_obj, subset = doublet_class == "Singlet")
}

saveRDS(seurat_obj, file = file.path(filt.dir,paste0(sample_names[i],"_qc_part1_filtered.rds")))
```



## Step 7: Cell-Level Quality Control: Visualization followed by Manual Threshold Setting

Now we filter individual cells based on their quality metrics. The key is visual inspection – look at the distributions and set biologically meaningful thresholds rather than relying on automatic statistical methods.

This is a three-step procedure:

- Visualization of unfiltered cell distributions, including % MT and % ribo genes
- Parameter selection and visualization of impact on individual cells with before & after plots
- Implement chosen filtering parameters

### Note on filtering in this tutorial: Human data

The % MT and % ribo filtering steps below are based on **human** gene symbols. For future workflow development, select according to species.

**For snRNAseq work**, consider visualization of MALAT1 as an additional violin plot.

### Step 7a: Unfiltered cells

```{r step_7a, fig.width = 15, fig.height = 5, eval = FALSE}
#-----------------------------------------------
# STEP 7: Calculate QC metrics, visualize, and filter cells
#-----------------------------------------------
 
# Calculate QC metrics
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^RP[SL]")
seurat_obj$log10GenesPerUMI <- log10(seurat_obj$nFeature_RNA) / log10(seurat_obj$nCount_RNA)
 
# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions:\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(seurat_obj$nCount_RNA), 
    "| Q1-Q3:", quantile(seurat_obj$nCount_RNA, 0.25), "-", 
    quantile(seurat_obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(seurat_obj$nFeature_RNA),
    "| Q1-Q3:", quantile(seurat_obj$nFeature_RNA, 0.25), "-",
    quantile(seurat_obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(seurat_obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(seurat_obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(seurat_obj$percent.ribo), 2), "%\n")
 
# Visualize distributions - THE CRITICAL STEP
(p3 <- VlnPlot(
  seurat_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) &
  theme(plot.title = element_text(face = "bold"))
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_03_qc_violins.png")), p3, width = 16, height = 4, dpi = 300)
 
# Scatter plots showing metric relationships
p4 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p5 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p6 <- FeatureScatter(seurat_obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
(p_scatter <- p4 + p5 + p6)
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_04_qc_scatter.png")), p_scatter, width = 15, height = 5, dpi = 300)
 
cat("\n→ EXAMINE plots/03_qc_violins.png and plots/04_qc_scatter.png\n")
cat("→ Look for:\n")
cat("   - Bimodal distributions (good vs bad cells)\n")
cat("   - Outliers in scatter plots\n")
cat("   - Relationship between metrics\n\n")

cat("Guidelines for threshold setting:\n")
cat("  • nFeature_RNA min: Where does the low-quality tail end? (typically 500-1000)\n")
cat("  • nFeature_RNA max: Where do potential doublets start? (typically 4000-7000)\n")
cat("  • percent.mt: PBMCs <10%, Brain <5%, Tumor <20%\n")
cat("  • If distributions are bimodal, set threshold between modes\n")
cat("  • Aim to remove no more than 25% of cells \n\n")
```
### Step 7b: Set thresholds in the following step

```{r step_7b, fig.width = 8, fig.height = 7, eval = FALSE}
# For an example case - adjust if necessary
nfeature_min <- 500    # Minimum genes per cell
nfeature_max <- 5000   # Maximum genes (above this = likely doublets)
ncount_min <- 800      # Minimum UMI per cell
ncount_max <- 20000    # Maximum UMI (above this = likely doublets)
mt_thresh <- 10        # Maximum mitochondrial % (adjust based on tissue)

removal_upper_limit <- 30 # upper % limit on cell removal before warning flag
removal_lower_limit <- 5  # lower % limit on cell removal before warning flag

cat("Chosen thresholds:\n")
cat("  nFeature_RNA: [", nfeature_min, ",", nfeature_max, "]\n")
cat("  nCount_RNA: [", ncount_min, ",", ncount_max, "]\n")
cat("  percent.mt: <", mt_thresh, "%\n\n")

# Visualize thresholds on data
qc_df <- data.frame(
  nCount_RNA = seurat_obj$nCount_RNA,
  nFeature_RNA = seurat_obj$nFeature_RNA,
  percent.mt = seurat_obj$percent.mt
)
 
qc_df$pass_qc <- (
  qc_df$nCount_RNA >= ncount_min &
  qc_df$nCount_RNA <= ncount_max &
  qc_df$nFeature_RNA >= nfeature_min &
  qc_df$nFeature_RNA <= nfeature_max &
  qc_df$percent.mt < mt_thresh
)
 
(p8 <- ggplot(qc_df, aes(x = log10(nCount_RNA + 1), y = log10(nFeature_RNA + 1), 
                        color = pass_qc)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_vline(xintercept = log10(c(ncount_min, ncount_max)), 
             linetype = "dashed", color = "red") +
  geom_hline(yintercept = log10(c(nfeature_min, nfeature_max)), 
             linetype = "dashed", color = "red") +
  scale_color_manual(values = c("TRUE" = "#06D6A0", "FALSE" = "#EF476F")) +
  labs(
    title = "Cell Filtering Thresholds",
    subtitle = paste0(sum(qc_df$pass_qc), " cells pass QC (", 
                      round(sum(qc_df$pass_qc)/nrow(qc_df)*100, 1), "%)"),
    x = "log10(UMI + 1)",
    y = "log10(Genes + 1)",
    color = "Pass QC"
  ) +
  theme_classic()
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_05_filtering_thresholds.png")), p8, width = 8, height = 7, dpi = 300)
 
# Report filtering impact
removal_pct <- sum(!qc_df$pass_qc) / nrow(qc_df) * 100
cat("Filtering impact:\n")
cat("  Cells before:", nrow(qc_df), "\n")
cat("  Cells passing QC:", sum(qc_df$pass_qc), "\n")
cat("  Cells removed:", sum(!qc_df$pass_qc), 
    "(", round(removal_pct, 1), "%)\n\n")
 
# Breakdown by criterion
cat("Removal breakdown:\n")
cat("  Low genes (<", nfeature_min, "):", 
    sum(qc_df$nFeature_RNA < nfeature_min), "\n")
cat("  High genes (>", nfeature_max, "):", 
    sum(qc_df$nFeature_RNA > nfeature_max), "\n")
cat("  High MT% (>", mt_thresh, "%):", 
    sum(qc_df$percent.mt > mt_thresh), "\n\n")
 
# Sanity check on removal rate

warn_icon <- '<img draggable="false" role="img" class="emoji" alt="⚠" src="https://s.w.org/images/core/emoji/17.0.2/svg/26a0.svg">'

if (removal_pct > removal_upper_limit) {
  cat(warn_icon, "WARNING: Removing", round(removal_pct, 1), "% of cells is high!\n")
  cat("   Consider relaxing thresholds (especially MT%)\n\n")
} else if (removal_pct < removal_lower_limit) {
  cat(warn_icon, "WARNING: Only removing", round(removal_pct, 1), "% of cells is low.\n")
  cat("   Check if you're retaining low-quality cells.\n\n")
} else {
  cat("✓ Removal rate of",round(removal_pct,1),"% is reasonable (typical: 10-25%)\n\n")
}
```

### Step 7c: Implement filtering thresholds

```{r step_7c, eval = FALSE}
# Track cells before cell-level QC filtering for summary
cells_before_cell_qc <- nrow(qc_df)
 
# Apply filters
seurat_obj <- subset(
  seurat_obj,
  subset = nCount_RNA >= ncount_min &
           nCount_RNA <= ncount_max &
           nFeature_RNA >= nfeature_min &
           nFeature_RNA <= nfeature_max &
           percent.mt < mt_thresh
)
 
cat("After filtering:", ncol(seurat_obj), "cells remaining\n")
```

## Step 8: Gene-Level Quality Control

After filtering cells, we filter genes based on how widely they’re detected across cells.

**Why Remove Hemoglobin?** In PBMC samples, hemoglobin genes (HBA1, HBA2, HBB) are typically from contaminating red blood cells during sample preparation, not true expression in PBMCs.

**Detection Threshold**: The percentage-based approach (e.g., 0.1% of cells) scales with dataset size. For 5,000 cells, 0.1% = 5 cells. This is more robust than fixed “3 cells” cutoffs when analyzing datasets of varying sizes.

### Note on filtering in this tutorial: Human data

The % MT, % ribo filtering and % HB steps below are based on **human** gene symbols. For future workflow development, select according to species.

### Step 8a: Visualization first

```{r step8a, fig.width = 8, fig.height = 5, eval = FALSE}
#-----------------------------------------------
# STEP 8: Gene-level QC with detection threshold
#-----------------------------------------------
 
# Calculate gene detection
counts_matrix <- LayerData(seurat_obj, layer = "counts")
gene_detection <- rowSums(counts_matrix > 0)
 
gene_qc <- data.frame(
  gene = rownames(seurat_obj),
  n_cells_detected = gene_detection,
  pct_cells_detected = (gene_detection / ncol(seurat_obj)) * 100,
  is_mt = grepl("^MT-", rownames(seurat_obj)),
  is_ribo = grepl("^RP[SL]", rownames(seurat_obj)),
  is_hb = grepl("^HB[AB]", rownames(seurat_obj))
)
 
cat("Total genes:", nrow(gene_qc), "\n")
cat("MT genes:", sum(gene_qc$is_mt), 
    "| Ribo genes:", sum(gene_qc$is_ribo), 
    "| Hb genes:", sum(gene_qc$is_hb), "\n\n")
 
# Visualize gene detection
(p9 <- ggplot(gene_qc, aes(x = pct_cells_detected)) +
  geom_histogram(bins = 50, fill = "#118AB2", alpha = 0.7) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Gene Detection Across Cells",
    subtitle = "Red line: 0.1% of cells threshold",
    x = "% of Cells Expressing Gene (log scale)",
    y = "Number of Genes"
  ) +
  theme_classic()
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_06_gene_detection.png")), p9, width = 8, height = 5, dpi = 300)
 
cat("→ EXAMINE plots/06_gene_detection.png\n")
cat("   Choose threshold based on where detection drops off\n\n")
 
# Set filtering threshold - ADJUST BASED ON YOUR DATA
# Common approaches:
#   - 0.1% of cells (very lenient, keeps most genes)
#   - 1% of cells (moderate, standard approach)
#   - 3 cells minimum (absolute count, conservative)

cat("Threshold options:\n")
cat("  • 0.1% of cells: Lenient (keeps rare cell type markers)\n")
cat("  • 1% of cells: Standard (balances detection vs noise)\n")
cat("  • 3-10 cells minimum: Conservative (removes very rare genes)\n\n")
 
```

### Step 8b: Implement gene per % or # cell threshold

```{r step_8b, eval = FALSE}
min_pct_cells <- 0.1  # Gene must be in ≥0.1% of cells
# Alternatively: min_cells <- 3  # Gene must be in ≥3 cells
 
cat("Setting gene filter threshold:\n")
cat("  Minimum detection: ≥", min_pct_cells, "% of cells\n")
cat("  (Equivalent to ≥", ceiling(ncol(seurat_obj) * min_pct_cells / 100), 
    "cells with current dataset)\n\n")
 
# Filter genes
genes_to_keep <- (gene_qc$pct_cells_detected >= min_pct_cells) & !gene_qc$is_hb
 
cat("Genes passing filter:", sum(genes_to_keep), "/", nrow(gene_qc), "\n")
cat("Genes removed:\n")
cat("  Low detection:", sum(gene_qc$pct_cells_detected < min_pct_cells), "\n")
cat("  Hemoglobin:", sum(gene_qc$is_hb), "\n\n")
 
seurat_obj <- seurat_obj[gene_qc$gene[genes_to_keep], ]
 
cat("After gene filtering:", nrow(seurat_obj), "genes remaining\n")
```

## Step 9 (optional): First Look at Highly Variable Genes

**Variable Features**: These genes show high variability across cells (not just high expression). They’re likely involved in defining cell types and states, making them most useful for downstream clustering and cell type identification (covered in the next tutorial).

```{r step_9, fig.width - 10, fig.height = 7, eval = FALSE}
#-----------------------------------------------
# STEP 9: Normalization and variable features
#-----------------------------------------------
 
# Log-normalize
seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", 
                            scale.factor = 10000, verbose = FALSE)
 
# Find highly variable genes
seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", 
                                   nfeatures = 2000, verbose = FALSE)
 
top10 <- head(VariableFeatures(seurat_obj), 10)
cat("Top 10 variable genes:", paste(top10, collapse = ", "), "\n")
# Top 10 variable genes: IGLC2, IGKC, LYPD2, LINGO2, IGLC3, PTGDS, FCER1A, IGHM, BANK1, S100A9
 
# Visualize variable features
p10 <- VariableFeaturePlot(seurat_obj)
p10 <- LabelPoints(plot = p10, points = top10, repel = TRUE)

print(p10)
ggsave(file.path(qc.dir,paste0("plots/",sample_name,"_07_variable_features.png")), p10, width = 10, height = 7, dpi = 300)


```

## Step 10: Export Filtered Cells and QC Summary

```{r step_10, eval = FALSE}
#-----------------------------------------------
# STEP 10: Save processed data and create summary
#-----------------------------------------------
 
# Save the clean Seurat object to filtered data directory
saveRDS(seurat_obj, file = file.path(filt.dir,paste0(sample_name,"_qc_filtered.rds")))
 
# Create ONE comprehensive QC summary (instead of many CSV files)
qc_summary <- data.frame(
  sample = sample_name,
 
  # Cell counts through QC pipeline
  initial_droplets = initial_droplet_count,  # From raw matrix before EmptyDrops
  after_emptydrops = length(validated_barcodes),  # Cells called by EmptyDrops
  after_soupx = length(validated_barcodes),       # SoupX doesn't remove cells
  after_doublets = cells_before_doublet_removal - n_doublets,  # After doublet removal
  after_cell_qc = ncol(seurat_obj),  # After cell-level QC filtering
  final_cells = ncol(seurat_obj),    # Final count (same as after_cell_qc)
 
  # Gene counts
  initial_genes = initial_gene_count,  # From raw matrix
  final_genes = nrow(seurat_obj),      # After gene filtering
 
  # Key metrics (from final filtered data)
  median_umi = median(seurat_obj$nCount_RNA),
  median_genes = median(seurat_obj$nFeature_RNA),
  median_mt_pct = median(seurat_obj$percent.mt),
 
  # QC parameters
  contamination_pct = ifelse(is.null(contamination_fraction), 0, 
                             round(contamination_fraction * 100, 2)),
  doublet_rate_pct = round(doublet_rate, 2),
 
  # Filtering thresholds used
  ncount_min = ncount_min,
  ncount_max = ncount_max,
  nfeature_min = nfeature_min,
  nfeature_max = nfeature_max,
  mt_threshold = mt_thresh
)
 
# Save comprehensive summary
write.csv(qc_summary, file.path(qc.dir,paste0("qc_metrics/",sample_name,"QC_summary.csv")), row.names = FALSE)
write.csv(seurat_obj@meta.data, file.path(filt.dir,paste0(sample_name,"_filt_cell_metadata.csv")))
 
cat("\n=== QC Pipeline Complete ===\n")
cat("Final cells:", ncol(seurat_obj), "\n")
cat("Final genes:", nrow(seurat_obj), "\n")
cat("Plots saved:", length(list.files("plots")), "files\n\n")
 
cat("Files created:\n")
cat(paste0("  • ",filt.dir,"/",sample_name,"_qc_filtered.rds - Clean Seurat object\n"))
cat(paste0("  • ",qc.dir,"/qc_metrics/",sample_name,"_QC_summary.csv - Comprehensive QC summary\n"))
cat(paste0("  • ",filt.dir,"/",sample_name,"_filt_cell_metadata.csv - Cell-level metadata\n"))

```

## Best Practices: Essential QC Principles

1. Visual Inspection is Key

- Plots are your primary QC tool
- Don’t just print metrics – visualize them
- One good plot > multiple summary statistics

2. Streamlined Reporting

- Brief console feedback during analysis
- Visual QC checks via plots
- ONE comprehensive summary at the end

3. Data-Driven Thresholds with Sanity Checks

- Use percentile-based thresholds
- Apply biological sanity checks – prevent unrealistic thresholds
- PBMCs: ≥500 genes, ≥800 UMI, <10% MT
- Warn if removal rate is >30% (too strict) or <5% (too lenient)
- Document all decisions and adjustments in your summary

4. Iterative QC

- QC is not one-and-done
- Return to QC plots after clustering
- Re-evaluate if results seem off

5. Biological Context

- Know your tissue’s expected metrics
- Different cell types have different profiles
- Stressed cells aren’t always technical artifacts

## Further Reading

### Official Documentation

- Seurat 5 vignettes: https://satijalab.org/seurat/
- Seurat GitHub: https://github.com/satijalab/seurat
- DropletUtils GitHub: https://github.com/MarioniLab/DropletUtils?tab=readme-ov-file
- SoupX GitHub: https://github.com/constantAmateur/SoupX
- scDblFinder GitHub: https://github.com/plger/scDblFinder

### References

- Hao Y, Stuart T, Kowalski MH, et al. Dictionary learning for integrative, multimodal and scalable single-cell analysis. Nat Biotechnol. 2024;42(2):293-304. [Seurat 5 paper]
- Luecken MD, Theis FJ. Current best practices in single-cell RNA-seq analysis: a tutorial. Mol Syst Biol. 2019;15(6):e8746.
- Lun ATL, Riesenfeld S, Andrews T, et al. EmptyDrops: distinguishing cells from empty droplets in droplet-based single-cell RNA sequencing data. Genome Biol. 2019;20(1):63.
- Young MD, Behjati S. SoupX removes ambient RNA contamination from droplet-based single-cell RNA sequencing data. Gigascience. 2020;9(12):giaa151.
- Germain PL, Lun A, Garcia Meixide C, Macnair W, Robinson MD. Doublet identification in single-cell sequencing data using scDblFinder. F1000Res. 2021;10:979.
- Osorio D, Cai JJ. Systematic determination of the mitochondrial proportion in human and mice tissues for single-cell RNA-sequencing data quality control. Bioinformatics. 2021;37(7):963-967.
- McCarthy DJ, Campbell KR, Lun ATL, Wills QF. Scater: pre-processing, quality control, normalization and visualization of single-cell RNA-seq data in R. Bioinformatics. 2017;33(8):1179-1186.
- Amezquita RA, Lun ATL, Becht E, et al. Orchestrating single-cell analysis with Bioconductor. Nat Methods. 2020;17(2):137-145.
- Zheng GXY, Terry JM, Belgrader P, et al. Massively parallel digital transcriptional profiling of single cells. Nat Commun. 2017;8:14049.
- Lee H, Joo JY, Sohn DH, et al. Single-cell RNA sequencing reveals rebalancing of immunological response in patients with periodontitis after non-surgical periodontal therapy. J Transl Med. 2022;20(1):504. [Dataset source]

