---
title: "ngs101_scRNAseq_walkthrough_2_QC - adapted to multiple samples"
subtitle: "part 1 - empty droplets, ambient RNA, doublet detection, initial visualization"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)
```

## Define sample paths and correction parameters

Also define vector objects to receive QC metrics

```{r step2}

# Define paths
sample_names <- c("Healthy_1",
                  "Healthy_2",
                  "Healthy_3",
                  "Healthy_4",
                  "Pre_Patient_1",
                  "Pre_Patient_2",
                  "Pre_Patient_3",
                  "Pre_Patient_4",
                  "Post_Patient_1",
                  "Post_Patient_2",
                  "Post_Patient_3",
                  "Post_Patient_4") # corresponding to cellranger_output/{SAMPLE}/outs

empty_lower <- 100 # 100 UMIs necessary to be counted as non-empty droplet
empty_niters <- 10000 # iterations for EmptyDrops p-value calculation
empty_FDR <- 0.01 # FDR threshold for detecting non-empty droplets

ambient_cutoff <- 0.05 # lowpass filter to decide whether to execute SoupX

initial_droplet_count <- vector(mode = "numeric", length = length(sample_names))
initial_gene_count <- vector(mode = "numeric", length = length(sample_names))
validated_barcode_count <- vector(mode = "numeric", length = length(sample_names))

ambient_fraction <- vector(mode = "numeric", length = length(sample_names))

n_doublets <- vector(mode = "numeric", length = length(sample_names))
doublet_rate <- vector(mode = "numeric", length = length(sample_names))
cells_before_doublet_removal <- vector(mode = "numeric", length = length(sample_names))

p1 <- list() # for collecting empty droplet ggplots
p2 <- list() # for collecting doublet ggplots

```

## Execute loop for empty droplet, ambient RNA and doublet detection

```{r loop1}

for (i in seq_along(sample_names)) {
  cat("Starting QC part 1 for",sample_names[i],"\n")
  cellranger_output.dir <- file.path(base.dir,
                                     paste0("cellranger_output/", sample_names[i], "/outs"))
  counts <- Read10X(data.dir = file.path(cellranger_output.dir, "raw_feature_bc_matrix"))
  seurat_obj <- CreateSeuratObject(
    counts = counts,
    project = sample_names[i],
    min.cells = 0,
    min.features = 0
  )
  seurat_obj$sample_id <- sample_names[i]
  
  ### EmptyDroplet procedures
  cat("EmptyDrops procedure\n")
  sce <- as.SingleCellExperiment(seurat_obj)
  
  # Track initial droplet count for QC summary
  initial_droplet_count[i] <- ncol(sce)
  initial_gene_count[i] <- nrow(sce)
  
  # Run EmptyDrops
  set.seed(27)
  empty_results <- emptyDrops(
    m = counts(sce),
    lower = empty_lower,
    # Droplets with <100 UMI assumed empty (for estimating ambient RNA)
    niters = empty_niters,
    # Iterations for p-value calculation
    test.ambient = TRUE # Test if droplets differ from ambient RNA profile
  )
  
  # Identify cells (FDR < 0.01)
  is_cell <- empty_results$FDR < empty_FDR
  is_cell[is.na(is_cell)] <- FALSE  # Treat NA as empty (very low UMI droplets)
  validated_barcodes <- colnames(sce)[is_cell]
  validated_barcode_count[i] <- length(validated_barcodes)
  
  # Visual QC - the critical check!
  empty_df <- data.frame(total_umi = colSums(counts(sce)), is_cell = is_cell)
  
  p1[[i]] <- ggplot(empty_df, aes(x = log10(total_umi + 1), fill = is_cell)) +
    geom_histogram(bins = 50,
                   alpha = 0.7,
                   position = "identity") +
    scale_fill_manual(
      values = c("TRUE" = "#2E86AB", "FALSE" = "#A23B72"),
      labels = c("Empty Droplet", "Cell"),
      name = "Classification"
    ) +
    labs(
      title = "EmptyDrops: Cell vs Empty Droplet Detection",
      subtitle = paste(
        sample_names[i],
        sum(is_cell),
        "cells called from",
        ncol(sce),
        "total droplets"
      ),
      x = "log10(UMI + 1)",
      y = "Number of Droplets"
    ) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold", size = 14))
  
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[i], "_01_empty_droplets.png")
    ),
    p1[[i]],
    width = 10,
    height = 6,
    dpi = 300
  )
  
  # Filter the Seurat object to keep only validated cells
  # IMPORTANT: Save raw counts BEFORE filtering (needed for SoupX)
  raw_counts_all_droplets <- LayerData(seurat_obj, layer = "counts")
  
  seurat_obj <- subset(seurat_obj, cells = validated_barcodes)
  rm(sce)
  
  ### SoupX procedures
  cat("SoupX procedure\n")
  # Prepare SoupX channel
  # tod (table of droplets) = raw counts from ALL droplets
  # toc (table of cells) = filtered counts from validated cells only
  tod <- raw_counts_all_droplets
  toc <- LayerData(seurat_obj, layer = "counts")
  sc <- SoupChannel(tod = tod,
                    toc = toc,
                    calcSoupProfile = TRUE)
  
  # Quick standard Seurat clustering for contamination estimation; clusters needed by SoupX
  temp_obj <- seurat_obj
  temp_obj <- NormalizeData(temp_obj, verbose = FALSE)
  temp_obj <- FindVariableFeatures(temp_obj, nfeatures = 2000, verbose = FALSE)
  temp_obj <- ScaleData(temp_obj, verbose = FALSE)
  temp_obj <- RunPCA(temp_obj, npcs = 30, verbose = FALSE)
  temp_obj <- FindNeighbors(temp_obj, dims = 1:30, verbose = FALSE)
  temp_obj <- FindClusters(temp_obj, resolution = 0.8, verbose = FALSE)
  sc <- setClusters(sc, setNames(
    as.character(temp_obj$seurat_clusters),
    colnames(temp_obj)
  ))
  
  # Estimate contamination
  sc <- tryCatch({
    autoEstCont(sc, verbose = FALSE)
  }, error = function(e) {
    sc$fit$rho <- NULL # if autoEstCont fails
    return(sc)
  })
  
  if (!is.null(sc$fit$rho) && sc$fit_rho != "") {
    ambient_fraction[i] <- sc$fit$rho
  } else {
    ambient_fraction[i] <- NA
  }
      
  # Decision: Skip or correct based on contamination level
  if (!is.na(ambient_fraction[i]) &&
      ambient_fraction[i] >= ambient_cutoff) {
    # Apply correction
    suppressWarnings({
      corrected_counts <- adjustCounts(sc)
    })
    seurat_obj <- SetAssayData(seurat_obj, layer = "counts", new.data = corrected_counts)
  }
  
  rm(sc, temp_obj, toc, tod, raw_counts_all_droplets)
  
  ### Doublet detection procedure (scDblFinder)
  cat("Doublet detection with scDblFinder\n")
  sce <- as.SingleCellExperiment(seurat_obj)
  
  set.seed(27)
  # Run scDblFinder (suppress expected warnings about empty layers and xgboost)
  suppressWarnings({
    sce <- scDblFinder(sce, dbr = NULL, verbose = FALSE)
  })
  
  # Extract results
  seurat_obj$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
  seurat_obj$doublet_score <- sce$scDblFinder.score
  
  # Brief feedback
  n_doublets[i] <- sum(seurat_obj$doublet_class == "Doublet")
  doublet_rate[i] <- n_doublets[i] / ncol(seurat_obj) * 100
  
  # Track cells before doublet removal for QC summary
  cells_before_doublet_removal[i] <- ncol(seurat_obj)
  
  # Quick UMAP for visualization
  seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
  seurat_obj <- FindVariableFeatures(seurat_obj, nfeatures = 2000, verbose = FALSE)
  seurat_obj <- ScaleData(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, npcs = 30, verbose = FALSE)
  suppressWarnings({
    seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, verbose = FALSE)
  })
  
  # Visualize doublets on UMAP
  p2[[i]] <- DimPlot(
    seurat_obj,
    reduction = "umap",
    group.by = "doublet_class",
    cols = c(
      "Singlet" = "#90E0EF",
      "Doublet" = "#EF233C"
    ),
    pt.size = 1
  ) +
    labs(title = paste0("Doublet Detection ",sample_names[i]," (", round(doublet_rate[i], 1), "%)")) +
    theme(plot.title = element_text(face = "bold", size = 14))
  
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[i], "_02_doublets_umap.png")
    ),
    p2[[i]],
    width = 8,
    height = 7,
    dpi = 300
  )
  
  # Remove doublets
  seurat_obj <- subset(seurat_obj, subset = doublet_class == "Singlet")
  
  cat("Saving qc_part1_filtered",sample_names[i],"rds object\n")
  saveRDS(seurat_obj,
          file = file.path(filt.dir, paste0(
    sample_names[i], "_qc_part1_filtered.rds"
  )))
  
  rm(seurat_obj)
}

```

### Save out QC metric objects so far

```{r save_QC_part1_metrics}

qc_part1.file <- "qc_part1.RData"

save(
sample_names,
empty_lower,
empty_niters,
ambient_cutoff,
initial_droplet_count,
initial_gene_count,
validated_barcode_count,
ambient_fraction,
n_doublets,
doublet_rate,
cells_before_doublet_removal,
file = file.path(qc.dir,"qc_metrics", qc_part1.file)
)
```

## Cell-Level Quality Control: Visualization (unfiltered cells)

Individual cells are filtered based on their quality metrics. The key is visual inspection – look at the distributions and set biologically meaningful thresholds rather than relying on automatic statistical methods.

This is a three-step procedure:

- Visualization of unfiltered cell distributions, including % MT and % ribo genes
- Parameter selection and visualization of impact on individual cells with before & after plots
- Implement chosen filtering parameters

### Note on filtering in this walkthrough: Human data

The % MT and % ribo filtering steps below are based on **human** gene symbols. For future workflow development, select according to species.

**For snRNAseq work**, consider visualization of MALAT1 as an additional violin plot.

### Unfiltered cells visualization (loop)

```{r step_7a, fig.width = 15, fig.height = 5, eval = TRUE}
#-----------------------------------------------
# STEP 7: Calculate QC metrics, visualize, and filter cells
#-----------------------------------------------

for (j in seq_along(sample_names)) {
  cat("Starting QC part 2 for",sample_names[j],"\n")
  seurat_obj <- readRDS(file = file.path(filt.dir,
                                         paste0(sample_names[j], "_qc_part1_filtered.rds")
                                         )
                        )
# Calculate QC metrics
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^RP[SL]")
seurat_obj$log10GenesPerUMI <- log10(seurat_obj$nFeature_RNA) / log10(seurat_obj$nCount_RNA)
 
# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions for",sample_names[j],"\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(seurat_obj$nCount_RNA), 
    "| Q1-Q3:", quantile(seurat_obj$nCount_RNA, 0.25), "-", 
    quantile(seurat_obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(seurat_obj$nFeature_RNA),
    "| Q1-Q3:", quantile(seurat_obj$nFeature_RNA, 0.25), "-",
    quantile(seurat_obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(seurat_obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(seurat_obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(seurat_obj$percent.ribo), 2), "%\n")
 
# Visualize distributions - THE CRITICAL STEP
(p3 <- VlnPlot(
  seurat_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) &
  theme(plot.title = element_text(face = "bold"))
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_names[j],"_03_qc_violins.png")), p3, width = 16, height = 4, dpi = 300)
 
# Scatter plots showing metric relationships
p4 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p5 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p6 <- FeatureScatter(seurat_obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
(p_scatter <- p4 + p5 + p6)
ggsave(file.path(qc.dir,paste0("plots/",sample_names[j],"_04_qc_scatter.png")), p_scatter, width = 15, height = 5, dpi = 300)

cat("Re-saving qc_part1_filtered",sample_names[j],"rds object with additional metrics\n")
  saveRDS(seurat_obj,
          file = file.path(filt.dir, paste0(
    sample_names[j], "_qc_part1_filtered.rds"
  )))
  rm(seurat_obj)
}

cat("\n→ EXAMINE plots/03_qc_violins.png and plots/04_qc_scatter.png\n")
cat("→ Look for:\n")
cat("   - Bimodal distributions (good vs bad cells)\n")
cat("   - Outliers in scatter plots\n")
cat("   - Relationship between metrics\n\n")

cat("Guidelines for threshold setting:\n")
cat("  • nFeature_RNA min: Where does the low-quality tail end? (typically 500-1000)\n")
cat("  • nFeature_RNA max: Where do potential doublets start? (typically 4000-7000)\n")
cat("  • percent.mt: PBMCs <10%, Brain <5%, Tumor <20%\n")
cat("  • If distributions are bimodal, set threshold between modes\n")
cat("  • Aim to remove no more than 25% of cells \n\n")
```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"QC_part1_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"QC_part1_sessionInfo_summary.txt"))

sessionInfo()
```




