---
title: "ngs101_scRNAseq_walkthrough_2_QC - adapted to multiple samples"
subtitle: "part 1 - doublet detection, initial visualization"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
params:
  base.dir: "~/GSE174609_scRNA" # working directory for analysis files; subdirectories will be created
  input.dir: "~/GSE174609_scRNA/counts_input" # directory with raw coun matrices
  species: "human" # for QC metrics; options are "human" or "mouse"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

Expects `cellranger_output` subdirectory with multiple samples within, to permit evaluation of empty droplets, doublet detection and ambient RNA correction.

If these are not available (such as scRNAseq data in the form of processed count matrices from GEO, or split-pool assays such as Scale or Parse that do not use microfludics), then the user should use an alternate .Rmd (_multi_sample_QC_part1_alternate.Rmd) that skips the empty droplet and ambient RNA steps.

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- params$base.dir # may contain a subdirectory with multiple samples within
input.dir <- params$input.dir
qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)
```

## Define sample paths and correction parameters

Also define vector objects to receive QC metrics

```{r step2}

# Define paths
sample_names <- c("Healthy_1",
                  "Healthy_2",
                  "Healthy_3",
                  "Healthy_4",
                  "Pre_Patient_1",
                  "Pre_Patient_2",
                  "Pre_Patient_3",
                  "Pre_Patient_4",
                  "Post_Patient_1",
                  "Post_Patient_2",
                  "Post_Patient_3",
                  "Post_Patient_4") # corresponding to cellranger_output/{SAMPLE}/outs

## Variables below not needed in absence of empty droplet and ambient RNA correction steps
# empty_lower <- 100 # 100 UMIs necessary to be counted as non-empty droplet
# empty_niters <- 10000 # iterations for EmptyDrops p-value calculation
# empty_FDR <- 0.01 # FDR threshold for detecting non-empty droplets

# ambient_cutoff <- 0.05 # lowpass filter to decide whether to execute SoupX

initial_droplet_count <- vector(mode = "numeric", length = length(sample_names))
initial_gene_count <- vector(mode = "numeric", length = length(sample_names))
validated_barcode_count <- vector(mode = "numeric", length = length(sample_names))

ambient_fraction <- vector(mode = "numeric", length = length(sample_names))

n_doublets <- vector(mode = "numeric", length = length(sample_names))
doublet_rate <- vector(mode = "numeric", length = length(sample_names))
cells_before_doublet_removal <- vector(mode = "numeric", length = length(sample_names))

# p1 <- list() # for collecting empty droplet ggplots
p2 <- list() # for collecting doublet ggplots

```

## Execute loop for Seurat object creation and doublet detection

In the section below, the `counts` matrix and the directory path for each individual sample will have to be altered for specific projects.

```{r loop1}

for (i in seq_along(sample_names)) {
  cat("Starting QC part 1 for",sample_names[i],"\n")
  # cellranger_output.dir <- file.path(base.dir,
  #                                    paste0("cellranger_output/", sample_names[i], "/outs"))
  # counts <- Read10X(data.dir = file.path(cellranger_output.dir, "raw_feature_bc_matrix"))

  ## Examples
  count_input.dir <- file.path(input.dir, sample_names[i])) 
  counts <- Read10X(data.dir = count_input.dir) # may need a different function or even read.csv() if just plain count matrix
  seurat_obj <- CreateSeuratObject(
    counts = counts,
    project = sample_names[i],
    min.cells = 0,
    min.features = 0
  )
  seurat_obj$sample_id <- sample_names[i]
  
  ### Doublet detection procedure (scDblFinder)
  cat("Doublet detection with scDblFinder\n")
  sce <- as.SingleCellExperiment(seurat_obj)
  
  set.seed(27)
  # Run scDblFinder (suppress expected warnings about empty layers and xgboost)
  suppressWarnings({
    sce <- scDblFinder(sce, dbr = NULL, verbose = FALSE)
  })
  
  # Extract results
  seurat_obj$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
  seurat_obj$doublet_score <- sce$scDblFinder.score
  
  # Brief feedback
  n_doublets[i] <- sum(seurat_obj$doublet_class == "Doublet")
  doublet_rate[i] <- n_doublets[i] / ncol(seurat_obj) * 100
  
  # Track cells before doublet removal for QC summary
  cells_before_doublet_removal[i] <- ncol(seurat_obj)
  
  # Quick UMAP for visualization
  seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
  seurat_obj <- FindVariableFeatures(seurat_obj, nfeatures = 2000, verbose = FALSE)
  seurat_obj <- ScaleData(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, npcs = 30, verbose = FALSE)
  suppressWarnings({
    seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, verbose = FALSE)
  })
  
  # Visualize doublets on UMAP
  p2[[i]] <- DimPlot(
    seurat_obj,
    reduction = "umap",
    group.by = "doublet_class",
    cols = c(
      "Singlet" = "#90E0EF",
      "Doublet" = "#EF233C"
    ),
    pt.size = 1
  ) +
    labs(title = paste0("Doublet Detection ",sample_names[i]," (", round(doublet_rate[i], 1), "%)")) +
    theme(plot.title = element_text(face = "bold", size = 14))
  
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[i], "_02_doublets_umap.png")
    ),
    p2[[i]],
    width = 8,
    height = 7,
    dpi = 300
  )
  
  # Remove doublets
  seurat_obj <- subset(seurat_obj, subset = doublet_class == "Singlet")
  
  cat("Saving qc_part1_filtered",sample_names[i],"rds object\n")
  saveRDS(seurat_obj,
          file = file.path(filt.dir, paste0(
    sample_names[i], "_qc_part1_filtered.rds"
  )))
  
  rm(seurat_obj)
}

```

### Save out QC metric objects so far

```{r save_QC_part1_metrics}

qc_part1.file <- "qc_part1.RData"

# Maintain compatibility with cellranger-based code for construction of final QC metrics in part3.Rmd
empty_lower <- NA
empty_niters <- NA
ambient_cutoff <- NA
initial_droplet_count <- rep(NA, length(sample_names))
initial_gene_count <- rep(NA, length(sample_names))
validated_barcode_count <- rep(NA, length(sample_names))
ambient_fraction <- rep(NA, length(sample_names))

save(
sample_names,
empty_lower,
empty_niters,
ambient_cutoff,
initial_droplet_count,
initial_gene_count,
validated_barcode_count,
ambient_fraction,
n_doublets,
doublet_rate,
cells_before_doublet_removal,
file = file.path(qc.dir,"qc_metrics", qc_part1.file)
)
```

## Cell-Level Quality Control: Visualization (unfiltered cells)

Individual cells are filtered based on their quality metrics. The key is visual inspection – look at the distributions and set biologically meaningful thresholds rather than relying on automatic statistical methods.

This is a three-step procedure:

- Visualization of unfiltered cell distributions, including % MT and % ribo genes
- Parameter selection and visualization of impact on individual cells with before & after plots
- Implement chosen filtering parameters

**For snRNAseq work**, consider visualization of MALAT1 as an additional violin plot.

### Unfiltered cells visualization (loop)

```{r step_7a, fig.width = 15, fig.height = 5, eval = TRUE}
#-----------------------------------------------
# STEP 7: Calculate QC metrics, visualize, and filter cells
#-----------------------------------------------
nCount_RNA <- vector(mode = "numeric", length = length(sample_names))
nFeature_RNA <- vector(mode = "numeric", length = length(sample_names))
percent_mt <- vector(mode = "numeric", length = length(sample_names))
percent_ribo <- vector(mode = "numeric", length = length(sample_names))
MALAT1_ratio <- vector(mode = "numeric", length = length(sample_names))

for (j in seq_along(sample_names)) {
  cat("Starting QC part 2 for",sample_names[j],"\n")
  seurat_obj <- readRDS(file = file.path(filt.dir,
                                         paste0(sample_names[j], "_qc_part1_filtered.rds")
                                         )
                        )

if (params$species == "human") {
  ribo_pattern <- "^RP[SL]"
  mt_pattern <- "^MT-"
  malat1_pattern <- "MALAT1"
} else if (params$species == "mouse") {
  ribo_pattern <- "^Rp[sl]"
  mt_pattern <- "^mt-"
  malat1_pattern <- "Malat1"
} else {
  stop("Invalid species parameter. Please use 'human' or 'mouse'.")
}
# Calculate QC metrics
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = mt_pattern)
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = ribo_pattern)
seurat_obj$log10GenesPerUMI <- log10(seurat_obj$nFeature_RNA) / log10(seurat_obj$nCount_RNA)
seurat_obj[["MALAT1_ratio"]] <- FetchData(seurat_obj, vars = malat1_pattern, layer = "counts") / seurat_obj$nCount_RNA

# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions for",sample_names[j],"\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(seurat_obj$nCount_RNA), 
    "| Q1-Q3:", quantile(seurat_obj$nCount_RNA, 0.25), "-", 
    quantile(seurat_obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(seurat_obj$nFeature_RNA),
    "| Q1-Q3:", quantile(seurat_obj$nFeature_RNA, 0.25), "-",
    quantile(seurat_obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(seurat_obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(seurat_obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(seurat_obj$percent.ribo), 2), "%\n")

cat("MALAT1_ratio:\n")
cat("  Median:", round(median(seurat_obj$MALAT1_ratio), 2), "%\n")
 
# Visualize distributions - THE CRITICAL STEP
(p3 <- VlnPlot(
  seurat_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) &
  theme(plot.title = element_text(face = "bold"))
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_names[j],"_03_qc_violins.png")), p3, width = 16, height = 4, dpi = 300)
 
# Scatter plots showing metric relationships
p4 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p5 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p6 <- FeatureScatter(seurat_obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
(p_scatter <- p4 + p5 + p6)
ggsave(file.path(qc.dir,paste0("plots/",sample_names[j],"_04_qc_scatter.png")), p_scatter, width = 15, height = 5, dpi = 300)

cat("Re-saving qc_part1_filtered",sample_names[j],"rds object with additional metrics\n")
  saveRDS(seurat_obj,
          file = file.path(filt.dir, paste0(
    sample_names[j], "_qc_part1_filtered.rds"
  )))
  rm(seurat_obj)
}
```

### Summary of unfiltered cells

All metrics printed above are captured and saved out as a data.frame `cellQC_unfilt_metrics_df` within `r file.path(qc.dir,"qc_metrics",qc_part1.file)`

```{r summary_unfilt}

cellQC_unfilt_metrics_df <- data.frame(
  sample = sample_names,
  median_nCount_RNA = nCount_RNA,
  median_nFeature_RNA = nFeature_RNA,
  median_percent_mt = percent_mt,
  median_percent_ribo = percent_ribo,
  median_MALAT1_ratio = MALAT1_ratio
)

# Update qc_part1 file
save(sample_names,
     empty_lower,
     empty_niters,
     ambient_cutoff,
     initial_droplet_count,
     initial_gene_count,
     validated_barcode_count,
     ambient_fraction,
     n_doublets,
     doublet_rate,
     cells_before_doublet_removal,
     cellQC_unfilt_metrics_df,
     file = file.path(qc.dir,"qc_metrics", qc_part1.file)
)
cat("\n→ EXAMINE plots/03_qc_violins.png and plots/04_qc_scatter.png\n")
cat("→ Look for:\n")
cat("   - Bimodal distributions (good vs bad cells)\n")
cat("   - Outliers in scatter plots\n")
cat("   - Relationship between metrics\n\n")

cat("Guidelines for threshold setting:\n")
cat("  • nFeature_RNA min: Where does the low-quality tail end? (typically 500-1000)\n")
cat("  • nFeature_RNA max: Where do potential doublets start? (typically 4000-7000)\n")
cat("  • percent.mt: PBMCs <10%, Brain <5%, Tumor <20%\n")
cat("  • If distributions are bimodal, set threshold between modes\n")
cat("  • Aim to remove no more than 25% of cells \n\n")
```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"QC_part1_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"QC_part1_sessionInfo_summary.txt"))

sessionInfo()
```




