---
title: "ngs101_scRNAseq_walkthrough_2_QC - adapted to multiple samples"
subtitle: "part 3 - gene-level QC: choose thresholds, visualize effects"
author: "Scot J Matkovich; Lei Guo"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

Primary source:

<https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-2-quality-control-and-cell-filtering>

## Libraries and directories

```{r step1_libs_data}

# Core single-cell analysis
library(Seurat)              # Main scRNA-seq toolkit
library(SeuratObject)        # Seurat object infrastructure
 
# QC-specific packages
library(DropletUtils)        # Empty droplet detection
library(scDblFinder)         # Doublet detection
library(SoupX)               # Ambient RNA correction
library(SingleCellExperiment) # SCE objects
 
# Visualization
library(ggplot2)             # Custom plots when needed
library(patchwork)           # Combine plots
library(dplyr)               # Data manipulation
 
# Set working directory
base.dir <- "~/GSE174609_scRNA"
qc.dir <- file.path(base.dir,"qc_analysis_multi_loop")
if (!dir.exists(qc.dir)) dir.create(qc.dir, recursive = TRUE)
filt.dir <- file.path(base.dir,"filtered_data")
if (!dir.exists(filt.dir)) dir.create(filt.dir, recursive = TRUE)

# Create output directories
if (!dir.exists(file.path(qc.dir,"plots"))) dir.create(file.path(qc.dir,"plots"), showWarnings = FALSE)
if (!dir.exists(file.path(qc.dir,"qc_metrics"))) dir.create(file.path(qc.dir,"qc_metrics"), showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(27)

qc_part1.file <- "qc_part1.RData"
load(file.path(qc.dir,"qc_metrics",qc_part1.file))

qc_part2b.file <- "qc_part2b.RData"
load(file.path(qc.dir,"qc_metrics",qc_part2b.file))
```

## Gene-Level Quality Control: Visualization

After filtering cells, genes are filtered based on how widely they’re detected across cells.

**Why Remove Hemoglobin?** In PBMC samples, hemoglobin genes (HBA1, HBA2, HBB) are typically from contaminating red blood cells during sample preparation, not true expression in PBMCs.

**Detection Threshold**: The percentage-based approach (e.g., 0.1% of cells) scales with dataset size. For 5,000 cells, 0.1% = 5 cells. This is more robust than fixed “3 cells” cutoffs when analyzing datasets of varying sizes.

### Note on filtering in this tutorial: Human data

The % MT, % ribo filtering and % HB steps below are based on **human** gene symbols. For future workflow development, select according to species.

### Visualization prior to filtering (loop)

```{r step8a, fig.width = 8, fig.height = 5, eval = TRUE}
#-----------------------------------------------
# STEP 8: Gene-level QC with detection threshold
#-----------------------------------------------

gene_qc_list <- list()

for (n in seq_along(sample_names)) {

  cat("Re-reading cell-filtered Seurat object for", sample_names[n],"\n")
  seurat_obj <- readRDS(file = file.path(filt.dir, paste0(
    sample_names[n], "_qc_part2_filtered.rds"
  )))

  # Calculate gene detection
counts_matrix <- LayerData(seurat_obj, layer = "counts")
gene_detection <- rowSums(counts_matrix > 0)
 
gene_qc_list[[n]] <- data.frame(
  gene = rownames(seurat_obj),
  n_cells_detected = gene_detection,
  pct_cells_detected = (gene_detection / ncol(seurat_obj)) * 100,
  is_mt = grepl("^MT-", rownames(seurat_obj)),
  is_ribo = grepl("^RP[SL]", rownames(seurat_obj)),
  is_hb = grepl("^HB[AB]", rownames(seurat_obj))
)
names(gene_qc_list)[n] <- sample_names[n]

cat("Total genes:", nrow(gene_qc_list[[n]]), "\n")
cat("MT genes:", sum(gene_qc_list[[n]]$is_mt), 
    "| Ribo genes:", sum(gene_qc_list[[n]]$is_ribo), 
    "| Hb genes:", sum(gene_qc_list[[n]]$is_hb), "\n\n")
 
# Visualize gene detection
(p9 <- ggplot(gene_qc_list[[n]], aes(x = pct_cells_detected)) +
  geom_histogram(bins = 50, fill = "#118AB2", alpha = 0.7) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Gene Detection Across Cells",
    subtitle = "Red line: 0.1% of cells threshold",
    x = "% of Cells Expressing Gene (log scale)",
    y = "Number of Genes"
  ) +
  theme_classic()
)
 
ggsave(file.path(qc.dir,paste0("plots/",sample_names[n],"_06_gene_detection.png")), p9, width = 8, height = 5, dpi = 300)

}

cat("→ EXAMINE plots/06_gene_detection.png\n")
cat("   Choose threshold based on where detection drops off\n\n")
 
# Set filtering threshold - ADJUST BASED ON YOUR DATA
# Common approaches:
#   - 0.1% of cells (very lenient, keeps most genes)
#   - 1% of cells (moderate, standard approach)
#   - 3 cells minimum (absolute count, conservative)

cat("Threshold options:\n")
cat("  • 0.1% of cells: Lenient (keeps rare cell type markers)\n")
cat("  • 1% of cells: Standard (balances detection vs noise)\n")
cat("  • 3-10 cells minimum: Conservative (removes very rare genes)\n\n")
 
```

### Save out QC metric objects generated before threshold setting

Outputs gathered into `qc_part3.RData` object

```{r save_qc_part3, eval = TRUE}
qc_part3.file <- "qc_part3.RData"

save(
sample_names,
gene_qc_list,
file = file.path(qc.dir,"qc_metrics", qc_part3.file)
)
```

## Re-import previous QC metric files if needed

Default is `eval = FALSE`

```{r re-import_previous_QC_part3_metrics, eval = TRUE}
qc_part3.file <- "qc_part3.RData"
load(file = file.path(qc.dir,"qc_metrics", qc_part3.file))
```

## Gene-Level Quality Control: Implement gene per % or # cell threshold

Set up a `min_pct_cells` vector to implement separate cutoffs for each Seurat object in `sample_names`

Don't run the following implementation section prior to inspecting gene-level QC plots and choosing cutoffs.

After implementing gene-level cutoffs, plots are generated for **Variable Features**. These genes show high variability across cells (not just high expression). They’re likely involved in defining cell types and states, making them most useful for downstream clustering and cell type identification.

```{r step_8b_9, fig.width = 10, fig.height = 7,eval = TRUE}

min_pct_cells <- rep(0.1, 12)  # Gene must be in ≥0.1% of cells
# Alternatively: min_cells <- 3  # Gene must be in ≥3 cells

after_cell_qc <- NULL
final_gene_count <- NULL
median_umi <- NULL
median_genes <- NULL
median_mt_pct <- NULL

for (p in seq_along(sample_names)) {
  cat("Re-reading cell-filtered Seurat object for", sample_names[p],"\n")
  seurat_obj <- readRDS(file = file.path(filt.dir, paste0(
    sample_names[p], "_qc_part2_filtered.rds"
  )))
  
  cat("Setting gene filter threshold for", sample_names[p], "\n")
  cat("  Minimum detection: ≥", min_pct_cells[p], "% of cells\n")
  cat(
    "  (Equivalent to ≥",
    ceiling(ncol(seurat_obj) * min_pct_cells[p] / 100),
    "cells with current dataset)\n\n"
  )
  
  # Filter genes
  genes_to_keep <- (gene_qc_list[[p]]$pct_cells_detected >= min_pct_cells[p]) &
    !gene_qc_list[[p]]$is_hb
  
  cat("Genes passing filter:",
      sum(genes_to_keep),
      "/",
      nrow(gene_qc_list[[p]]),
      "\n")
  cat("Genes removed:\n")
  cat("  Low detection:",
      sum(gene_qc_list[[p]]$pct_cells_detected < min_pct_cells[p]),
      "\n")
  cat("  Hemoglobin:", sum(gene_qc_list[[p]]$is_hb), "\n\n")
  
  seurat_obj <- seurat_obj[gene_qc_list[[p]]$gene[genes_to_keep], ]
  
  # Capture QC metrics
  cat("After gene filtering:",
      nrow(seurat_obj),
      "genes remaining\n")
  after_cell_qc[p] <- nrow(seurat_obj)
  final_gene_count[p] <- ncol(seurat_obj)
  median_umi[p] <- median(seurat_obj$nCount_RNA)
  median_genes[p] <- median(seurat_obj$nFeature_RNA)
  median_mt_pct[p] <- median(seurat_obj$percent.mt)
  
  write.csv(seurat_obj@meta.data, file.path(filt.dir,paste0(sample_names[p],"_filt_cell_metadata.csv")))
  
  cat("Saving new qc_final_filtered Seurat object for",
      sample_names[p],
      "\n")
  saveRDS(seurat_obj, file = file.path(filt.dir, paste0(
    sample_names[p], "_qc_final_filtered.rds"
  )))
  
  #-----------------------------------------------
  # STEP 9: Normalization and variable features (first looks)
  #-----------------------------------------------
  
  ## The following steps perform initial normalization and HVG steps in Seurat objects
  ## These won't be saved, facilitating later combination of the objects into one
  ## (or multi-layer assay)
  
  # Log-normalize
  seurat_obj <- NormalizeData(
    seurat_obj,
    normalization.method = "LogNormalize",
    scale.factor = 10000,
    verbose = FALSE
  )
  
  # Find highly variable genes
  seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures = 2000,
    verbose = FALSE
  )
  
  top10 <- head(VariableFeatures(seurat_obj), 10)
  cat("Top 10 variable genes:", paste(top10, collapse = ", "), "\n\n")
  # Top 10 variable genes: IGLC2, IGKC, LYPD2, LINGO2, IGLC3, PTGDS, FCER1A, IGHM, BANK1, S100A9
  
  # Visualize variable features
  p10 <- VariableFeaturePlot(seurat_obj)
  p10 <- LabelPoints(plot = p10,
                     points = top10,
                     repel = TRUE)
  
  print(p10)
  ggsave(
    file.path(
      qc.dir,
      paste0("plots/", sample_names[p], "_07_variable_features.png")
    ),
    p10,
    width = 10,
    height = 7,
    dpi = 300
  )
  
  rm(seurat_obj)
}
```

## Export QC summary for all processed Seurat objects

```{r step_10, eval = TRUE}
#-----------------------------------------------
# STEP 10: Save processed data and create summary
#-----------------------------------------------
 
# Create ONE comprehensive QC summary (instead of many CSV files)
qc_summary <- data.frame(
  sample = sample_names,
 
  # Cell counts through QC pipeline
  initial_droplets = initial_droplet_count,  # From raw matrix before EmptyDrops
  after_emptydrops = validated_barcode_count,  # Cells called by EmptyDrops
  after_soupx = validated_barcode_count,       # SoupX doesn't remove cells
  
  # Cell QC parameters
  ambient_pct = ifelse(is.null(ambient_fraction), 0, 
                             round(ambient_fraction * 100, 2)),
  doublet_rate_pct = round(doublet_rate, 2),
 
  # Cell filtering thresholds used
  ncount_min = threshold_df_final$ncount_min,
  ncount_max = threshold_df_final$ncount_max,
  nfeature_min = threshold_df_final$nfeature_min,
  nfeature_max = threshold_df_final$nfeature_max,
  mt_threshold = threshold_df_final$mt_thresh,
  
  # Cell counts
  initial_cells_after_doublet_removal = cells_before_doublet_removal - n_doublets,  # After doublet removal
  final_cells_after_cell_qc = after_cell_qc,  # After cell-level QC filtering
 
  # Gene counts
  initial_genes = initial_gene_count,  # From raw matrix
  final_genes = final_gene_count,      # After gene filtering
 
  # Key metrics (from final filtered data)
  median_umi = median_umi,
  median_genes = median_genes,
  median_mt_pct = median_mt_pct
)
 
# Save comprehensive summary
write.csv(qc_summary, file.path(qc.dir,paste0("qc_metrics/all_samples_QC_summary.csv")), row.names = FALSE)

cat("Files created:\n")
cat(paste0("  • ",qc.dir,"/qc_metrics/all_samples_QC_summary.csv - Comprehensive QC summary\n"))
cat(paste0("  • ",filt.dir,"/_qc_final_filtered.rds - Post-QC per-sample Seurat objects\n"))
cat(paste0("  • ",filt.dir,"/_filt_cell_metadata.csv - Cell-level metadata for per-sample Seurat objects\n"))
```

### Save out final QC metric objects

Outputs gathered into `qc_final.RData` object

```{r save_qc_final, eval = TRUE}

qc_final.file <- "qc_final.RData"

save(
sample_names,
empty_lower,
empty_niters,
ambient_cutoff,
initial_droplet_count,
initial_gene_count,
validated_barcode_count,
ambient_fraction,
n_doublets,
doublet_rate,
cells_before_doublet_removal,
qc_list,
removal_pct,
threshold_df_final,
gene_qc_list,
after_cell_qc,
final_gene_count,
median_umi,
median_genes,
median_mt_pct,
file = file.path(qc.dir,"qc_metrics", qc_final.file)
)
```

## R sessionInfo

```{r export_sessionInfo}
session.dir <- file.path(base.dir,"sessionInfo")
if (!dir.exists(session.dir)){
  dir.create(session.dir, recursive = T)
}

write.csv(unlist(sessionInfo()), file=file.path(session.dir,"QC_part3_sessionInfo.csv"))
sessionInfo() |>
  capture.output(file=file.path(session.dir,"QC_part3_sessionInfo_summary.txt"))

sessionInfo()
```

## Best Practices: Essential QC Principles

1. Visual Inspection is Key

- Plots are your primary QC tool
- Don’t just print metrics – visualize them
- One good plot > multiple summary statistics

2. Streamlined Reporting

- Brief console feedback during analysis
- Visual QC checks via plots
- ONE comprehensive summary at the end

3. Data-Driven Thresholds with Sanity Checks

- Use percentile-based thresholds
- Apply biological sanity checks – prevent unrealistic thresholds
- PBMCs: ≥500 genes, ≥800 UMI, <10% MT
- Warn if removal rate is >30% (too strict) or <5% (too lenient)
- Document all decisions and adjustments in your summary

4. Iterative QC

- QC is not one-and-done
- Return to QC plots after clustering
- Re-evaluate if results seem off

5. Biological Context

- Know your tissue’s expected metrics
- Different cell types have different profiles
- Stressed cells aren’t always technical artifacts

## Further Reading

### Official Documentation

- Seurat 5 vignettes: https://satijalab.org/seurat/
- Seurat GitHub: https://github.com/satijalab/seurat
- DropletUtils GitHub: https://github.com/MarioniLab/DropletUtils?tab=readme-ov-file
- SoupX GitHub: https://github.com/constantAmateur/SoupX
- scDblFinder GitHub: https://github.com/plger/scDblFinder

### References

- Hao Y, Stuart T, Kowalski MH, et al. Dictionary learning for integrative, multimodal and scalable single-cell analysis. Nat Biotechnol. 2024;42(2):293-304. [Seurat 5 paper]
- Luecken MD, Theis FJ. Current best practices in single-cell RNA-seq analysis: a tutorial. Mol Syst Biol. 2019;15(6):e8746.
- Lun ATL, Riesenfeld S, Andrews T, et al. EmptyDrops: distinguishing cells from empty droplets in droplet-based single-cell RNA sequencing data. Genome Biol. 2019;20(1):63.
- Young MD, Behjati S. SoupX removes ambient RNA contamination from droplet-based single-cell RNA sequencing data. Gigascience. 2020;9(12):giaa151.
- Germain PL, Lun A, Garcia Meixide C, Macnair W, Robinson MD. Doublet identification in single-cell sequencing data using scDblFinder. F1000Res. 2021;10:979.
- Osorio D, Cai JJ. Systematic determination of the mitochondrial proportion in human and mice tissues for single-cell RNA-sequencing data quality control. Bioinformatics. 2021;37(7):963-967.
- McCarthy DJ, Campbell KR, Lun ATL, Wills QF. Scater: pre-processing, quality control, normalization and visualization of single-cell RNA-seq data in R. Bioinformatics. 2017;33(8):1179-1186.
- Amezquita RA, Lun ATL, Becht E, et al. Orchestrating single-cell analysis with Bioconductor. Nat Methods. 2020;17(2):137-145.
- Zheng GXY, Terry JM, Belgrader P, et al. Massively parallel digital transcriptional profiling of single cells. Nat Commun. 2017;8:14049.
- Lee H, Joo JY, Sohn DH, et al. Single-cell RNA sequencing reveals rebalancing of immunological response in patients with periodontitis after non-surgical periodontal therapy. J Transl Med. 2022;20(1):504. [Dataset source]